### 2022-10-01 main.tex
Initial Overleaf Import

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[letterpaper]{geometry}

\title{\bfseries%
Don't interpolate your data!}
\author{Hogg, Casey, Daunt, others?}
\date{October 2022}

\sloppy\sloppypar\raggedbottom\frenchspacing
\begin{document}

\maketitle

\begin{abstract}\noindent
When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
Bound-saturating measurements are made not by manipulating data, but instead by optimizing (or otherwise using) likelihood functions, where the data are treated as fixed, and the model is modified to fit the data.
The shifting and stacking of data can be simply converted into a model-fitting procedure, at small additional computational cost.
The key component of this conversion is a spectral model that is a continuous function of wavelength (or position in the case of imaging) that can represent the psf- and pixel-convolved signal being measured by the device after any reasonable translation.
The advantages of a modeling approach are myriad:
The sacred and expensive data never have to be interpolated or otherwise modified.
Noise maps, data gaps, and bad-data masks, none of which can easily be interpolated, don't have to be manipulated either.
The simplest model-fitting procedure, like the original shift-and-add procedure, is linear in the data, so noise propagation is straightforward.
We demonstrate all these things with toy data and we provide open-source sample code for re-use.
\end{abstract}

\section{Introduction}

Hello World!

\section{Concepts and assumptions}

\paragraph{Multi-epoch spectra with shifts:}

\paragraph{Noise model:}

\paragraph{Bad-pixel mosk:}

\paragraph{Data gaps:}

\paragraph{Pixel-convolved point-spread function:}

\paragraph{Continuous spectral model:}

\paragraph{Linear basis or mixture models:}

\paragraph{Non-uniform fast Fourier transform:}

\paragraph{Band limit:}

\paragraph{Known or fitted image offsets:}

\paragraph{Time variability:}

\paragraph{Sky, tellurics, flat-field:}

\end{document}

### 2022-10-01 main.tex
Updates from Overleaf

@@
 \usepackage[utf8]{inputenc}
+
+% page setup
 \usepackage[letterpaper]{geometry}
+\addtolength{\topmargin}{-0.7in}
+\addtolength{\textheight}{1.8in}
 
@@
 When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
-Bound-saturating measurements are made not by manipulating data, but instead by optimizing (or otherwise using) likelihood functions, where the data are treated as fixed, and the model is modified to fit the data.
-The shifting and stacking of data can be simply converted into a model-fitting procedure, at small additional computational cost.
+Bound-saturating measurements are made not by manipulating data, but instead by optimizing (or otherwise using) 
+a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
+Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed.
 The key component of this conversion is a spectral model that is a continuous function of wavelength (or position in the case of imaging) that can represent the psf- and pixel-convolved signal being measured by the device after any reasonable translation.
-The advantages of a modeling approach are myriad:
+The benefits of a modeling approach are myriad:
 The sacred and expensive data never have to be interpolated or otherwise modified.
@@
 The simplest model-fitting procedure, like the original shift-and-add procedure, is linear in the data, so noise propagation is straightforward.
-We demonstrate all these things with toy data and we provide open-source sample code for re-use.
+The only cost is a small increase in computational complexity.
+We demonstrate all these things with toy data; we provide open-source sample code for re-use.
 \end{abstract}
@@
 
-\paragraph{Bad-pixel mosk:}
+\paragraph{Bad-pixel mask:}
 
@@
 
-\paragraph{Sky, tellurics, flat-field:}
+\paragraph{Sky, tellurics, flat-field, calibration:}
+
+\section{Method}
+
+\section{Toy experiments and results}
+
+\section{Discussion}
 

### 2022-10-07 main.tex
notes to self
section 2 done

@@
 
-% page setup
+% page and document setup
 \usepackage[letterpaper]{geometry}
@@
 \addtolength{\textheight}{1.8in}
+\renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
+
+% text macros
+\newcommand{\sectionname}{Section}
 
@@
 
-\begin{abstract}\noindent
+\paragraph{Abstract}
 When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
@@
 We demonstrate all these things with toy data; we provide open-source sample code for re-use.
-\end{abstract}
 
@@
 
-\paragraph{Multi-epoch spectra with shifts:}
+What we do here can apply to almost any astronomical signals, but we are going to specialize to astronomical spectra for specificity.
+We are going to make many other assumptions too, all of which are discussed in this \sectionname.
 
-\paragraph{Noise model:}
+\paragraph{Multi-epoch spectra with shifts}
+We will assume that there are $n$ observations (epochs) of the same star (or maybe of different stars or quasars, say, that are assumed to be more-or-less identical, intrinsically, but let's think of a single star for now).
+Each of these observations $i$ (with $1\leq i\leq n$) produces a one-dimensional observed (noisy) image $y_i$.
+The $y_i$ can be thought of as $m_i$-vectors in what follows, where $m_i$ is the number of pixels in the spectrum.
+The different observations are made at different relative velocities (star minus spectrograph) such that the different images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
 
-\paragraph{Bad-pixel mask:}
+\paragraph{Shift operators}
+Because the shifts might be non-linear, the shifts $\Delta x_i$ won't be pure numbers but something more like shift operators.
+That is, if there were a true spectrum $\tilde{y}$, each individual noisy image $y_i$ would be related to that true spectrum by a shift operator acting on that true spectrum, plus noise.
+We will begin by assuming that (from some external information) we know these shift operators very accurately.
+At the end we may (do we?) return to the question of how those shift operators are determined.
 
-\paragraph{Data gaps:}
+\paragraph{Wavelength calibration}
+We will assume that, for every $m_i$-pixel image $y_i$ we also have a $m_i$-vector (or list) $x_i$ of pixel positions, such that each element of $x_i$ gives an accurate and precise value for the wavelength (or position in the spectrograph, in some settings) for each pixel of the image $y_i$.
+That is, the device is well calibrated, and we know all the housekeeping data we need for image $y_i$.
 
-\paragraph{Pixel-convolved point-spread function:}
+\paragraph{Noise model}
+Each observation is noisy.
+We will assume that the noise has known properties, and especially that it is additive, (nearly) zero-mean and (nearly) Gaussian in form, with known variance.
+Under these assumptions, each spectrum $y_i$ has an associated variance tensor $C_i$.
+If $y_i$ is an $m_i$-vector, then $C_i$ is a nonnegative-definite $m_i\times m_i$ 2-tensor.
+We don't need to assume that the different pixels of the spectrum received independent noise, but we do assume that each epoch $i$ has an independent noise draw.
 
-\paragraph{Continuous spectral model:}
+\paragraph{Bad-pixel masks}
+Occasionally (or frequently) some of the pixels of a spectrum $y_i$ might be bad---affected by cosmic rays or electronics issues---such that there is a bad-pixel mask $b_i$ associated with each spectrum $y_i$.
+This mask has value 1 in the locations in all good pixels, and 0 in the locations of all bad pixels.
 
-\paragraph{Linear basis or mixture models:}
+\paragraph{Data gaps}
+There is no assumption here that all the $y_i$ have the same number of pixels $m_i$, nor that the pixel grid is in any sense uniform or complete.
+That is, there can be gaps and spaces in the wavelength coverage of the spectra, as there are at chip gaps and where the detector has bad columns.
 
-\paragraph{Non-uniform fast Fourier transform:}
+\paragraph{Pixel-convolved line-spread function}
+The spectrograph has something like a point-spread function or line-spread function, which, in the case of spectrosccopy, sets the shape of an unresolved spectral line.
+In general it depends on the slit or fiber width, the spectrograph resolution and optics, instrument focus, and the properties of the detector.
+In what follows we will not build any model of any of this.
+Instead we will assume that there is a relatively constant pixel-convolved line-spread function (PCLSF), constant both in time, and constant with respect to the shifts $\Delta x_i$ we see in our data set.
+We consider only the PCLSF, because then the sampling by the detector pixels does not require any additional convolution.
+That is, when the PCLSF is used to model a spectrum, the projection onto the pixels is just a sampling of the function at the pixel centers, and does not involve any convolution over the face of the pixel.
+All that said, we won't explicitly model or use the PCLSF in what follows; we will just assume that our continuous spectral model is a model for the pixel-convolved, finite-resolution spectrum; we won't be doing any convolutions in comparing models and data.
 
-\paragraph{Band limit:}
+\paragraph{Continuous spectral model}
+In what follows we are going to avoid interpolating the \emph{data} by having a spectral \emph{model} that can be arbitrarily and losslessly interpolated.
+That means that the spectral model must be a representation of a continuous one-dimensional function.
+If we were working on two-dimensional images, the model would have to be a representation of a continous two-dimensional function.
 
-\paragraph{Known or fitted image offsets:}
+\paragraph{Linear basis or mixture models}
+The simplest kind of model for a continuous function is a sum or mixture or linear combination of continuous basis functions.
+These could be sines and cosines for a Fourier series model.
+They could be spline (or sinc or Lanczos) cardinal basis functions.
+They could be wavelets.
+The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination.
+Thus we will use a linear basis.
 
-\paragraph{Time variability:}
+\paragraph{Band limit}
+We say that a signal is \emph{band-limited} if it has no frequency content above some highest possible cutoff frequency.
+There is a weird sense in which data from a spectrograph both is, and is not, band limited.
+It is band limited in that the spectrograph is finite in resolution, such that no real spectroscopic signal can show features narrower than the PCLSF.
+It is not band-limited in that the photon noise in the device is independent (or close to independent) from pixel to pixel, such that the noise contribution to each observation $y_i$ has support at all spatial frequencies.
 
-\paragraph{Sky, tellurics, flat-field, calibration:}
+One interesting question in what follows is whether the forward model we build for the data should itself be restricted to the band limit defined by the spectrograph resolution, or whether it should permit higher frequencies.
+It might surprise the reader to learn that we prefer to let the model capture higher frequencies, even when we think they shouldn't be there for any reasons other than noise, given the hardware.
+More on this below.
+
+\paragraph{Critical sampling}
+Because of considerations related to the Nyquist frequency on a uniform grid, it is valuable for a spectrograph to have a pixel spacing on the device that is smaller than the width of the PCLSF.
+Specifically, it is valuable for there to be two or three pixels across the full-width at half-maximum (FWHM) of the PCLSF.
+This keeps the band-limited part of the spectroscopic signal well-sampled in the data.
+Spectrographs that don't have pixel scales this fine sometimes do clever things to dither \cite{apogeehardware}.
+
+\paragraph{Non-uniform fast Fourier transform}
+In the special case that the spectral model is a Fourier series, there are very useful and fast tools for interpolation, including especially the non-uniform fast Fourier transform (for example, \cite{finufft}).
+
+\paragraph{Time variability}
+When we average data (or build a static model of multi-epoch data), we are implicitly assuming that there is no intrinsic variability to the source being observed.
+That is, we are assuming that the changes from observation to observation are solely from either the shifts $\Delta x_i$ or from the noise.
+We interpolate the model to account for the shifts, and we combine data to beat down the noise.
+There are situations in which there is substantial time variability and these approaches are still valid, and there are modified approaches.
+We will discuss these changes and generalizations at the end.
+
+\paragraph{Bias, sky, tellurics, flat-field}
+In what follows, we will assume that the input data are well calibrated in various ways.
+In particular, we will assume that the images or spectra have been bias-corrected, sky-subtracted, flat-fielded, and tellurics-corrected sufficiently well that the different observations $y_i$ can be seen as observations of the same underlying or intrinsic signal.
 
@@
 
+Give some summary of the above.
+
+What happens if you don't know your shifts?
+
+What happens if you have multiple shifting signals?
+\texttt{wobble} \cite{wobble} is a solution to one version of that problem.
+
+Discuss time-variable inputs, and how you would modify to deal with that.
+One way involves modeling the time variability with some process.
+Another way involves ignoring it and just getting the average anyway!
+
+What about non-Gaussian noise?
+
+What about data from different instruments with different LSFs / PSFs? What about data with different pixel scales?
+
 \end{document}

### 2022-10-08 main.tex
notes to self; start the model specification

@@
 \usepackage[utf8]{inputenc}
+\usepackage{amsmath}
 
@@
 When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
-Bound-saturating measurements are made not by manipulating data, but instead by optimizing (or otherwise using) 
-a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
-Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed.
-The key component of this conversion is a spectral model that is a continuous function of wavelength (or position in the case of imaging) that can represent the psf- and pixel-convolved signal being measured by the device after any reasonable translation.
+Bound-saturating measurements are made not by manipulating data, but instead by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
+Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean spectrum.
+The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation.
 The benefits of a modeling approach are myriad:
-The sacred and expensive data never have to be interpolated or otherwise modified.
-Noise maps, data gaps, and bad-data masks, none of which can easily be interpolated, don't have to be manipulated either.
-The simplest model-fitting procedure, like the original shift-and-add procedure, is linear in the data, so noise propagation is straightforward.
+The sacred data never have to be interpolated or otherwise modified.
+Noise maps, data gaps, and bad-data masks, none of which can easily be interpolated, don't have to be manipulated.
+The procedure is linear in the data, so noise propagation is straightforward.
+The output can take the form of a spectrum evaluated on a pixel grid, as is expected by the spectroscopic community.
 The only cost is a small increase in computational complexity.
-We demonstrate all these things with toy data; we provide open-source sample code for re-use.
+We demonstrate all these things with both toy and real data; we provide open-source sample code for re-use.
 
@@
 
-Hello World!
+Things to emphasize here and below:
+\begin{itemize}
+\item The input--output relationship of this modeling approach is identical to the input--output relationship of the shift-and-coadd approach. It is just better.
+\item People have a lot of hangups about the word ``model''. We have to use another word, or be very very careful how we use this word.
+\item We have to explain why and how we are better than other approaches in terms that people care about (like radial-velocity precision and abundances etc).
+\item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
+\end{itemize}
 
@@
 
+The problem set-up is that there are $N$ spectral measurements $y_i$ (with $1\leq i\leq N$), each of which is a $M_i$-pixel list of intensities (fluxes maybe).
+Different observations $i$ might have different numbers $M_i$ of pixels.
+The $M_i$ pixels are associated with an $M_i$-length list of positions (wavelengths or log-wavelengths) $x_i$.
+Each measurement or observation $i$ is shifted by a shift operator $\Delta x_i$, which can be thought of as a length-$M_i$ list of pixel offsets or shifts, such that the ``rest frame'' or ``unshifted'' positions corresponding to the length-$M_i$ list of data $y_i$ is the list of differences $x_i - \Delta x_i$.
+The goal is to get a combined spectrum that represents the mean of these spectra, in the common rest frame.
+
+The spectral expectation model $f(x;\theta)$, which will be a representation for our combined spectrum, is a function of $x$ controlled by a list $\theta$ of $P$ linear parameters $a_j$ (with $1\leq j\leq P$).
+This function will be very flexible; possibly even an interpolation of control points or a Fourier series; the number $P$ will be roughly the number of pixels we want in our output combined spectrum.
+The statistical model underlying can be expressed qualitatively as
+\begin{align}
+    y_i &= f(x_i + \Delta x_i;\theta) + \mbox{noise} \\
+    f(x;\theta) &= \sum_{j=1}^P a_j\,g_j(x) ~,
+\end{align}
+where we are being a bit loose with terminology (since $x_i$ is a set of positions, not just one position), the noise is additive (but not yet specified), and each of the $P$ functions $g_j(x)$ is a basis function of the representation.
+
+HOGG: NOISE MODEL
+
+HOGG: BASIS FUNCTIONS.
+
 \section{Toy experiments and results}
@@
 
+\paragraph{Acknowledgements}
+It is a pleasure to thank
+Adrian Price-Whelan (Flatiron),
+and the Astronomical Data Group at the Flatiron Institute
+for valuable discussions.
+
 \end{document}

### 2022-10-08 main.tex
Noise model

@@
 
-HOGG: NOISE MODEL
+The noise model is additive (as noted above) and we will assume that the noise contributing to observation $y_i$ is zero-mean and Gaussian, with a known $M_i\times M_i$ variance tensor $C_i$ (or its inverse $C_i^{-1}$).
+If the raw pixel measurements are independent, which they often are (and are often assumed to be), then the $C_i$ (or $C_i^{-1}$) matrices will be diagonal matrices with the individual pixel variances (or inverse variances) down the diagonals.
+In many cases you don't know (or aren't provided with) the individual pixel variances, and you are just co-adding data with uniform weights.
+That choice, for our purposes, is equivalent to setting every inverse variance matrix $C_i^{-1}$ to the $M_i\times M_i$ identity matrix.
+That's permitted!
+It is no worse a choice than it was (in the past) when you averaged your input data without thinking about noise variances or weights.
+Another sensible default choice would be to set each inverse matrix $C_i^{-1}$ to the identity times the exposure time, such that the eventual math we do will weight the data by the exposure times.
+That is, you do not need to know the noise variances accurately (or even at all) to execute the forthcoming procedure; but if you do know them, it is good to use them.
+
+If you additionally have a bad-pixel mask $b_i$ associated with each observation $y_i$, the bad-pixel mask can be used simply to zero out the corresponding bad rows and columns of the inverse covariance matrix $C_i^{-1}$.
+Importantly, the bad data get zeros in the \emph{inverse} covariance matrix, not the covariance matrix!
 
@@
 
-\section{Toy experiments and results}
+HOGG: LINEAR ALGEBRA.
+
+HOGG: WHAT TO REPORT BACK TO THE USER?
+
+\paragraph{Implementation notes}
+Don't actually construct the very sparse $C_i$ matrices.
+Don't use \texttt{inv()}, use \texttt{solve()}.
+Check your condition numbers.
+
+\section{Experiments and results}
 
@@
 Adrian Price-Whelan (Flatiron),
-and the Astronomical Data Group at the Flatiron Institute
+and the Astronomical Data Group at the Flatiron Institute,
 for valuable discussions.

### 2022-10-08 main.tex
smol things

@@
 Don't interpolate your data!}
-\author{Hogg, Casey, Daunt, others?}
+\author{Hogg, Casey, others?}
 \date{October 2022}
@@
 \item We have to explain why and how we are better than other approaches in terms that people care about (like radial-velocity precision and abundances etc).
+\item All this is even more important in ill-sampled contexts than well-sampled contexts. See, for example, \textsl{APOGEE} raw data!
 \item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
@@
 This keeps the band-limited part of the spectroscopic signal well-sampled in the data.
-Spectrographs that don't have pixel scales this fine sometimes do clever things to dither \cite{apogeehardware}.
+There are spectrograph that don't have pixel scales this fine \cite{apogeehardware}; in these cases what we are proposing here is even more important than it is in the well-sampled case.
 
@@
 
+\paragraph{Software}
+\texttt{numpy} \cite{numpy} ---
+\texttt{finufft} \cite{finufft} ---
+\texttt{matplotlib} \cite{matplotlib}.
+
+\paragraph{Facilities}
+
 \paragraph{Acknowledgements}
 It is a pleasure to thank
+Matt Daunt (NYU),
 Adrian Price-Whelan (Flatiron),
@@
 
+HOGG: SDSS ACK?
+
 \end{document}

### 2022-10-09 main.tex

@@
 \item The input--output relationship of this modeling approach is identical to the input--output relationship of the shift-and-coadd approach. It is just better.
-\item People have a lot of hangups about the word ``model''. We have to use another word, or be very very careful how we use this word.
+\item People have a lot of hangups about the word ``model''.
+We have to use another word, or be very very careful how we use this word.
+For instance, some people I talked to about this thought I meant a physical model of the spectrum starting with atomic physics!
 \item We have to explain why and how we are better than other approaches in terms that people care about (like radial-velocity precision and abundances etc).
@@
 
-HOGG: BASIS FUNCTIONS.
+HOGG: CHOICE OF BASIS FUNCTION. BAND LIMIT. ETC.
 
-HOGG: LINEAR ALGEBRA.
+Now for the method:
+We stack all of the observations $y_i$ into one huge $M$-pixel list $Y$, where $M=\sum_{i=1}^N M_i$.
+We make a $M\times M$ block-diagonal matrix $C^{-1}$ from all the individual inverse variance matrices $C_i^{-1}$.
+We make a $M\times P$ design matrix $X$ which is the evaluation of all the $P$ basis functions at all the pixel locations in all of the $x_i$.
+Once we have these things, the best-fit parameters $\hat\theta$ are
+\begin{align}
+    \hat\theta &= (X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1}\,Y ~,
+\end{align}
+which is the maximum-likelihood or minimum-chi-squared solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \cite{hoggfitting}).
+And now if we choose a final (uniform, say) output pixel grid $x^\star$ in the rest-frame (unshifted) wavelength space, the combined (output, maybe?) spectrum is just
+\begin{align}
+    y^\star &= X^\star\,\hat\theta ~,
+\end{align}
+where $X^\star$ is the evaluation of all of the $P$ basis functions at the locations of the pixels in the output-spectrum pixel grid $x^\star$.
+Putting these things together, the output spectrum $y^\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$ according to
+\begin{align}
+    y^\star &= A\,Y \\
+    A &\equiv X^\star\,(X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1} ~.
+\end{align}
+That is, the method we are delivering is a linear combination of the input data, just like traditional interpolate-and-average.
+But our method never involves, not even implicitly, shifting or interpolating the input data.
 
@@
 Check your condition numbers.
+If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra.
 

### 2022-10-11 main.tex
basis functions, uncertainty estimates

@@
 The procedure is linear in the data, so noise propagation is straightforward.
-The output can take the form of a spectrum evaluated on a pixel grid, as is expected by the spectroscopic community.
+The output can take the form of a spectrum evaluated on a pixel grid, as is often expected by the user.
 The only cost is a small increase in computational complexity.
@@
 
-HOGG: CHOICE OF BASIS FUNCTION. BAND LIMIT. ETC.
+Because we are often dealing with a spectrograph with a (fairly) well defined spectral resolution ($\delta\lambda/\lambda$ roughly constant), and because we care a lot about Doppler shifts (which are uniform in log wavelength), it makes sense to think of the pixel positions $x_i$ and shifts $\Delta x_i$ as being in log-wavelength (or ln-wavelength) units.
+Indeed, the \textsl{Sloan Digital Sky Survey} family of spectrographs extract one-dimensional spectra on a uniform-in-log-wavelength basis.
+Once that choice is made, there are still many choices for the linear basis,
+which include canonical functions from interpolation bases, wavelets, and Fourier series.
+In what follows, somewhat arbitrarily, we will use a Fourier basis, which can be expressed as follows:
+\begin{align}
+    g_j(x) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,x\right) & \mbox{for $j$ odd} \\[3ex]
+                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,x\right) & \mbox{for $j$ even}\end{array}\right. ~,
+\end{align}
+where $L$ is a (long) length-scale in the $x$ space (the log-wavelength space).
 
@@
 \begin{align}
-    y^\star &= X^\star\,\hat\theta ~,
+    y^\star &= X_\star\,\hat\theta ~,
 \end{align}
-where $X^\star$ is the evaluation of all of the $P$ basis functions at the locations of the pixels in the output-spectrum pixel grid $x^\star$.
-Putting these things together, the output spectrum $y^\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$ according to
+where $X_\star$ is the evaluation of all of the $P$ basis functions at the locations of the pixels in the output-spectrum pixel grid $x^\star$.
+Putting these things together, the output spectrum $y^\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
 \begin{align}
     y^\star &= A\,Y \\
-    A &\equiv X^\star\,(X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1} ~.
+    A &\equiv X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1} ~.
 \end{align}
@@
 But our method never involves, not even implicitly, shifting or interpolating the input data.
+
+The uncertainty on the combined spectrum $y^\star$ can be estimated in a few ways.
+The information-theoretic answer is that the covariance matrix $C_\star$ delivering the uncertainties on $y^\star$ is
+\begin{align}
+    C_\star &= [X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top]^{-1} ~.
+\end{align}
+The naive uncertainty variances are the diagonals of this matrix (and not the inverses of the diagonals of its inverse).
+But, \emph{very importantly}, this is only a correct usage when the input $C_i$ matrices really represented the true uncertainties going in.
+If you have any concerns about that, then it is way safer to get your uncertainty estimates from bootstrap or jackknife \cite{hoggflexible}.
 
@@
 If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra.
+If you use the nufft, order your frequencies in the way the software wants, and deal with the complex numbers.
+Specific choice of $L$ and $P$ in the Fourier case.
 

### 2022-10-11 main.tex
deliver to user...

@@
 which is the maximum-likelihood or minimum-chi-squared solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \cite{hoggfitting}).
-And now if we choose a final (uniform, say) output pixel grid $x^\star$ in the rest-frame (unshifted) wavelength space, the combined (output, maybe?) spectrum is just
+And now if we choose a final (uniform, say) output pixel grid $x_\star$ in the rest-frame (unshifted) wavelength space, the combined (output, maybe?) spectrum is just
 \begin{align}
-    y^\star &= X_\star\,\hat\theta ~,
+    y_\star &= X_\star\,\hat\theta ~,
 \end{align}
-where $X_\star$ is the evaluation of all of the $P$ basis functions at the locations of the pixels in the output-spectrum pixel grid $x^\star$.
-Putting these things together, the output spectrum $y^\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
+where $X_\star$ is the $M_\star\times P$ evaluation of all of the $P$ basis functions at the locations of the $M_\star$ pixels in the output-spectrum pixel grid $x_\star$.
+Putting these things together, the output spectrum $y_\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
 \begin{align}
-    y^\star &= A\,Y \\
-    A &\equiv X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1} ~.
+    y_\star &= A_\star\,Y \\
+    A_\star &\equiv X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1} ~.
 \end{align}
@@
 
-The uncertainty on the combined spectrum $y^\star$ can be estimated in a few ways.
-The information-theoretic answer is that the covariance matrix $C_\star$ delivering the uncertainties on $y^\star$ is
+The uncertainty on the combined spectrum $y_\star$ can be estimated in a few ways.
+The information-theoretic answer is that the covariance matrix $C_\star$ delivering the uncertainties on $y_\star$ is
 \begin{align}
@@
 
-HOGG: WHAT TO REPORT BACK TO THE USER?
+In the end, most users want the combined spectrum $y_\star$ (plus the pixel positions $x_\star$ and uncertainty informaion).
+Some users will additionally want (or want to reconstruct) the parameters $\hat\theta$ and the basis functions $g_j(x)$.
+For this reasons it often makes sense to set $M_\star\geq P$ such that the delivered $y_\star$ fully determines the parameters.
 
@@
 If you use the nufft, order your frequencies in the way the software wants, and deal with the complex numbers.
-Specific choice of $L$ and $P$ in the Fourier case.
+Specific choice of $L$ and $P$ and $M_\star$ in the Fourier case.
 

### 2022-10-14 main.tex
start to write the experiments section.

@@
 
-In the end, most users want the combined spectrum $y_\star$ (plus the pixel positions $x_\star$ and uncertainty informaion).
+In the end, most users want the combined spectrum $y_\star$ (plus the pixel positions $x_\star$ and uncertainty information).
 Some users will additionally want (or want to reconstruct) the parameters $\hat\theta$ and the basis functions $g_j(x)$.
-For this reasons it often makes sense to set $M_\star\geq P$ such that the delivered $y_\star$ fully determines the parameters.
+For these reasons it often makes sense to set $M_\star\geq P$ such that the delivered $y_\star$ fully determines the parameters.
+But that's a detailed question about your users and your goals.
 
 \paragraph{Implementation notes}
-Don't actually construct the very sparse $C_i$ matrices.
-Don't use \texttt{inv()}, use \texttt{solve()}.
-Check your condition numbers.
-If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra.
+Don't actually construct the very sparse $C_i$ matrices, just do pixel-wise multiplies.
+Don't use \texttt{inv()}, use \texttt{solve()} or \texttt{lstsq()}.
+Check your condition numbers before you go.
+If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra; research those.
 If you use the nufft, order your frequencies in the way the software wants, and deal with the complex numbers.
@@
 \section{Experiments and results}
+
+\paragraph{Fake data}
+We demonstrate the effectiveness of the spectral combination method on artificial data.
+The spectral data were generated from a mean spectral expectation that is a continuum plus a randomly generated (not at all physical) line list, where each line equivalent width was set by a random draw from a power law.
+Each spectral epoch was given a velocity (Doppler shift) by putting the epochs equally spaced on a sinusoidal variation with a semi-amplitude of $30\,\kmps$ (to emulate the motion of the spectrograph with respect to the Solar System barycenter).
+The data were generated with an assumption of unresolved spectral lines and a Gaussian line-spread function (LSF) by simply evaluating the spectral expectation model at pixel centers, so the LSF is effectively the pixel-convolved LSF.
+The pixel grids were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
+After sampling the mean spectral expectation model at the pixels, we add Gaussian noise with a uniform variance such that the signal-to-noise in the continuum is XXXX per pixel in each epoch spectrum.
+At a rate of XXX pixels were randomly and independently marked as ``bad'' and offset by a large offset.
+At each epoch $i$, there is a spectrum $y_i$ with $M_i=XXX$ pixels, a spectrograph rest-frame pixel grid $x_i$ (in log wavelength $x$), and a $M_i$-pixel binary bad-pixel mask $b_i$.
+The full data set is shown in Figure~\label{fig:data}.
+
+For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by YYYY, which (just barely) samples the pixel-convolved LSF well.
+We treat every pixel in the data as having the same noise variance $\sigma_y$ (which is indeed how they were generated).
+
+\paragraph{Standard practice}
+HOGG: Choose an interpolator.
+
+HOGG: Interpolate to the $x_\star$ grid and average.
+
+HOGG: Deal with the bad pixels how?
+
+HOGG: Standard practice is way sucky. It is shown in Figure~XX.
+
+\paragraph{Our method HOGG THIS NEEDS A NAME}
+
+We do what we say and show it in Figure~XXX.
 
@@
 
+What happens if your data aren't continuum normalized properly?
+
 Discuss time-variable inputs, and how you would modify to deal with that.
@@
 What about data from different instruments with different LSFs / PSFs? What about data with different pixel scales?
+
+Code is available at HOGG THIS URL.
 

### 2022-10-15 main.tex
tiny

@@
 \newcommand{\sectionname}{Section}
+
+% math macros
+\newcommand{\unit}[1]{\mathrm{#1}}
+\newcommand{\mps}{\unit{m\,s^{-1}}}
+\newcommand{\kmps}{\unit{km\,s^{-1}}}
 
@@
 
-\paragraph{Bias, sky, tellurics, flat-field}
+\paragraph{Bias, sky, tellurics, flat-field, normalization}
 In what follows, we will assume that the input data are well calibrated in various ways.
 In particular, we will assume that the images or spectra have been bias-corrected, sky-subtracted, flat-fielded, and tellurics-corrected sufficiently well that the different observations $y_i$ can be seen as observations of the same underlying or intrinsic signal.
+Alternatively, the input data might be continuum-normalized or pseudo-continuum-normalized.
+Again, the assumption is that the different calibrated or processed observations are of the same underlying signal.
 
@@
 The data were generated with an assumption of unresolved spectral lines and a Gaussian line-spread function (LSF) by simply evaluating the spectral expectation model at pixel centers, so the LSF is effectively the pixel-convolved LSF.
-The pixel grids were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
+The individual pixel grids $x_i$ were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
 After sampling the mean spectral expectation model at the pixels, we add Gaussian noise with a uniform variance such that the signal-to-noise in the continuum is XXXX per pixel in each epoch spectrum.
@@
 At each epoch $i$, there is a spectrum $y_i$ with $M_i=XXX$ pixels, a spectrograph rest-frame pixel grid $x_i$ (in log wavelength $x$), and a $M_i$-pixel binary bad-pixel mask $b_i$.
-The full data set is shown in Figure~\label{fig:data}.
+The full data set is shown in Figure~\label{fig:data}, and a zoom-in on one epoch is shown in \figurename~\ref{fig:datazoom}.
 
@@
 
-What about data from different instruments with different LSFs / PSFs? What about data with different pixel scales?
+What about data from different instruments with different LSFs / PSFs? What about data with different pixel scales? What about data with unknown wavelength calibration?
 

### 2022-10-16 main.tex
standard practice

@@
 The only cost is a small increase in computational complexity.
-We demonstrate all these things with both toy and real data; we provide open-source sample code for re-use.
+We demonstrate all of these things with a small data example and we provide open-source sample code for re-use.
 
@@
 \paragraph{Standard practice}
-HOGG: Choose an interpolator.
+In standard practice, the individual spectra $y_i$ are interpolated to rest-frame spectra $y'_i$ using an interpolator.
+In detail, the data are shifted and also resampled onto the output pixel grid $x_\star$.
+The interpolator is a choice; cubic spline, Lanczos, and sinc interpolations are common, since they have good properties with respect to the spectrograph band limit.
+Here we choose a cubic spline for convenience ANDY: WHAT SHOULD WE CHOOSE??.
 
-HOGG: Interpolate to the $x_\star$ grid and average.
+The interpolated spectra $y'_i$ are, by construction, all on the same rest-frame wavelength grid $x_\star$ so they can be averaged, with a straight average, a median, an inverse-variance-weighted mean, or any more sophisticated algorithm.
+Here we choose an inverse-variance-weighted mean, but zeroing out the inverse variance at the locations of the bad pixels.
 
-HOGG: Deal with the bad pixels how?
+This censorship of the bad pixels requires some attention:
+Naive interpolation of an integer (binary actually) bad-pixel mask $b_i$ will not deliver an integer mask $b'_i$.
+The conservative move (which we adopt) is to naively interpolate $b_i$ and then zero out any pixels that are significantly below unity.
+In general this substantially grows the binary mask, and reduces the amount of data available for combination.
 
-HOGG: Standard practice is way sucky. It is shown in Figure~XX.
+We show the results of this combintation, for the fake data, in \figurename~\ref{fig:standard}.
+We also show the comparison with the true spectral expectation model that was employed to make the fake data.
+We find HOGG WHAT??
 
-\paragraph{Our method HOGG THIS NEEDS A NAME}
+\paragraph{Our method: The forward model}
+HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
 
-We do what we say and show it in Figure~XXX.
+The combined spectrum resulting from the method for forward modeling described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:combined}.
+Again, it is compared to the true spectral expectation model.
+We find HOGG WHAT?? HERE SAY ABSOLUTELY WHAT WE FIND.
+
+HOGG: HERE SAY WHAT WE FIND RELATIVELY to the standard practice.
 
@@
 
-\paragraph{Facilities}
-
 \paragraph{Acknowledgements}
@@
 
-HOGG: SDSS ACK?
-
 \end{document}

### 2022-10-16 main.tex
tweaks

@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
@@
 \item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
+\item This model is a sub-model of \textsl{wobble} \cite{wobble}.
 \end{itemize}
 
-\section{Concepts and assumptions}
+\section{Concepts and assumptions}\label{sec:assumptions}
 
@@
 
-\section{Method}
+\section{Method}\label{sec:method}
 
@@
 
-\section{Experiments and results}
+\section{Experiments and results}\label{sec:results}
 
@@
 
-We show the results of this combintation, for the fake data, in \figurename~\ref{fig:standard}.
+We show the results of Standar Practice (tm), for the fake data, in \figurename~\ref{fig:standard}.
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
@@
 
-The combined spectrum resulting from the method for forward modeling described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:combined}.
+The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:combined}.
 Again, it is compared to the true spectral expectation model.
@@
 
-HOGG: HERE SAY WHAT WE FIND RELATIVELY to the standard practice.
+HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
 
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 

### 2022-10-16 main.tex
notes to self

@@
 All that said, we won't explicitly model or use the PCLSF in what follows; we will just assume that our continuous spectral model is a model for the pixel-convolved, finite-resolution spectrum; we won't be doing any convolutions in comparing models and data.
+
+HOGG: Something about the LSF being different in different exposures!
 
@@
 
-\paragraph{Standard practice}
+\paragraph{Standard Practice (tm)}
 In standard practice, the individual spectra $y_i$ are interpolated to rest-frame spectra $y'_i$ using an interpolator.
@@
 
-\paragraph{Our method: The forward model}
+\paragraph{Our method: Forward Model (tm)}
 HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
@@
 What happens if your data aren't continuum normalized properly?
+
+How do you modify to deal with LSF variations (like we have in \textsl{SDSS-V APOGEE}?
 

### 2022-10-17 main.tex
notes to self

@@
 
-We show the results of Standar Practice (tm), for the fake data, in \figurename~\ref{fig:standard}.
+We show the results of Standard Practice (tm), for the fake data, in \figurename~\ref{fig:standard}.
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
@@
 
-\paragraph{Our method: Forward Model (tm)}
+\paragraph{Fiducial case}
 HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
+
+HOGG: Give full list of input parameters for the Fiducial.
 
@@
 HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
+
+\paragraph{Dependence on raw-data pixel scale}
+HOGG show that things with Standard Practice (tm) depend VERY STRONGLY on the input pixel grid.
+
+\paragraph{Noise and noise covariances}
+HOGG do repeated trials to empirically measure the noise covariances.
+
+Plot the empirical covariances as a function of input-data pixel scale, interpolation method, and number of epochs.
 
@@
 How do you modify to deal with LSF variations (like we have in \textsl{SDSS-V APOGEE}?
+What if you don't know the LSF differences?
+Difference between modeling LSF changes and modeling LSFs. The former is good, the latter is bad.
 

### 2022-10-21 main.tex

@@
 The benefits of a modeling approach are myriad:
-The sacred data never have to be interpolated or otherwise modified.
-Noise maps, data gaps, and bad-data masks, none of which can easily be interpolated, don't have to be manipulated.
-The procedure is linear in the data, so noise propagation is straightforward.
-The output can take the form of a spectrum evaluated on a pixel grid, as is often expected by the user.
+The sacred data never have to be modified.
+Noise maps, data gaps, and bad-data masks don't have to be interpolated.
+The output can take the form of a spectrum evaluated on a pixel grid, as is traditional.
+The noise in the output spectrum becomes uncorrelated across neighboring pixels.
 The only cost is a small increase in computational complexity.

### 2022-10-22 main.tex
Figures! They need captions

@@
+% Copyright 2022 David W. Hogg (NYU) and Andy Casey (Monash)
+
+% To-Do items
+% -----------
+% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code?
+% - Write about the two input-data cases.
+% - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
+% - Copy in a figure macro from Hogg & Villar and use it.
+
 \documentclass[11pt]{article}
@@
 \renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
+
+% figure setup
+\usepackage{graphicx}
+\usepackage{xcolor}
+\usepackage[framemethod=tikz]{mdframed}
+\usetikzlibrary{shadows}
+\definecolor{captiongray}{HTML}{555555}
+\mdfsetup{%
+  innertopmargin=2ex,
+  innerbottommargin=1.8ex,
+  linecolor=captiongray,
+  linewidth=0.5pt,
+  roundcorner=1pt,
+  shadow=false,
+}
+\newlength{\figurewidth}
+\setlength{\figurewidth}{0.75\textwidth}
 
@@
 At each epoch $i$, there is a spectrum $y_i$ with $M_i=XXX$ pixels, a spectrograph rest-frame pixel grid $x_i$ (in log wavelength $x$), and a $M_i$-pixel binary bad-pixel mask $b_i$.
-The full data set is shown in Figure~\label{fig:data}, and a zoom-in on one epoch is shown in \figurename~\ref{fig:datazoom}.
+The full data set is shown in Figure~\label{fig:data1}, and a zoom-in on one epoch is shown in \figurename~\ref{fig:data2}.
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=1.3\figurewidth]{notebooks/data1.pdf}\\
+    \includegraphics[width=\figurewidth]{notebooks/datazoom.pdf}
+    \end{center}
+    \caption{HOGG: Something.}
+    \label{fig:data1}
+    \end{mdframed}
+\end{figure}
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=1.3\figurewidth]{notebooks/data2.pdf}
+    \end{center}
+    \caption{HOGG: Something.}
+    \label{fig:data2}
+    \end{mdframed}
+\end{figure}
 
@@
 We find HOGG WHAT??
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/standard1.pdf}\\
+    \includegraphics[width=\figurewidth]{notebooks/standard2.pdf}
+    \end{center}
+    \caption{HOGG: Something.}
+    \label{fig:standard}
+    \end{mdframed}
+\end{figure}
 
@@
 
-The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:combined}.
+The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
 Again, it is compared to the true spectral expectation model.
 We find HOGG WHAT?? HERE SAY ABSOLUTELY WHAT WE FIND.
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/forward1.pdf}\\
+    \includegraphics[width=\figurewidth]{notebooks/forward2.pdf}
+    \end{center}
+    \caption{HOGG: Something.}
+    \label{fig:forward}
+    \end{mdframed}
+\end{figure}
 
 HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
-
-\paragraph{Dependence on raw-data pixel scale}
-HOGG show that things with Standard Practice (tm) depend VERY STRONGLY on the input pixel grid.
 
@@
 Plot the empirical covariances as a function of input-data pixel scale, interpolation method, and number of epochs.
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/noise.pdf}
+    \end{center}
+    \caption{HOGG: Something.}
+    \label{fig:noise}
+    \end{mdframed}
+\end{figure}
 

### 2022-10-23 main.tex

@@
 % - Copy in a figure macro from Hogg & Villar and use it.
+% - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.
 
@@
 At a rate of XXX pixels were randomly and independently marked as ``bad'' and offset by a large offset.
+
+HOGG: MAKE 2 DATA SETS, one poorly sampled, one well sampled.
+
 At each epoch $i$, there is a spectrum $y_i$ with $M_i=XXX$ pixels, a spectrograph rest-frame pixel grid $x_i$ (in log wavelength $x$), and a $M_i$-pixel binary bad-pixel mask $b_i$.
@@
     \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:data1}
+    \caption{The top 8 panels show the $N=8$ epochs of the poorly sampled input data.
+The bad pixels are indicated with grey bars.
+The bottom 2 panels show a zoom in on one of the epochs, along with the true spectrum used to generate the data.
+The true spectrum is shown at zero Doppler shift, while each epoch spectrum is at a finite Doppler shift $\Delta x_i$.\label{fig:data1}}
     \end{mdframed}
@@
     \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:data2}
+    \caption{Similar to the top panels of \figurename~\ref{fig:data1} except showing the well sampled input data.\label{fig:data2}}
     \end{mdframed}
@@
 
-For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by YYYY, which (just barely) samples the pixel-convolved LSF well.
+For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by $1/R$, which (just barely) samples the pixel-convolved LSF well.
 We treat every pixel in the data as having the same noise variance $\sigma_y$ (which is indeed how they were generated).
-
-\paragraph{Standard Practice (tm)}
-In standard practice, the individual spectra $y_i$ are interpolated to rest-frame spectra $y'_i$ using an interpolator.
-In detail, the data are shifted and also resampled onto the output pixel grid $x_\star$.
-The interpolator is a choice; cubic spline, Lanczos, and sinc interpolations are common, since they have good properties with respect to the spectrograph band limit.
-Here we choose a cubic spline for convenience ANDY: WHAT SHOULD WE CHOOSE??.
-
-The interpolated spectra $y'_i$ are, by construction, all on the same rest-frame wavelength grid $x_\star$ so they can be averaged, with a straight average, a median, an inverse-variance-weighted mean, or any more sophisticated algorithm.
-Here we choose an inverse-variance-weighted mean, but zeroing out the inverse variance at the locations of the bad pixels.
-
-This censorship of the bad pixels requires some attention:
-Naive interpolation of an integer (binary actually) bad-pixel mask $b_i$ will not deliver an integer mask $b'_i$.
-The conservative move (which we adopt) is to naively interpolate $b_i$ and then zero out any pixels that are significantly below unity.
-In general this substantially grows the binary mask, and reduces the amount of data available for combination.
-
-We show the results of Standard Practice (tm), for the fake data, in \figurename~\ref{fig:standard}.
-We also show the comparison with the true spectral expectation model that was employed to make the fake data.
-We find HOGG WHAT??
-\begin{figure}[t!]
-    \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/standard1.pdf}\\
-    \includegraphics[width=\figurewidth]{notebooks/standard2.pdf}
-    \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:standard}
-    \end{mdframed}
-\end{figure}
 
@@
     \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:forward}
+    \caption{The top two panels show the result of running the Forward Model (tm) on the poorly sampled input data, plus the residual away from the truth.
+The bottom two panels show the same but for the well sampled input data.
+In each panel, the red line shows the true model used to generate the data.\label{fig:forward}}
     \end{mdframed}
@@
 
+\paragraph{Standard Practice (tm)}
+In standard practice, the individual spectra $y_i$ are interpolated to rest-frame spectra $y'_i$ using an interpolator.
+In detail, the data are shifted and also resampled onto the output pixel grid $x_\star$.
+The interpolator is a choice; cubic spline, Lanczos, and sinc interpolations are common, since they have good properties with respect to the spectrograph band limit.
+Here we choose a cubic spline for convenience ANDY: WHAT SHOULD WE CHOOSE??.
+
+The interpolated spectra $y'_i$ are, by construction, all on the same rest-frame wavelength grid $x_\star$ so they can be averaged, with a straight average, a median, an inverse-variance-weighted mean, or any more sophisticated algorithm.
+Here we choose an inverse-variance-weighted mean, but zeroing out the inverse variance at the locations of the bad pixels.
+
+This censorship of the bad pixels requires some attention:
+Naive interpolation of an integer (binary actually) bad-pixel mask $b_i$ will not deliver an integer mask $b'_i$.
+The conservative move (which we adopt) is to naively interpolate $b_i$ (also by the cubic spline) and then zero out any pixels that are significantly below unity.
+In general this substantially grows the binary mask, and reduces the amount of data available for combination.
+
+An example of a poorly-sampled spectrum interpolated to the rest frame is shown in \figurename~\ref{fig:interpolated}.
+The interpolation introduces substantial ringing artifacts in the spectrum (compare to \figurename~\ref{fig:data1}).
+It also grows the bad-pixel mask around the bad pixel.
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/datazoom_interpolated.pdf}
+    \end{center}
+    \caption{HOGG: WRITE CAPTION.\label{fig:interpolated}}
+    \end{mdframed}
+\end{figure}
+
+We show the final result of Standard Practice (tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
+We also show the comparison with the true spectral expectation model that was employed to make the fake data.
+We find HOGG WHAT??
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/standard1.pdf}\\
+    \includegraphics[width=\figurewidth]{notebooks/standard2.pdf}
+    \end{center}
+    \caption{Same as \figurename~\ref{fig:forward} but for the output of Standard Practice (tm).
+Note that the residuals appear to be spatially correlated; more about this in the text and in \figurename~\ref{fig:noise}.\label{fig:standard}}
+    \end{mdframed}
+\end{figure}
+
 \paragraph{Noise and noise covariances}
-HOGG do repeated trials to empirically measure the noise covariances.
+HOGG do repeated trials to empirically measure the noise covariances. How exactly did we do that?
 
-Plot the empirical covariances as a function of input-data pixel scale, interpolation method, and number of epochs.
+In \figurename~\ref{fig:noise} we show the empirical covariances for the poorly sampled and well sampled data sets.
+In each case, Standard Practice (tm) has a lower noise variance at zero lag, but shows spatially correlated noise over many pixels.
+The Forward Model (tm), in contrast, shows no pixel-to-pixel noise covariances, even when the input data are ill sampled.
 \begin{figure}[t!]

### 2022-10-23 main.tex
add a figure; notes to self

@@
 \end{align}
-where $L$ is a (long) length-scale in the $x$ space (the log-wavelength space).
+where $L$ is a (long) length-scale in the $x$ space (the log-wavelength space), and $1\leq j\leq P$.
 
@@
 The data were generated with an assumption of unresolved spectral lines and a Gaussian line-spread function (LSF) by simply evaluating the spectral expectation model at pixel centers, so the LSF is effectively the pixel-convolved LSF.
-The individual pixel grids $x_i$ were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
-After sampling the mean spectral expectation model at the pixels, we add Gaussian noise with a uniform variance such that the signal-to-noise in the continuum is XXXX per pixel in each epoch spectrum.
+After sampling the mean spectral expectation model onto an observational pixel grid, we add Gaussian noise with a uniform variance (HOGG NO WE DON'T!) such that the signal-to-noise in the continuum is XXXX per pixel in each epoch spectrum.
 At a rate of XXX pixels were randomly and independently marked as ``bad'' and offset by a large offset.
@@
 HOGG: MAKE 2 DATA SETS, one poorly sampled, one well sampled.
+The individual pixel grids $x_i$ were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
 
@@
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
-We find HOGG WHAT??
+We find HOGG WHAT?? CORRELATIONS; WORSE RESIDUALS NEAR SPECTRAL FEATURES (UH OH).
 \begin{figure}[t!]
@@
     \caption{Same as \figurename~\ref{fig:forward} but for the output of Standard Practice (tm).
-Note that the residuals appear to be spatially correlated; more about this in the text and in \figurename~\ref{fig:noise}.\label{fig:standard}}
+Note that in the poorly sampled case, the residuals appear to be spatially correlated, and the residuals appear to be larger near strong spectral features; more about these issues in the text and in \figurename~\ref{fig:noise}.\label{fig:standard}}
     \end{mdframed}
@@
 
+We assumed no sky subtraction necessary. What gives?
+
 How do you modify to deal with LSF variations (like we have in \textsl{SDSS-V APOGEE}?

### 2022-10-23 paper/main.tex
moving furniture

--- 2022-10-22 main.tex
@@
 
-For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by YYYY, which (just barely) samples the pixel-convolved LSF well.
+For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by $1/R$, which (just barely) samples the pixel-convolved LSF well.
 We treat every pixel in the data as having the same noise variance $\sigma_y$ (which is indeed how they were generated).
+
+\paragraph{Fiducial case}
+HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
+
+HOGG: Give full list of input parameters for the Fiducial.
+
+The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
+Again, it is compared to the true spectral expectation model.
+We find HOGG WHAT?? HERE SAY ABSOLUTELY WHAT WE FIND.
+\begin{figure}[t!]
+    \begin{mdframed}\begin{center}
+    \includegraphics[width=\figurewidth]{notebooks/forward1.pdf}\\
+    \includegraphics[width=\figurewidth]{notebooks/forward2.pdf}
+    \end{center}
+    \caption{The top two panels show the result of running the Forward Model (tm) on the poorly sampled input data, plus the residual away from the truth.
+The bottom two panels show the same but for the well sampled input data.
+In each panel, the red line shows the true model used to generate the data.\label{fig:forward}}
+    \end{mdframed}
+\end{figure}
+
+HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
 
@@
     \end{center}
-    \caption{HOGG: Something.}
+    \caption{Same as \figurename~\ref{fig:forward} but for the output of Standard Practice (tm). Note that the residuals appear to be spatially correlated.}
     \label{fig:standard}
@@
 \end{figure}
-
-\paragraph{Fiducial case}
-HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
-
-HOGG: Give full list of input parameters for the Fiducial.
-
-The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
-Again, it is compared to the true spectral expectation model.
-We find HOGG WHAT?? HERE SAY ABSOLUTELY WHAT WE FIND.
-\begin{figure}[t!]
-    \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/forward1.pdf}\\
-    \includegraphics[width=\figurewidth]{notebooks/forward2.pdf}
-    \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:forward}
-    \end{mdframed}
-\end{figure}
-
-HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
 

### 2022-10-23 paper/main.tex

--- 2022-10-23 main.tex
@@
 Each spectral epoch was given a velocity (Doppler shift) by putting the epochs equally spaced on a sinusoidal variation with a semi-amplitude of $30\,\kmps$ (to emulate the motion of the spectrograph with respect to the Solar System barycenter).
-The data were generated with an assumption of unresolved spectral lines and a Gaussian line-spread function (LSF) by simply evaluating the spectral expectation model at pixel centers, so the LSF is effectively the pixel-convolved LSF.
-After sampling the mean spectral expectation model onto an observational pixel grid, we add Gaussian noise with a uniform variance (HOGG NO WE DON'T!) such that the signal-to-noise in the continuum is XXXX per pixel in each epoch spectrum.
-At a rate of XXX pixels were randomly and independently marked as ``bad'' and offset by a large offset.
+The spectral expectation model is built on the assumption of a normalized (unit) continuum and unresolved spectral lines with a Gaussian line-spread function (LSF) with one-sigma width $1/R$, where $R=135\,000$ is the spectrograph resolution).
 
-HOGG: MAKE 2 DATA SETS, one poorly sampled, one well sampled.
-The individual pixel grids $x_i$ were set to slightly under-sample the pixel-convolved LSF, and they were fixed in the rest frame of the spectrograph.
+The sampled data at each epoch were made by simply evaluating the spectral expectation model (shifted by the Doppler shift) at the pixel centers, such that the $R=135\,000$ LSF is effectively the pixel-convolved LSF.
+After sampling the mean spectral expectation model onto an observational pixel grid, Gaussian noise is added with a variance that grows linearly with expected flux, such that the signal-to-noise per pixel in the continuum has a definite known value in each epoch spectrum.
+At a rate of 0.01 (one percent), sets of three adjacent pixels were randomly and independently marked as ``bad'' and offset by a large positive offset.
 
-At each epoch $i$, there is a spectrum $y_i$ with $M_i=XXX$ pixels, a spectrograph rest-frame pixel grid $x_i$ (in log wavelength $x$), and a $M_i$-pixel binary bad-pixel mask $b_i$.
-The full data set is shown in Figure~\label{fig:data1}, and a zoom-in on one epoch is shown in \figurename~\ref{fig:data2}.
+According to these rules we made two input data sets.
+One is poorly sampled, with $M=171$ spectral pixels in the input data separated by $2/R$ in $x$ (natural logarithm of wavelength), and signal-to-noise per pixel of 17 in the continuum.
+The other is well sampled, with $M=340$ spectral pixels in the input data separated by $1/R$ in $x$, and signal-to-noise per pixel of 12 in the continuum.
+In each case there are $N=8$ epochs.
+The two complete data sets are shown in \figurename~\label{fig:data1} and \figurename~\ref{fig:data2}.
 \begin{figure}[t!]
@@
     \end{center}
-    \caption{The top 8 panels show the $N=8$ epochs of the poorly sampled input data.
+    \caption{The top 8 panels show the $N=8$ epochs of the poorly sampled input data, with $N=8$, $M=171$, and signal-to-noise per pixel of 17 in the continuum.
 The bad pixels are indicated with grey bars.
 The bottom 2 panels show a zoom in on one of the epochs, along with the true spectrum used to generate the data.
-The true spectrum is shown at zero Doppler shift, while each epoch spectrum is at a finite Doppler shift $\Delta x_i$.\label{fig:data1}}
+The true spectrum is shown at zero Doppler shift, while each epoch spectrum is at a finite Doppler shift $\Delta x_i$ relative to the spectrograph pixel grid.\label{fig:data1}}
     \end{mdframed}
@@
     \end{center}
-    \caption{Similar to the top panels of \figurename~\ref{fig:data1} except showing the well sampled input data.\label{fig:data2}}
+    \caption{Similar to the top panels of \figurename~\ref{fig:data1} except showing the well sampled input data with $N=8$, $M=340$, and signal-to-noise per pixel of 12 in the continuum.\label{fig:data2}}
     \end{mdframed}
@@
 
-For the combined spectrum we choose a pixel grid $x_\star$ with $M_\star$ pixels evenly spaced in log wavelength $x$, separated by $1/R$, which (just barely) samples the pixel-convolved LSF well.
-We treat every pixel in the data as having the same noise variance $\sigma_y$ (which is indeed how they were generated).
-
-\paragraph{Fiducial case}
-HOGG: OUR METHOD NEEDS A CONSISTENT NAME USED THROUGHOUT. FORWARD MODEL??
-
-HOGG: Give full list of input parameters for the Fiducial.
+\paragraph{Forward Model (tm)}
+HOGG: Give full list of input parameters for the Fiducial. NOTE THAT WE DON'T DO THE NOISE IVAR RIGHT.
 
@@
 
-HOGG: HERE SAY WHAT WE FIND RELATIVELY to Standard Practice (tm).
-
 \paragraph{Standard Practice (tm)}
@@
 The interpolator is a choice; cubic spline, Lanczos, and sinc interpolations are common, since they have good properties with respect to the spectrograph band limit.
-Here we choose a cubic spline for convenience ANDY: WHAT SHOULD WE CHOOSE??.
+Here we choose a cubic spline for convenience, but there are many other choices, including linear, sinc, and Lanczos interpolators, among many others.
 
 The interpolated spectra $y'_i$ are, by construction, all on the same rest-frame wavelength grid $x_\star$ so they can be averaged, with a straight average, a median, an inverse-variance-weighted mean, or any more sophisticated algorithm.
-Here we choose an inverse-variance-weighted mean, but zeroing out the inverse variance at the locations of the bad pixels.
+Here we choose an inverse-variance-weighted mean (HOGG: NO WE DON'T), but zeroing out the inverse variance at the locations of the bad pixels.
 
@@
 
+HOGG: HERE SAY WHAT WE FIND RELATIVELY between the Forward Model (tm) and Standard Practice (tm).
+
 \paragraph{Noise and noise covariances}
@@
 \end{figure}
+
+HOGG: Note that if we choose $P<M$ we start to see noise correlations in the Forward Model (tm) too.
+That's one of a few reasons we like $P=M$.
 

### 2022-10-24 paper/main.tex

@@
 
-\documentclass[11pt]{article}
+\documentclass[modern]{aastex631}
 \usepackage[utf8]{inputenc}
@@
 % page and document setup
-\usepackage[letterpaper]{geometry}
-\addtolength{\topmargin}{-0.7in}
-\addtolength{\textheight}{1.8in}
-\renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
+\addtolength{\topmargin}{-0.35in}
+\addtolength{\textheight}{0.6in}
+% \renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
 
@@
 % text macros
+\shorttitle{don't interpolate your data!}
+\shortauthors{hogg \& casey}
 \newcommand{\sectionname}{Section}
@@
 
-\title{\bfseries%
-Don't interpolate your data!}
-\author{Hogg, Casey, others?}
-\date{October 2022}
-
 \sloppy\sloppypar\raggedbottom\frenchspacing
@@
 
-\maketitle
+\title{\Huge Don't interpolate your data!}
 
-\paragraph{Abstract}
+\author[0000-0003-2866-9403]{David W. Hogg}
+\affiliation{Center for Cosmology and Particle Physics, Department of Physics, New York University}
+\affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
+\affiliation{Flatiron Institute, a division of the Simons Foundation}
+
+\author[0000-0003-0174-0564]{Andrew R. Casey}
+\affiliation{School of Physics \& Astronomy, Monash University}
+\affiliation{Faculty of Information Technology, Monash University}
+
+\begin{abstract}\noindent
 When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
 Bound-saturating measurements are made not by manipulating data, but instead by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
-Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean spectrum.
+Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean.
 The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation.
 The benefits of a modeling approach are myriad:
-The sacred data never have to be modified.
-Noise maps, data gaps, and bad-data masks don't have to be interpolated.
+The sacred data never are modified.
+Noise maps, data gaps, and bad-data masks don't require interpolation.
 The output can take the form of a spectrum evaluated on a pixel grid, as is traditional.
@@
 We demonstrate all of these things with a small data example and we provide open-source sample code for re-use.
+\end{abstract}
 
+\keywords{Foo --- Bar --- HOGG}
+
+\section*{}\clearpage
 \section{Introduction}\label{sec:intro}
@@
 According to these rules we made two input data sets.
-One is poorly sampled, with $M=171$ spectral pixels in the input data separated by $2/R$ in $x$ (natural logarithm of wavelength), and signal-to-noise per pixel of 17 in the continuum.
+One is poorly sampled, with $M=171$ spectral pixels in the input data separated by $2/R$ in $x$ (natural logarithm of wavelength), and signal-to-noise per pixel of 18 in the continuum.
 The other is well sampled, with $M=340$ spectral pixels in the input data separated by $1/R$ in $x$, and signal-to-noise per pixel of 12 in the continuum.
 In each case there are $N=8$ epochs.
-The two complete data sets are shown in \figurename~\label{fig:data1} and \figurename~\ref{fig:data2}.
+The two complete data sets are shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
 \begin{figure}[t!]
@@
 \paragraph{Forward Model (tm)}
-HOGG: Give full list of input parameters for the Fiducial. NOTE THAT WE DON'T DO THE NOISE IVAR RIGHT.
+As our main experiment or result, we apply the Forward Model (tm) as described above in \sectionname~\ref{sec:method} to the input data shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
+In each case, the output pixel grid $x_\star$ is chosen to be well sampled, with a pixel spacing of $1/R$ in the (natural) logarithm of wavelength.
+In order to give the model maximum flexibility, the number $P$ of Fourier modes in the spectral model is set to be equal to the number of pixels $M_\star$ in the pixel representation.
+The log-wavelength scale $L$ of the Fourier basis was set to the pixel spacing times the number of pixels.
+We didn't employ any data weighting; we treated every epoch as identical in terms of inverse variance.
 
-The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
-Again, it is compared to the true spectral expectation model.
-We find HOGG WHAT?? HERE SAY ABSOLUTELY WHAT WE FIND.
 \begin{figure}[t!]
@@
 \end{figure}
+The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
+It is compared to the true spectral expectation model that was used to generate the data.
+Also shown in \figurename~\ref{fig:forward} is the residuals, Forward Model (tm) minus truth.
+The residuals look stationary, approximately Gaussian, and uncorrelated (more on this below).
 
@@
 
-An example of a poorly-sampled spectrum interpolated to the rest frame is shown in \figurename~\ref{fig:interpolated}.
-The interpolation introduces substantial ringing artifacts in the spectrum (compare to \figurename~\ref{fig:data1}).
-It also grows the bad-pixel mask around the bad pixel.
 \begin{figure}[t!]
@@
     \end{center}
-    \caption{HOGG: WRITE CAPTION.\label{fig:interpolated}}
+    \caption{As part of Standard Practice (tm), it is necessary to interpolate the individual epoch spectra before averaging them.
+This shows one epoch of the poorly sampled data, interpolated to the output $x_\star$ pixel grid in the rest frame.
+The interpolated spectrum is compared to the true spectral expectation employed to generate it.
+Note the ringing induced by the interpolation.
+Also shown is the censored data, near an interpolated bad pixel and at the edge of the domain.
+Compare to the bottom panels of \figurename~\ref{fig:data1}.\label{fig:interpolated}}
     \end{mdframed}
 \end{figure}
+An example of a poorly-sampled spectrum interpolated to the rest frame is shown in \figurename~\ref{fig:interpolated}.
+The interpolation introduces substantial ringing artifacts in the spectrum (compare to \figurename~\ref{fig:data1}).
+It also grows the bad-pixel mask around the bad pixel.
 
-We show the final result of Standard Practice (tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
-We also show the comparison with the true spectral expectation model that was employed to make the fake data.
-We find HOGG WHAT?? CORRELATIONS; WORSE RESIDUALS NEAR SPECTRAL FEATURES (UH OH).
 \begin{figure}[t!]
@@
 \end{figure}
-
-HOGG: HERE SAY WHAT WE FIND RELATIVELY between the Forward Model (tm) and Standard Practice (tm).
+We show the final result of Standard Practice (tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
+We also show the comparison with the true spectral expectation model that was employed to make the fake data.
+The residuals look correlated (there is some ringing), and the residuals are larger near strong spectral features, especially in the case of the poorly sampled input data.
 
@@
     \end{center}
-    \caption{HOGG: Something.}
-    \label{fig:noise}
+    \caption{The empirical noise variance in the output (combined) spectra, and pixel-to-pixel covariances, as a function of pixel lag (pixel offset).
+The variances were estimated by performing the experiments repeated 64 times, with the same spectral expectations and sampling but unique noise draws.
+The Forward Model (tm) has slightly larger variance per pixel at zero lag, but vanishing pixel-to-pixel correlations.
+In contrast, the Standard Practice (tm) leads to correlated pixel values, with correlations extending out to many pixels.\label{fig:noise}}
     \end{mdframed}

### 2022-10-24 paper/main.tex
Updates from Overleaf

@@
     \end{center}
-    \caption{The top 8 panels show the $N=8$ epochs of the poorly sampled input data, with $N=8$, $M=171$, and signal-to-noise per pixel of 17 in the continuum.
+    \caption{The top 8 panels show the $N=8$ epochs of the poorly sampled input data, with $N=8$, $M=171$, and signal-to-noise per pixel of 18 in the continuum.
 The bad pixels are indicated with grey bars.

### 2022-11-01 paper/main.tex
Intro text start

@@
 \addtolength{\textheight}{0.6in}
-% \renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
+\setlength{\parindent}{3.5ex}
+\renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
 
@@
 
+It would be wrong to say that the practice of astronomy is, basically, \emph{staring at the sky}.
+That said, staring at the sky is a big part of what we do.
+And the way we do it, usually, is this:
+We take many individual images of our target with exposure times that are long enough to make read noise irrelevant (if possible) while at the same time short enough to avoid saturation (of what we care about).
+We then shift and average (maybe median?) those images to get the highest possible signal-to-noise on our target.
+This, fundamentally, is the technology underlying the \textsl{Hubble Deep Field} (\citealt{hdf}),
+the \textsl{APOGEE} spectroscopic survey (\citealt{apogee}),
+and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.
+
+Is it astronomical blasphemy to say that this practice is \emph{wrong?}
+It is wrong because minimum-variance unbiased estimators are (often) maximum-likelihood estimators (\citealt{something}).
+The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given noisy data.
+Noisy data are combined to make a measurement by building and optimizing a likelihood, not shifting and stacking!
+Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}.
+But it is.
+It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
+
+Our view is that image combination should be seen as a problem in statistical estimation.
+Let's find the best estimator for the combined image we can find, and let's deliver that (along with an uncertainty estimate).
+
 Things to emphasize here and below:
@@
 \end{itemize}
+
+Most important: Say here why you should never want to do this at all!
 
@@
 Matt Daunt (NYU),
+Dustin Lang (Perimeter),
 Adrian Price-Whelan (Flatiron),

### 2022-11-02 paper/main.tex
intro text

@@
 Is it astronomical blasphemy to say that this practice is \emph{wrong?}
-It is wrong because minimum-variance unbiased estimators are (often) maximum-likelihood estimators (\citealt{something}).
-The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given noisy data.
-Noisy data are combined to make a measurement by building and optimizing a likelihood, not shifting and stacking!
-Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}.
-But it is.
+It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}).
+The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
+Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
+Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator},
+but it is.
 It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
 
-Our view is that image combination should be seen as a problem in statistical estimation.
-Let's find the best estimator for the combined image we can find, and let's deliver that (along with an uncertainty estimate).
+We are in a period in astrophysics in which exceedingly precise measurements are expected, in both imaging and spectroscopy projects.
+In spectroscopy, in particular, high signal-to-noise spectral measurements are expected to obtain surface-abundance measurements with precisions a few percent, and radial-velocity measurements with precisions of up to one ten-thousandth of a \emph{pixel} (in terms of the Doppler shift).
+Spectral combination methods that distort absorption (or emission) lines will translate into biases, errors, or increased variance in these measurements.
+Spectral combination methods that correlate the noise in the outputs will translate into increased variance, and necessitate more sophisticated downstream measurement techniques (which don't really exist at present).
+
+HOGG: Of course if your GOAL IS RV you should DO SOMETHING DIFFERENT.
+
+One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
+It isn't really a ``model'' for the data; it is more like a ``summary'' of the data.
+When we switch to a statistical-inference framework, we have to choose some form or basis for expressing the combined image.
+In what follows we use a maximally parameterized model, in the sense that we choose an expression for the functions or function space that has as many parameters as the output image has pixels.
+That means that the approach we take does not get its power from the choice of parameterized functions for the output:
+Indeed it has just as much freedom as the shifting-and-stacking standard practice.
+The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation.
 
@@
 \begin{itemize}
-\item The input--output relationship of this modeling approach is identical to the input--output relationship of the shift-and-coadd approach. It is just better.
-\item People have a lot of hangups about the word ``model''.
+\item Maybe move this to the method section?
+People have a lot of hangups about the word ``model''.
 We have to use another word, or be very very careful how we use this word.
 For instance, some people I talked to about this thought I meant a physical model of the spectrum starting with atomic physics!
-\item We have to explain why and how we are better than other approaches in terms that people care about (like radial-velocity precision and abundances etc).
 \item All this is even more important in ill-sampled contexts than well-sampled contexts. See, for example, \textsl{APOGEE} raw data!
 \item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
-\item This model is a sub-model of \textsl{wobble} \cite{wobble}.
+\item This problem is a sub-problem of \textsl{wobble} \cite{wobble}.
+\item The problem we solve here is the easy case.
+In the hard case, there is a slit, and the object is extended, and different observations are at different slit angles, different slit positions, and with different atmospheric seeing.
+That problem is a hard problem!
+But it probably solves the same way.
+Say a bit here and more in the discussion.
 \end{itemize}

### 2022-11-02 paper/main.tex
comments and add text to section 1

--- 2022-11-01 paper/main.tex
@@
 \affiliation{School of Physics \& Astronomy, Monash University}
-\affiliation{Faculty of Information Technology, Monash University}
 
@@
 We take many individual images of our target with exposure times that are long enough to make read noise irrelevant (if possible) while at the same time short enough to avoid saturation (of what we care about).
-We then shift and average (maybe median?) those images to get the highest possible signal-to-noise on our target.
+We then shift and average sdfa(maybe median?) those images to get the highest possible signal-to-noise on our target.
 This, fundamentally, is the technology underlying the \textsl{Hubble Deep Field} (\citealt{hdf}),
@@
 The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given noisy data.
-Noisy data are combined to make a measurement by building and optimizing a likelihood, not shifting and stacking!
+Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking!
 Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}.
@@
 Let's find the best estimator for the combined image we can find, and let's deliver that (along with an uncertainty estimate).
+The data are never manipulated when using an estimator, but the relationship between the per epoch observations (inputs) and `combined image' or `combined spectrum' (outputs) is identical to the input--output relationship of a shift-and-coadd approach.
+Unlike the shift-and-coadd approach, an estimator is insensitive to under- or poorly-sampled data (e.g., \textsl{APOGEE} raw data).
+As we will show, an estimator also yields a more faithful representation\footnote{
+    An estimator needs a \emph{model} to produce a combined spectrum, but we use the word \emph{representation} to avoid overloading the word `model', and to indicate that we will not construct a physical model of a spectrum from atomic physics.
+    We only need a flexible representation!
+}
+of the underlying \emph{true} spectrum, critically important for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
 

### 2022-11-02 paper/main.tex
add bit-flag mask concept and minor edits to other assumptions

@@
 \newcommand{\kmps}{\unit{km\,s^{-1}}}
+
+% notes
 
@@
 Each of these observations $i$ (with $1\leq i\leq n$) produces a one-dimensional observed (noisy) image $y_i$.
-The $y_i$ can be thought of as $m_i$-vectors in what follows, where $m_i$ is the number of pixels in the spectrum.
+The $y_i$ can be thought of as $m_i$-vectors in what follows, where $m_i$ is the number of pixels in the $i$-th spectrum.
 The different observations are made at different relative velocities (star minus spectrograph) such that the different images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
@@
 We will begin by assuming that (from some external information) we know these shift operators very accurately.
-At the end we may (do we?) return to the question of how those shift operators are determined.
+At the end we may (do we?) return to the question of how those shift operators are determined. ANDY: SMASH! Out of scope!
 
@@
 This mask has value 1 in the locations in all good pixels, and 0 in the locations of all bad pixels.
+
+\paragraph{Bit-flag masks (or is it bit-masks? bit-pixel masks?)}
+A bit-flag mask is often used to encode meta data about an observation that otherwise cannot be represented by the spectrum $y_i$ or its asssociated variance tensor $C_i$ (e.g., poorly described line spread function), and we often want to propagate this meta data to appropriate pixels in the combined spectrum.
+These bit-flag masks are usually encoded as integers with up to $p$ potential flags, allowing for multiple flags to be encoded as a single value $\sum^{P}_{p=0} 2^p$. 
+This bit-flag mask can be separated into $p$ masks where each has value 1 in the locations where the flag is marked, and 0 in the locations where it is not.
 

### 2022-11-02 paper/main.tex
add final comments to concepts/assumptions

--- 2022-11-02 paper/main.tex
@@
 \shorttitle{don't interpolate your data!}
+% Or...
+%\shorttitle{Binning is sinning!}
 \shortauthors{hogg \& casey}
@@
 
+\paragraph{Bit-flag masks}
+A bit-flag mask is often used to store meta data about specific pixels in an observation that otherwise cannot be encoded by the spectrum $y_i$ or its asssociated variance tensor $C_i$ (e.g., poorly described line spread function).
+We often want to propagate this meta data to appropriate pixels in the combined spectrum. 
+These bit-flag masks are usually encoded as integers with up to $F$ potential flags, allowing for all (or a subset of) flags to be encoded as a single value $\sum^{F}_{f=0} 2^f$. 
+This bit-flag mask can be separated into $F$ masks where each has value 1 in the locations where the flag is marked, and 0 in the locations where it is not.
+
 \paragraph{Data gaps}
@@
 In general it depends on the slit or fiber width, the spectrograph resolution and optics, instrument focus, and the properties of the detector.
-In what follows we will not build any model of any of this.
+In what follows we will not build a model of any of this.
 Instead we will assume that there is a relatively constant pixel-convolved line-spread function (PCLSF), constant both in time, and constant with respect to the shifts $\Delta x_i$ we see in our data set.
@@
 If we were working on two-dimensional images, the model would have to be a representation of a continous two-dimensional function.
+
+ANDY: discuss with Hogg and check earlier language on spectral representation. Could introduce it earlier as continuous spectral model.
 
@@
 Again, the assumption is that the different calibrated or processed observations are of the same underlying signal.
+
+ANDY: Should we mention something about if it is not telluric corrected, and you have many epochs of a source that has perceived radial velocity variations, then you want something like wobble?
 

### 2022-11-02 paper/main.tex
up to discussion; add ack for holtz. need to discuss discussion (meta meta) with hogg

@@
 Importantly, the bad data get zeros in the \emph{inverse} covariance matrix, not the covariance matrix!
+We will propagate the meta data from the bit-flag mask later.
 
@@
 Now for the method:
-We stack all of the observations $y_i$ into one huge $M$-pixel list $Y$, where $M=\sum_{i=1}^N M_i$.
+% ANDY: this used to say "We stack" and it triggered me because I just read 7 pages about how we shouldn't "stack"!
+We append all of the observations $y_i$ into one huge $M$-pixel list $Y$, where $M=\sum_{i=1}^N M_i$.
 We make a $M\times M$ block-diagonal matrix $C^{-1}$ from all the individual inverse variance matrices $C_i^{-1}$.
@@
 \end{align}
-which is the maximum-likelihood or minimum-chi-squared solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \cite{hoggfitting}).
-And now if we choose a final (uniform, say) output pixel grid $x_\star$ in the rest-frame (unshifted) wavelength space, the combined (output, maybe?) spectrum is just
+which is the maximum-likelihood or minimum-$\chi^2$ solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \cite{hoggfitting}).
+And now if we choose a final (uniform, say) output pixel grid $x_\star$ in the rest-frame (unshifted) wavelength space, the combined spectrum is just
 \begin{align}
@@
 
-In the end, most users want the combined spectrum $y_\star$ (plus the pixel positions $x_\star$ and uncertainty information).
+In the end, most users want the combined spectrum $y_\star$ (and perhaps the pixel positions $x_\star$, uncertainty information, and propagated meta data).
 Some users will additionally want (or want to reconstruct) the parameters $\hat\theta$ and the basis functions $g_j(x)$.
@@
 Don't use \texttt{inv()}, use \texttt{solve()} or \texttt{lstsq()}.
-Check your condition numbers before you go.
+Check your condition numbers.
 If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra; research those.
-If you use the nufft, order your frequencies in the way the software wants, and deal with the complex numbers.
+If you use the non-uniform fast Fourier transform, order your frequencies in the way the software wants, and deal with the complex numbers.
 Specific choice of $L$ and $P$ and $M_\star$ in the Fourier case.
@@
 The spectral expectation model is built on the assumption of a normalized (unit) continuum and unresolved spectral lines with a Gaussian line-spread function (LSF) with one-sigma width $1/R$, where $R=135\,000$ is the spectrograph resolution).
+
+ANDY: continuum-rectified (unit-less) ?
 
@@
 The bottom two panels show the same but for the well sampled input data.
-In each panel, the red line shows the true model used to generate the data.\label{fig:forward}}
+In each panel, the red line shows the true model used to generate the data.\label{fig:forward} ANDY: THIS IS BEAUTIFUL}
     \end{mdframed}
@@
 In general this substantially grows the binary mask, and reduces the amount of data available for combination.
+The non-conservative approach is just as bad: if we restrict the number (or fraction) of pixels in the combined spectrum to equal those in the individual epochs, we risk un-masking genuinely bad pixels while growing the binary mask elsewhere!
 
@@
 It is a pleasure to thank
+Jon Holtzman (NMSU), 
 Matt Daunt (NYU),

### 2022-11-03 paper/main.tex
merge changes with hogg main

@@
 Is it astronomical blasphemy to say that this practice is \emph{wrong?}
-It is wrong because minimum-variance unbiased estimators are (often) maximum-likelihood estimators (\citealt{something}).
-The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given noisy data.
-Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking!
-Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}.
-But it is.
+It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}).
+The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
+Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
+Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator},
+but it is.
 It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
 
-Our view is that image combination should be seen as a problem in statistical estimation.
-Let's find the best estimator for the combined image we can find, and let's deliver that (along with an uncertainty estimate).
+We are in a period in astrophysics in which exceedingly precise measurements are expected, in both imaging and spectroscopy projects.
+In spectroscopy, in particular, high signal-to-noise spectral measurements are expected to obtain surface-abundance measurements with precisions a few percent, and radial-velocity measurements with precisions of up to one ten-thousandth of a \emph{pixel} (in terms of the Doppler shift).
+Spectral combination methods that distort absorption (or emission) lines will translate into biases, errors, or increased variance in these measurements.
+Spectral combination methods that correlate the noise in the outputs will translate into increased variance, and necessitate more sophisticated downstream measurement techniques (which don't really exist at present).
+
+HOGG: Of course if your GOAL IS RV you should DO SOMETHING DIFFERENT.
+
+One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
+It isn't really a ``model'' for the data; it is more like a ``summary'' of the data.
+When we switch to a statistical-inference framework, we have to choose some form or basis for expressing the combined image.
+In what follows we use a maximally parameterized model, in the sense that we choose an expression for the functions or function space that has as many parameters as the output image has pixels.
+That means that the approach we take does not get its power from the choice of parameterized functions for the output:
+Indeed it has just as much freedom as the shifting-and-stacking standard practice.
+The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation.
+
 The data are never manipulated when using an estimator, but the relationship between the per epoch observations (inputs) and `combined image' or `combined spectrum' (outputs) is identical to the input--output relationship of a shift-and-coadd approach.
@@
 \begin{itemize}
-\item The input--output relationship of this modeling approach is identical to the input--output relationship of the shift-and-coadd approach. It is just better.
-\item People have a lot of hangups about the word ``model''.
+    \item Maybe move this to the method section?
+People have a lot of hangups about the word ``model''.
 We have to use another word, or be very very careful how we use this word.
 For instance, some people I talked to about this thought I meant a physical model of the spectrum starting with atomic physics!
-\item We have to explain why and how we are better than other approaches in terms that people care about (like radial-velocity precision and abundances etc).
     \item All this is even more important in ill-sampled contexts than well-sampled contexts. See, for example, \textsl{APOGEE} raw data!
 \item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
-\item This model is a sub-model of \textsl{wobble} \cite{wobble}.
+\item This problem is a sub-problem of \textsl{wobble} \cite{wobble}.
+\item The problem we solve here is the easy case.
+In the hard case, there is a slit, and the object is extended, and different observations are at different slit angles, different slit positions, and with different atmospheric seeing.
+That problem is a hard problem!
+But it probably solves the same way.
+Say a bit here and more in the discussion.
 \end{itemize}

### 2022-12-04 paper/main.tex
discussion text dump

@@
 % -----------
+% - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
 % - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code?
@@
 % - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
-% - Copy in a figure macro from Hogg & Villar and use it.
+% - Audit for its / it's. Hogg has issues with this.
 % - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.
@@
 % page and document setup
+\renewcommand{\twocolumngrid}{}
 \addtolength{\topmargin}{-0.35in}
@@
 % text macros
-\shorttitle{don't interpolate your data!}
-% Or...
-%\shorttitle{Binning is sinning!}
+\shorttitle{don't interpolate your data}
 \shortauthors{hogg \& casey}
+\newcommand{\documentname}{\textsl{Article}}
 \newcommand{\sectionname}{Section}
@@
 Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean.
-The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation.
+The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
 The benefits of a modeling approach are myriad:
@@
 We take many individual images of our target with exposure times that are long enough to make read noise irrelevant (if possible) while at the same time short enough to avoid saturation (of what we care about).
-We then shift and average sdfa(maybe median?) those images to get the highest possible signal-to-noise on our target.
+We then shift and average (maybe median?) those images to get the highest possible signal-to-noise on our target.
 This, fundamentally, is the technology underlying the \textsl{Hubble Deep Field} (\citealt{hdf}),
@@
 
-HOGG: Of course if your GOAL IS RV you should DO SOMETHING DIFFERENT.
-
 One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
@@
 
-The data are never manipulated when using an estimator, but the relationship between the per epoch observations (inputs) and `combined image' or `combined spectrum' (outputs) is identical to the input--output relationship of a shift-and-coadd approach.
-Unlike the shift-and-coadd approach, an estimator is insensitive to under- or poorly-sampled data (e.g., \textsl{APOGEE} raw data).
-As we will show, an estimator also yields a more faithful representation\footnote{
-    An estimator needs a \emph{model} to produce a combined spectrum, but we use the word \emph{representation} to avoid overloading the word `model', and to indicate that we will not construct a physical model of a spectrum from atomic physics.
-    We only need a flexible representation!
-}
-of the underlying \emph{true} spectrum, critically important for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
+The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
+We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates.
+That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
+Unlike the shift-and-coadd approach, the estimator we will deliver will be insensitive to under- or poorly-sampled data, which is not uncommon when astronomical seeing accidentally gets excellent, or in spectrographs that have limited detector pixels (for example, \citealt{apogeehardware}).
+As we will show, an estimator also often yields a more faithful estimate of the \emph{true} spectrum, and almost always better noise properties.
+Both of these results make our results better for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
+But importantly, we aren't proposing a change to the \emph{format} of the input or outputs of spectral or image combination.
 
-Things to emphasize here and below:
-\begin{itemize}
-    \item Maybe move this to the method section?
-People have a lot of hangups about the word ``model''.
-We have to use another word, or be very very careful how we use this word.
-For instance, some people I talked to about this thought I meant a physical model of the spectrum starting with atomic physics!
-    \item All this is even more important in ill-sampled contexts than well-sampled contexts. See, for example, \textsl{APOGEE} raw data!
-\item We should choose a rigid terminology and stick to it here. HOGG is currently liking ``combined spectrum'' for the model expectation.
-\item This problem is a sub-problem of \textsl{wobble} \cite{wobble}.
-\item The problem we solve here is the easy case.
-In the hard case, there is a slit, and the object is extended, and different observations are at different slit angles, different slit positions, and with different atmospheric seeing.
-That problem is a hard problem!
-But it probably solves the same way.
-Say a bit here and more in the discussion.
-\end{itemize}
+The ideas here aren't novel; many projects have avoided data interpolation previously.
+For one example, the \textsl{COMBO-17} survey (\citealt{combo17}) took many images of the sky in many bands, and did not co-add the data prior to making measurements, in part to avoid mixing different point-spread functions and pixel samplings.
+For another, what we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models.
+Plus it learns (it isn't told) the epoch-to-epoch spectral shifts.
+There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{froe}, \citealt{sp}):
+Optimal extraction is an unfortunate name for the process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
+One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is a better method that moves this spectral combination method directly into the two-dimensional data.
+We hope for a future in which one-dimensional spectra are never extracted from two-dimensional images; in this future the two-dimensional data are the sacred data, not the extractions.
 
-Most important: Say here why you should never want to do this at all!
+In real applications, the spectra we extract are often of variable sources, or taken at different position angles on extended objects, such that we don't expect each epoch to be identical.
+Similarly, the \textsl{LSST} data are being taken the way they because the sky (and point-spread function) will vary.
+That means that we are only considering here the very easiest case.
+However, we note that all standard shift-and-coadd methods make these non-variable assumptions implicitly.
+Part of our contribution in this \documentname{} is to make implicit assumptions explicit, in \sectionname~\ref{sec:assumptions}.
+
+HOGG: Most important: Say here why you should never want to do this at all!
+Comment on a spectrum being too low in s/n to measure a Na line...!!
 
@@
 That is, if there were a true spectrum $\tilde{y}$, each individual noisy image $y_i$ would be related to that true spectrum by a shift operator acting on that true spectrum, plus noise.
-We will begin by assuming that (from some external information) we know these shift operators very accurately.
-At the end we may (do we?) return to the question of how those shift operators are determined. ANDY: SMASH! Out of scope!
+We will assume here that (from some external information) we know these shift operators very accurately.
+This is often not true, of course; we will return to this point in the discussion.
 
@@
 
-HOGG: Something about the LSF being different in different exposures!
+In many cases the LSF or PSF is actually substantially different in different exposures.
+This leads to spectral (or image) variations, and violate the assumptions of this model (and also the implicit assumptions underlying standard practice).
+We won't address this kind of variability here, except in some discussion at the end.
 
@@
 This keeps the band-limited part of the spectroscopic signal well-sampled in the data.
-There are spectrograph that don't have pixel scales this fine \cite{apogeehardware}; in these cases what we are proposing here is even more important than it is in the well-sampled case.
+There are spectrograph that don't have pixel scales this fine (for example, \textsl{APOGEE} \citealt{apogee}); in these cases what we are proposing here is \emph{even more important} than it is in the well-sampled case.
+This will become obvious in the experiments below.
 
 \paragraph{Non-uniform fast Fourier transform}
-In the special case that the spectral model is a Fourier series, there are very useful and fast tools for interpolation, including especially the non-uniform fast Fourier transform (for example, \cite{finufft}).
+In the special case that the spectral model is a Fourier series, there are very useful and fast tools for interpolation, including especially the non-uniform fast Fourier transform (for example, \citealt{finufft}).
 
@@
 Again, the assumption is that the different calibrated or processed observations are of the same underlying signal.
-
-ANDY: Should we mention something about if it is not telluric corrected, and you have many epochs of a source that has perceived radial velocity variations, then you want something like wobble?
+If there are multiple signals participating, the model would have to expand into something like \textsl{wobble} (\citealt{wobble}).
 
@@
 Now for the method:
-% ANDY: this used to say "We stack" and it triggered me because I just read 7 pages about how we shouldn't "stack"!
 We append all of the observations $y_i$ into one huge $M$-pixel list $Y$, where $M=\sum_{i=1}^N M_i$.
@@
 \end{align}
-which is the maximum-likelihood or minimum-$\chi^2$ solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \cite{hoggfitting}).
+which is the maximum-likelihood or minimum-$\chi^2$ solution for a linear weighted least squares problem or a linear model with Gaussian noise (see, for example, \citealt{fitting}).
 And now if we choose a final (uniform, say) output pixel grid $x_\star$ in the rest-frame (unshifted) wavelength space, the combined spectrum is just
@@
 But, \emph{very importantly}, this is only a correct usage when the input $C_i$ matrices really represented the true uncertainties going in.
-If you have any concerns about that, then it is way safer to get your uncertainty estimates from bootstrap or jackknife \cite{hoggflexible}.
+If you have any concerns about that, then it is way safer to get your uncertainty estimates from bootstrap or jackknife (\citealt{fittingflexible}).
 
@@
 
-ANDY: continuum-rectified (unit-less) ?
+ANDY: continuum-rectified (unit-less) ?  HOGG: THE LINE FORMATION MODEL DESCRIBED ABOVE IS NOT RIGHT.
 
@@
 The bottom two panels show the same but for the well sampled input data.
-In each panel, the red line shows the true model used to generate the data.\label{fig:forward} ANDY: THIS IS BEAUTIFUL}
+In each panel, the red line shows the true model used to generate the data.\label{fig:forward} ANDY: THIS IS BEAUTIFUL.
+HOGG SAY: I KNOW!}
     \end{mdframed}
@@
 
-What happens if you don't know your shifts?
+In the Introduction, we were agnostic about whether we were talking about imaging or spectroscopic data.
+In the end, our examples are spectroscopic.
+What differences are there between these cases?
+The main difference is that there are many more ``shift'' parameters for imaging than spectroscopy:
+Two spectra (according to our assumptions) differ only (or primarily) in terms of a one-dimensional Doppler shift with respect to pixels that are located in one dimension (the wavelength direction).
+Imaging epochs, on the other hand, can differ in terms of two shifts and a rotation, or (equivalently) three Euler angles.
+And, since every pixel has its own two-dimensional position, the problem of locating pixels in an image is generally harder than the problem of locating pixels in an extracted spectrum.
+That is, the imaging case is generally more difficult.
+But it is not different, conceptually, from what is presented above.
 
-What happens if you have multiple shifting signals?
-\texttt{wobble} \cite{wobble} is a solution to one version of that problem.
+Everything presented above depends on a set of strong assumptions, listed in \sectionname~\ref{sec:assumptions}.
+Perhaps the strongest assumption is that the spectrum or image is not a function of time.
+That is, each epoch spectrum or image is a sampling of the same underlying, true signal.
+This assumption can be wrong for a myriad of reasons:
+The source can vary, the instrument can vary, there can be variable foregrounds or backgrounds, and there can be mistakes or variations in calibration or data processing.
+All of these things happen in real data sets.
 
-What happens if your data aren't continuum normalized properly?
+There are (at least) three attitudes for an investigator to take in the face of variations in the data, from epoch to epoch or exposure to exposure:
+The first is that the procedures we design are designed to deliver the exposure (or signal-to-noise-squared) weighted mean of the observed epochs.
+That is, the fact that the data are varying does not mean that the methodology needs to be modified; without modification, the procedures given above---both the Forward Model (tm) and the Standard Practice (tm)---will produce something like the mean spectrum.
 
-We assumed no sky subtraction necessary. What gives?
+The second attitude to take---and the necessary attitude for many projects---is that the variations must be accounted for, by complexifying the forward model.
+For example, if there is a varying line-spread function between epochs (as there is for the \textsl{APOGEE} spectrograph, because the LSF is a function of the slit position of the optical fiber), then in addition to a shift at each epoch there is also an LSF adjustment at each epoch.
+The LSF adjustment is then applied to the spectral model before it is compared to each epoch of data, just as the shift is applied in the current methodological scope.
+Similar model adustments can account for intrinsic spectral variations, additive or multiplicative nuisance signals in the data (like backgrounds or telluric absorptions), and mistakes in data pre-processing (like continuum modeling errors).
 
-How do you modify to deal with LSF variations (like we have in \textsl{SDSS-V APOGEE}?
-What if you don't know the LSF differences?
-Difference between modeling LSF changes and modeling LSFs. The former is good, the latter is bad.
+One side note about epoch-to-epoch variations in instrument or data-processing parameters:
+It is usually a good idea to model not the full instrument or data properties (as does \citealt{sp}) if that can be avoided, but rather just the \emph{differences} in instrument or data properties from one epoch to the next, taking those differences away from some fiducial data or instrument state.
+In the case of the LSF, for example, that obviates the need to model the infinite-resolution spectrum; the spectrum only needs to be modeled at fiducial resolution and convolved to other resolutions with difference kernels.
 
-Discuss time-variable inputs, and how you would modify to deal with that.
-One way involves modeling the time variability with some process.
-Another way involves ignoring it and just getting the average anyway!
+Although \emph{conceptually} changes to account for data variability aren't that big, \emph{in practice} they can be very hard to implement.
+For one, the method we have presented assumes that we \emph{know} the individual epoch shifts \textsl{a priori}, and all new epoch-to-epoch adjustments would also depend on per-epoch housekeeping data or parameters that might, in practice, be hard to know.
+That brings us to the third attitude that the investigator can take.
+What's presented above can be seen as a model for the \emph{mean} of the process that generates individual-epoch spectra; that is the first term in an expansion of moments.
+The next term would be the \emph{variance} of the process.
+That is, a more sophisticated model could have not just $P$ parameters for the mean spectrum, but additionally $P'$ parameters for a variance model.
+If the variance is low dimensional, or sparse, or compact in important ways, there should be very effective models of this form.
+For example, we could imagine forward-modeling generalizations of some of sparse methods (for example, \citealt{candes}) for applications in which the variability is localized in time or space; or we could imagine generalizations of matrix factorizations (for example, \citealt{hmf}) for applications in which variability is low dimensional (the \textsl{wobble} method has some such functionality; \citealt{wobble}).
+
+All the above considerations of variability bring us to another of our strong assumptions:
+We have assumed that we know the shifts $\Delta x_i$ exactly.
+In practice, we rarely do, or maybe more accurately: We rarely know them as well as we could know them, in principle.
+When the shifts $\Delta x_i$ are not known, or not known accurately enough, we recommend a simultaneous-fitting procedure, in which the shifts are learned at the same time as the combined spectrum is estimated or inferred.
+This can be achieved through a challenging nonlinear optimization, or through an iteration of linearized optimizations, alternating improvements to the combined spectrum and improvements to the shifts.
+This is, more or less, how the \textsl{wobble} system works (\citealt{wobble}).
+
+We assumed no sky subtraction necessary. No tellurics. What gives if you have multiple shifting signals. Causal structure. etc.
 
 What about non-Gaussian noise?
-
-What about data from different instruments with different LSFs / PSFs? What about data with different pixel scales? What about data with unknown wavelength calibration?
 
@@
 \paragraph{Software}
-\texttt{numpy} \cite{numpy} ---
-\texttt{finufft} \cite{finufft} ---
-\texttt{matplotlib} \cite{matplotlib}.
+\texttt{numpy} \citep{numpy} ---
+\texttt{matplotlib} \citep{matplotlib}.
 
@@
 It is a pleasure to thank
+Mike Blanton (NYU),
+Matt Daunt (NYU),
 Jon Holtzman (NMSU), 
-Matt Daunt (NYU),
 Dustin Lang (Perimeter),
@@
 and the Astronomical Data Group at the Flatiron Institute,
-for valuable discussions.
+for valuable discussions and input.
+
+\begin{thebibliography}{dummy}
+\bibitem[Barnett et al.(2019)]{finufft}
+Barnett, A. H., Magland, J. F., \& af~Klinteberg, L., 2019, SIAM J. Sci. Comput. 41(5), C479
+\bibitem[Bedell et al.(2019)]{wobble} Bedell, M., Hogg, D.~W., Foreman-Mackey, D., et al.\ 2019, \aj, 158, 164. doi:10.3847/1538-3881/ab40a7
+\bibitem[Bolton \& Schlegel(2010)]{sp} Bolton, A.~S. \& Schlegel, D.~J.\ 2010, \pasp, 122, 248. doi:10.1086/651008
+\bibitem[Harris et al.(2020)]{numpy} Harris, C.~R., Millman, K.~J., van der Walt, S.~J., et al.\ 2020, \nat, 585, 357. doi:10.1038/s41586-020-2649-2
+\bibitem[Hewett et al.(1985)]{oe} Hewett, P.~C., Irwin, M.~J., Bunclark, P., et al.\ 1985, \mnras, 213, 971. doi:10.1093/mnras/213.4.
+\bibitem[Hogg et al.(2010)]{fitting} Hogg, D.~W., Bovy, J., \& Lang, D.\ 2010, arXiv:1008.4686
+\bibitem[Hogg \& Villar(2021)]{fittingflexible} Hogg, D.~W. \& Villar, S.\ 2021, \pasp, 133, 093001. doi:10.1088/1538-3873/ac20ac
+\bibitem[Hunter(2007)]{matplotlib} Hunter, J.~D.\ 2007, Computing in Science and Engineering, 9, 90. doi:10.1109/MCSE.2007.55
+\bibitem[Ivezi{\'c} et al.(2019)]{lsst} Ivezi{\'c}, {\v{Z}}., Kahn, S.~M., Tyson, J.~A., et al.\ 2019, \apj, 873, 111. doi:10.3847/1538-4357/ab042c
+\bibitem[Majewski et al.(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2017, \aj, 154, 94. doi:10.3847/1538-3881/aa784d
+\bibitem[Williams et al.(1996)]{hdf} Williams, R.~E., Blacker, B., Dickinson, M., et al.\ 1996, \aj, 112, 1335. doi:10.1086/118105
+\bibitem[Wolf et al.(2003)]{combo17} Wolf, C., Meisenheimer, K., Rix, H.-W., et al.\ 2003, \aap, 401, 73. doi:10.1051/0004-6361:20021513
+\bibitem[Zechmeister et al.(2014)]{froe} Zechmeister, M., Anglada-Escud{\'e}, G., \& Reiners, A.\ 2014, \aap, 561, A59. doi:10.1051/0004-6361/201322746
+\end{thebibliography}
 

### 2023-09-21 code/sampling.py
update nomenclature to match paper

@@ -1,8 +1,8 @@
 import numpy as np
+import warnings
 from collections import OrderedDict
 from typing import List, Tuple, Optional
 from itertools import cycle
-
 
 def resample_spectrum(
     resample_wavelength: np.array,
@@ -11,46 +11,52 @@ def resample_spectrum(
     ivar: Optional[np.array] = None,
     flags: Optional[np.array] = None,
     mask: Optional[np.array] = None,
-    A: Optional[np.array] = None,
     L: Optional[int] = None,
     P: Optional[int] = None,
+    X_star: Optional[np.array] = None,
     min_resampled_flag_value: Optional[float] = 0.1,
+    Lambda: Optional[float] = 0.0,
+    rcond: Optional[float] = 1e-15,
     full_output: Optional[bool] = False,
 ) -> Tuple[np.array, np.array]:
     """
     Sample a spectrum on a wavelength array given a set of pixels recorded from one or many visits.
     
     :param resample_wavelength:
-        A ``N``-length array of wavelengths to sample the spectrum on.
+        A ``N``-length array of wavelengths to sample the spectrum on. In the paper, this is equivalent 
+        to the output-spectrum pixel grid $x_\star$.
 
     :param wavelength:
-        A ``D``-length array of wavelengths from individual visits.
+        A ``M``-length array of wavelengths from individual visits. In the paper, this is equivalent to 
+        the input-spectrum pixel grid $x_i$.
     
     :param flux:
-        A ``D``-length array of flux values from individual visits.
+        A ``M``-length array of flux values from individual visits. In the paper, this is equivalent to 
+        the observations $y_i$.
     
     :param ivar: [optional]
-        A ``D``-length array of inverse variance values from individual visits.
+        A ``M``-length array of inverse variance values from individual visits. In the paper, this is 
+        equivalent to the individual inverse variance matrices $C_i^{-1}$.
 
     :param flags: [optional]
-        A ``D``-length array of bitmask flags from individual visits.
+        A ``M``-length array of bitmask flags from individual visits.
     
     :param mask: [optional]
-        A ``D``-length array of boolean values indicating whether a pixel should be used or not in
+        A ``M``-length array of boolean values indicating whether a pixel should be used or not in
[…]

### 2023-09-21 setup.py
moving furniture

@@ -1,0 +1,8 @@
+from setuptools import setup, find_packages
+
+setup(
+    name="ndi",
+    version="0.0.1",
+    packages=find_packages(where="ndi"),
+    install_requires=["numpy"]
+)

### 2023-09-21 code/__init__.py
add test structure

@@ -1,0 +1,414 @@
+import numpy as np
+import warnings
+from collections import OrderedDict
+from typing import List, Tuple, Optional
+from itertools import cycle
+
+def resample_spectrum(
+    resample_wavelength: np.array,
+    wavelength: np.array,
+    flux: np.array,
+    ivar: Optional[np.array] = None,
+    flags: Optional[np.array] = None,
+    mask: Optional[np.array] = None,
+    L: Optional[int] = None,
+    P: Optional[int] = None,
+    X_star: Optional[np.array] = None,
+    min_resampled_flag_value: Optional[float] = 0.1,
+    Lambda: Optional[float] = 0.0,
+    rcond: Optional[float] = 1e-15,
+    full_output: Optional[bool] = False,
+) -> Tuple[np.array, np.array]:
+    """
+    Sample a spectrum on a wavelength array given a set of pixels recorded from one or many visits.
+    
+    :param resample_wavelength:
+        A ``N``-length array of wavelengths to sample the spectrum on. In the paper, this is equivalent 
+        to the output-spectrum pixel grid $x_\star$.
+
+    :param wavelength:
+        A ``M``-length array of wavelengths from individual visits. In the paper, this is equivalent to 
+        the input-spectrum pixel grid $x_i$.
+    
+    :param flux:
+        A ``M``-length array of flux values from individual visits. In the paper, this is equivalent to 
+        the observations $y_i$.
+    
+    :param ivar: [optional]
+        A ``M``-length array of inverse variance values from individual visits. In the paper, this is 
+        equivalent to the individual inverse variance matrices $C_i^{-1}$.
+
+    :param flags: [optional]
+        A ``M``-length array of bitmask flags from individual visits.
+    
+    :param mask: [optional]
+        A ``M``-length array of boolean values indicating whether a pixel should be used or not in
+        the resampling (`True` means mask the pixel, `False` means use the pixel). If `None` is
+        given then all pixels will be used. The `mask` is only relevant for sampling the flux and
+        inverse variance values, and not the flags.
+    
+    :param X_star: [optional]
+        The design matrix to use when solving for the combined spectrum. If you are resampling
[…]

### 2023-09-21 ndi/__init__.py
scaffolding for tests

@@ -1,414 +1,0 @@
-import numpy as np
-import warnings
-from collections import OrderedDict
-from typing import List, Tuple, Optional
-from itertools import cycle
-
-def resample_spectrum(
-    resample_wavelength: np.array,
-    wavelength: np.array,
-    flux: np.array,
-    ivar: Optional[np.array] = None,
-    flags: Optional[np.array] = None,
-    mask: Optional[np.array] = None,
-    L: Optional[int] = None,
-    P: Optional[int] = None,
-    X_star: Optional[np.array] = None,
-    min_resampled_flag_value: Optional[float] = 0.1,
-    Lambda: Optional[float] = 0.0,
-    rcond: Optional[float] = 1e-15,
-    full_output: Optional[bool] = False,
-) -> Tuple[np.array, np.array]:
-    """
-    Sample a spectrum on a wavelength array given a set of pixels recorded from one or many visits.
-    
-    :param resample_wavelength:
-        A ``N``-length array of wavelengths to sample the spectrum on. In the paper, this is equivalent 
-        to the output-spectrum pixel grid $x_\star$.
-
-    :param wavelength:
-        A ``M``-length array of wavelengths from individual visits. In the paper, this is equivalent to 
-        the input-spectrum pixel grid $x_i$.
-    
-    :param flux:
-        A ``M``-length array of flux values from individual visits. In the paper, this is equivalent to 
-        the observations $y_i$.
-    
-    :param ivar: [optional]
-        A ``M``-length array of inverse variance values from individual visits. In the paper, this is 
-        equivalent to the individual inverse variance matrices $C_i^{-1}$.
-
-    :param flags: [optional]
-        A ``M``-length array of bitmask flags from individual visits.
-    
-    :param mask: [optional]
-        A ``M``-length array of boolean values indicating whether a pixel should be used or not in
-        the resampling (`True` means mask the pixel, `False` means use the pixel). If `None` is
-        given then all pixels will be used. The `mask` is only relevant for sampling the flux and
-        inverse variance values, and not the flags.
-    
-    :param X_star: [optional]
-        The design matrix to use when solving for the combined spectrum. If you are resampling
[…]

### 2023-09-21 setup.py

@@ -3,6 +3,6 @@ from setuptools import setup, find_packages
 setup(
     name="ndi",
     version="0.0.1",
-    packages=find_packages(where="ndi"),
+    packages=find_packages(where="code"),
     install_requires=["numpy"]
 )

### 2023-09-21 tests/test_ndi.py

@@ -1,5 +1,139 @@
-def func(x):
-    return x + 1
+# imports and initialize seeds, fundamental constants
+import numpy as np
+import pylab as plt
+import scipy.interpolate as interp
+np.random.seed(17)
+c = 299792458. # m / s
+sqrt2pi = np.sqrt(2. * np.pi)
 
-def test_answer():
-    assert func(3) == 4
+# make the synthetic spectrum (spectral expectation), and also add noise
+
+def oned_gaussian(dxs, sigma):
+    return np.exp(-0.5 * dxs ** 2 / sigma ** 2) / (sqrt2pi * sigma)
+
+def true_spectrum(xs, doppler, lxs, ews, sigma):
+    """
+    """
+    return np.exp(-1. * np.sum(ews[None, :] *
+                               oned_gaussian(xs[:, None] - doppler
+                                             - lxs[None, :], sigma), axis=1))
+
+def ivar(ys, continuum_ivar):
+    return continuum_ivar / ys
+
+def noisy_true_spectrum(xs, doppler, continuum_ivar, line_xs, line_ews, sigma_x):
+    """
+    """
+    ys_true = true_spectrum(xs, doppler, line_xs, line_ews, sigma_x)
+    y_ivars = ivar(ys_true, continuum_ivar)
+    return  ys_true + np.random.normal(size=xs.shape) / np.sqrt(y_ivars), y_ivars
+
+def doppler_information(xs, doppler, continuum_ivar, dx):
+    """
+    # Bugs:
+    - Horrifying numerical derivative!
+    """
+    dys_dx = (true_spectrum(xs, doppler + dx)
+              - true_spectrum(xs, doppler - dx)) / (2. * dx)
+    y_ivars = ivar(true_spectrum(xs, doppler), continuum_ivar)
+    return np.sum(y_ivars * dys_dx ** 2)
+
+def badify(yy, badfrac):
+    """
+    Make bad-pixel masks and badify the bad pixels.
+    """
+    bady = 1. * yy
+    bs = (np.random.uniform(size=len(bady)) > badfrac).astype(int)
+    bs = np.minimum(bs, np.roll(bs, 1))
+    bs = np.minimum(bs, np.roll(bs, -1))
+    nbad = np.sum(bs < 0.5)
+    if nbad > 0:
+        bady[bs < 0.5] += 2. * np.random.uniform(size=nbad)
+    return bs, bady
+
[…]

### 2023-09-21 setup.py
tests require that forward model is better than standard practice

@@ -3,6 +3,7 @@ from setuptools import setup, find_packages
 setup(
     name="ndi",
     version="0.0.1",
-    packages=find_packages(where="code"),
+    package_dir={"": "python"},
+    packages=find_packages(where="python"),
     install_requires=["numpy"]
 )

### 2023-09-21 tests/test_ndi.py

@@ -1,10 +1,13 @@
 # imports and initialize seeds, fundamental constants
 import numpy as np
-import pylab as plt
 import scipy.interpolate as interp
+import matplotlib.pyplot as plt
+
 np.random.seed(17)
 c = 299792458. # m / s
 sqrt2pi = np.sqrt(2. * np.pi)
+
+from ndi import resample_spectrum
 
 # make the synthetic spectrum (spectral expectation), and also add noise
 
@@ -26,7 +29,7 @@ def noisy_true_spectrum(xs, doppler, continuum_ivar, line_xs, line_ews, sigma_x)
     """
     ys_true = true_spectrum(xs, doppler, line_xs, line_ews, sigma_x)
     y_ivars = ivar(ys_true, continuum_ivar)
-    return  ys_true + np.random.normal(size=xs.shape) / np.sqrt(y_ivars), y_ivars
+    return  (ys_true + np.random.normal(size=xs.shape) / np.sqrt(y_ivars), y_ivars)
 
 def doppler_information(xs, doppler, continuum_ivar, dx):
     """
@@ -51,7 +54,7 @@ def badify(yy, badfrac):
         bady[bs < 0.5] += 2. * np.random.uniform(size=nbad)
     return bs, bady
 
-def make_one_dataset(dx, SNR, x_min, x_max, line_xs, line_ews, sigma_x, badfrac, N=8):
+def make_one_dataset(dx, SNR, x_min, x_max, line_xs, line_ews, sigma_x, badfrac, xstar, N=8):
     # create true Doppler shifts on a sinusoid of epoch number
     Delta_xs = (3.e4 / c) * np.cos(np.arange(N) / 3.)
     # set the ivar
@@ -61,14 +64,23 @@ def make_one_dataset(dx, SNR, x_min, x_max, line_xs, line_ews, sigma_x, badfrac,
     ys = np.zeros((N, len(xs)))
     y_ivars = np.zeros_like(ys)
     bs = np.zeros_like(ys).astype(int)
+    y_true = true_spectrum(xstar, 0, line_xs, line_ews, sigma_x)
     for j in range(N):
         ys[j], y_ivars[j] = noisy_true_spectrum(xs, Delta_xs[j], continuum_ivar, line_xs, line_ews, sigma_x)
         bs[j], ys[j] = badify(ys[j], badfrac)
-    return xs, ys, y_ivars, bs, Delta_xs
+    return xs, ys, y_ivars, bs, Delta_xs, y_true
 
 
[…]

### 2023-09-21 paper/main.tex
intro edits

@@
 
-\title{\Huge Don't interpolate your data!}
+\title{\Huge Don't interpolate the data!}
+% ARC: I changed this to "Don't interpolate the data!"? because you can't like, own data, man.
+% (but you can change it back if you prefer)
 
@@
 \affiliation{School of Physics \& Astronomy, Monash University}
+\affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}
 
@@
 Is it astronomical blasphemy to say that this practice is \emph{wrong?}
-It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}).
+It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}). % HOGG
 The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
 Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
-Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator},
-but it is.
+Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
 It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
@@
 The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
-We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates.
+We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps metadata flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
 That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.

### 2023-09-21 paper/main.tex
try to distinguish from interpolation

@@
 
+The approach we propose differs from interpolation in a subtle way. 
+Interpolation could be described as computing new values at arbitrary positions based on an exising set of known data points. 
+Instead, here we propose to construct a forward model for the data, where we then evaluate that model on the given set of output positions (pixels).
+The \emph{output product} is the same form as what one would expect from an interpolation procedure (e.g., computed values along some array of positions), but the \emph{procedure} differs from interpolation.
+When interpolating data the basis functions are explicitly constructed from the data, whereas here we construct a forward model that is maximally consistent with the data, and evaluated at new positions.
+
 The ideas here aren't novel; many projects have avoided data interpolation previously.
@@
 There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{froe}, \citealt{sp}):
-Optimal extraction is an unfortunate name for the process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
+Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
 One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is a better method that moves this spectral combination method directly into the two-dimensional data.
@@
 
-HOGG: Most important: Say here why you should never want to do this at all!
-Comment on a spectrum being too low in s/n to measure a Na line...!!
+HOGG: Most important: Say here why you should never want to do this at all! % ARC: I think you've covered this by talking about the two-dimensional data.
+Comment on a spectrum being too low in s/n to measure a Na line...!! % ARC: I don't remember what the context is for this
 
@@
 \bibitem[Williams et al.(1996)]{hdf} Williams, R.~E., Blacker, B., Dickinson, M., et al.\ 1996, \aj, 112, 1335. doi:10.1086/118105
+\bibitem[Wilson et al.(2019)]{apogeehardware} Wilson, J.C., Hearty, F.R., Skrutskie, M.F., Majewski, S.R., Holtzman, J.A., Eisenstein, D., et al.\ 2019, \pasp, 131, 055001.
+doi:10.1088/1538-3873/ab0075.
 \bibitem[Wolf et al.(2003)]{combo17} Wolf, C., Meisenheimer, K., Rix, H.-W., et al.\ 2003, \aap, 401, 73. doi:10.1051/0004-6361:20021513

### 2023-09-21 paper/main.tex
smallest change i could think of that addresses the 'what about single epoch cases'
minor

@@
 The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
-We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps metadata flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
+We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
 That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
@@
 Interpolation could be described as computing new values at arbitrary positions based on an exising set of known data points. 
-Instead, here we propose to construct a forward model for the data, where we then evaluate that model on the given set of output positions (pixels).
+Instead, here we propose to construct a continuous spectral model for the data, where we then evaluate that model on the given set of output positions (pixels).
 The \emph{output product} is the same form as what one would expect from an interpolation procedure (e.g., computed values along some array of positions), but the \emph{procedure} differs from interpolation.
@@
 
-What we do here can apply to almost any astronomical signals, but we are going to specialize to astronomical spectra for specificity.
+What we do here can apply to almost any astronomical signals, but we are going to specialize to multi-epoch astronomical spectra for specificity.
 We are going to make many other assumptions too, all of which are discussed in this \sectionname.
@@
 The $y_i$ can be thought of as $m_i$-vectors in what follows, where $m_i$ is the number of pixels in the $i$-th spectrum.
-The different observations are made at different relative velocities (star minus spectrograph) such that the different images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
+The different observations are made at different relative velocities (star minus spectrograph) such that the images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
 
@@
 
-\paragraph{Bit-flag masks}
-A bit-flag mask is often used to store meta data about specific pixels in an observation that otherwise cannot be encoded by the spectrum $y_i$ or its asssociated variance tensor $C_i$ (e.g., poorly described line spread function).
+\paragraph{Bit-flags}
+A bit-flag is often used to store meta data about a pixel that otherwise cannot be encoded by the spectrum $y_i$ or its asssociated variance tensor $C_i$ (e.g., poorly described line spread function).
 We often want to propagate this meta data to appropriate pixels in the combined spectrum. 
@@
 
-ANDY: discuss with Hogg and check earlier language on spectral representation. Could introduce it earlier as continuous spectral model.
-
 \paragraph{Linear basis or mixture models}
@@
 This keeps the band-limited part of the spectroscopic signal well-sampled in the data.
-There are spectrograph that don't have pixel scales this fine (for example, \textsl{APOGEE} \citealt{apogee}); in these cases what we are proposing here is \emph{even more important} than it is in the well-sampled case.
+There are spectrograph that don't have pixel scales this fine (for example, \textsl{APOGEE}; \citealt{apogee}); in these cases what we are proposing here is \emph{even more important} than it is in the well-sampled case.
 This will become obvious in the experiments below.
@@
 If you use the non-uniform fast Fourier transform, order your frequencies in the way the software wants, and deal with the complex numbers.
-Specific choice of $L$ and $P$ and $M_\star$ in the Fourier case.
+Specify your choice of $L$ and $P$ and $M_\star$ in the Fourier case.
 
@@
 We demonstrate the effectiveness of the spectral combination method on artificial data.
-The spectral data were generated from a mean spectral expectation that is a continuum plus a randomly generated (not at all physical) line list, where each line equivalent width was set by a random draw from a power law.
+The spectral data were generated from a mean spectral expectation that is a continuum plus a randomly generated line list, where each line equivalent width was set by a random draw from a power law. 
+This line formation model is unphysical, but it is sufficient for our purposes in that it generates fake data that look like stellar spectra.
 Each spectral epoch was given a velocity (Doppler shift) by putting the epochs equally spaced on a sinusoidal variation with a semi-amplitude of $30\,\kmps$ (to emulate the motion of the spectrograph with respect to the Solar System barycenter).
-The spectral expectation model is built on the assumption of a normalized (unit) continuum and unresolved spectral lines with a Gaussian line-spread function (LSF) with one-sigma width $1/R$, where $R=135\,000$ is the spectrograph resolution).
-
-ANDY: continuum-rectified (unit-less) ?  HOGG: THE LINE FORMATION MODEL DESCRIBED ABOVE IS NOT RIGHT.
+The spectral expectation model is built on the assumption of a normalized (unit) continuum and unresolved spectral lines with a Gaussian line-spread function (LSF) with one-sigma width $1/R$, where $R=135\,000$ is the spectrograph resolution.
 
@@
 The bottom two panels show the same but for the well sampled input data.
-In each panel, the red line shows the true model used to generate the data.\label{fig:forward} ANDY: THIS IS BEAUTIFUL.
-HOGG SAY: I KNOW!}
+In each panel, the red line shows the true model used to generate the data.\label{fig:forward}}
     \end{mdframed}
@@
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
-The residuals look correlated (there is some ringing), and the residuals are larger near strong spectral features, especially in the case of the poorly sampled input data.
+The residuals look correlated (there is some ringing which mimics weak spectral features), and the residuals are larger near strong spectral features, especially in the case of the poorly sampled input data.
 
 \paragraph{Noise and noise covariances}
-HOGG do repeated trials to empirically measure the noise covariances. How exactly did we do that?
 
-In \figurename~\ref{fig:noise} we show the empirical covariances for the poorly sampled and well sampled data sets.
+We repeated the above experiments to empirically measure the noise covariances, and to estimate how correlated neighboring pixels are in the output spectrum.
+For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using the Forward Model (tm) and Standard Practice (tm).
+We then computed the empirical covariance in the output spectrum minus the true spectrum, which are shown in \figurename~\ref{fig:noise}.
 In each case, Standard Practice (tm) has a lower noise variance at zero lag, but shows spatially correlated noise over many pixels.
@@
 Give some summary of the above.
+% ARC: Did you want to add summary of the above, or is this comment stale and the summary is now below?
 
@@
 The source can vary, the instrument can vary, there can be variable foregrounds or backgrounds, and there can be mistakes or variations in calibration or data processing.
-All of these things happen in real data sets.
+All of these things happen in real data sets, but the Forward Model (tm) is no more impacted by this than what is done in Standard Practice (tm)!
 
@@
 For example, we could imagine forward-modeling generalizations of some of sparse methods (for example, \citealt{candes}) for applications in which the variability is localized in time or space; or we could imagine generalizations of matrix factorizations (for example, \citealt{hmf}) for applications in which variability is low dimensional (the \textsl{wobble} method has some such functionality; \citealt{wobble}).
+% HOGG: need candes and hmf citations
 

### 2023-09-21 python/ndi/__init__.py
docstring

@@ -23,12 +23,14 @@ def resample_spectrum(
     Sample a spectrum on a wavelength array given a set of pixels recorded from one or many visits.
     
     :param resample_wavelength:
-        A ``N``-length array of wavelengths to sample the spectrum on. In the paper, this is equivalent 
+        A ``M_star``-length array of wavelengths to sample the spectrum on. In the paper, this is equivalent 
         to the output-spectrum pixel grid $x_\star$.
 
     :param wavelength:
-        A ``M``-length array of wavelengths from individual visits. In the paper, this is equivalent to 
-        the input-spectrum pixel grid $x_i$.
+        A ``M``-length array of wavelengths from individual visits. If you have $N$ spectra where 
+        the $i$th spectrum has $m_i$ pixels, then $M = \sum_{i=1}^N m_i$, and this array represents a
+        flattened 1D array of all wavelength positions. In the paper, this is equivalent to the 
+        input-spectrum pixel grid $x_i$.
     
     :param flux:
         A ``M``-length array of flux values from individual visits. In the paper, this is equivalent to 

### 2023-09-21 setup.py
register as NoDataInterpolation; ndi was taken

@@ -1,7 +1,7 @@
 from setuptools import setup, find_packages
 
 setup(
-    name="ndi",
+    name="NoDataInterpolation",
     version="0.0.1",
     package_dir={"": "python"},
     packages=find_packages(where="python"),

### 2023-09-21 python/ndi/__init__.py
removed superfluous tests

@@ -253,164 +253,3 @@ def _check_shapes(wavelength, flux, ivar, mask):
         
 
 
-if __name__ == "__main__":
-
-    # This is instantaneous now..
-    x_min = 8.7000 # minimum ln wavelength
-    x_max = 8.7025 # maximum ln wavelength
-
-    # So let's test with 6750 pixels, which is something we could NOT afford when we had to construct all the 2D arrays slowly.
-    #x_min = 8.7000 # minimum ln wavelength
-    #x_max = 8.7500 # maximum ln wavelength
-
-    import numpy as np
-    import pylab as plt
-    import scipy.interpolate as interp
-    from time import time
-
-    np.random.seed(34)
-    c = 299792458. # m / s
-    sqrt2pi = np.sqrt(2. * np.pi)
-
-
-    R = 1.35e5 # resolution
-    dxstar = 1. / R # output pixel grid spacing
-    xstar = np.arange(x_min + 0.5 * dxstar, x_max, dxstar) # output pixel grid
-    P = np.round((x_max - x_min) * R).astype(int) # number of Fourier modes (ish)
-
-
-    # define high-level parameters, especially including spectrograph parameters
-    sigma_x = 1. / R # LSF sigma in x units
-    lines_per_x = 2.0e4 # mean density (Poisson rate) of lines per unit ln wavelength
-    ew_max_x = 3.0e-5 # maximum equivalent width in x units
-    ew_power = 5.0 # power parameter in EW maker
-    badfrac = 0.01 # fraction of data to mark bad    
-
-    # Set the pixel grid and model complexity for the output combined spectrum
-    Mstar = len(xstar) # number of output pixels
-
-    # set up the line list for the true spectral model
-    x_margin = 1.e6/c # hoping no velocities are bigger than 1000 km/s
-    x_range = x_max - x_min + 2. * x_margin # make lines in a bigger x range than the data range
-    nlines = np.random.poisson(x_range * lines_per_x) # set the total number of lines
-    line_xs = (x_min - x_margin) + x_range * np.random.uniform(size=nlines)
-
-    # give those lines equivalent widths from a power-law distribution
-    line_ews = ew_max_x * np.random.uniform(size=nlines) ** ew_power # don't ask
-
-
[…]

### 2023-09-21 paper/main.tex

@@
 But importantly, we aren't proposing a change to the \emph{format} of the input or outputs of spectral or image combination.
+
+% Hogg's email epiphany:
+% 1.
+I think I figured out what Blanton means by saying that interpolation of the data is a kind of fitting.
+It is!
+But those fits can’t be combined well because the fits are infinitely over-parameterized.
+If the interpolation was done approximately with a finite-parameter fit, we could interpolate and then average just fine.
+I’m writing this here to remind myself of this epiphany for when I come back online and write a subsection about this in the paper.
+
+% Hogg: what do you want to do about this?
 
@@
 
-We assumed no sky subtraction necessary. No tellurics. What gives if you have multiple shifting signals. Causal structure. etc.
+% What about non-Gaussian noise?
 
-What about non-Gaussian noise?
+We assumed that the noise was Gaussian, but this is not always true of astronomical data.
+In the case of Poisson noise that is not background dominated, the variance tensor $C_i$ will be biased, and all our assumptions about additive Gaussian variance will be broken.
+% ARC: what else should we say about this?
 
-Code is available at HOGG THIS URL.
+Code is available at \url{https://github.com/davidwhogg/NoDataInterpolation} and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
 

### 2023-09-21 paper/main.tex
Updates from Overleaf

--- 2023-09-21 paper/main.tex
@@
-% Copyright 2022 David W. Hogg (NYU) and Andy Casey (Monash)
+% Copyright 2022, 2023 David W. Hogg (NYU) and Andy Casey (Monash)
 
@@
 % - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.
+% - Blanton sent an email with comments.
 
@@
 
-\title{\Huge Don't interpolate the data!}
-% ARC: I changed this to "Don't interpolate the data!"? because you can't like, own data, man.
-% (but you can change it back if you prefer)
+\title{\Huge Don't interpolate your data!}
 
@@
 \affiliation{School of Physics \& Astronomy, Monash University}
-\affiliation{Centre of Excellence for Astrophysics in Three Dimensions (ASTRO-3D)}
 
@@
 Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean.
-The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
+The key component of this conversion is an image (or spectrum) model that is completely flexible but also a continuous function of position (or wavelength) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
 The benefits of a modeling approach are myriad:
@@
 Noise maps, data gaps, and bad-data masks don't require interpolation.
-The output can take the form of a spectrum evaluated on a pixel grid, as is traditional.
-The noise in the output spectrum becomes uncorrelated across neighboring pixels. 
+The output can take the form of an image or spectrum evaluated on a standardized pixel grid, as is traditional.
+The uncertainties in the output become uncorrelated across neighboring pixels. 
 The only cost is a small increase in computational complexity.
@@
 Is it astronomical blasphemy to say that this practice is \emph{wrong?}
-It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}). % HOGG
+It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}).
 The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
 Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
-Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
+Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator},
+but it is.
 It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
@@
 The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
-We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps metadata flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
+We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates.
 That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
@@
 
-The approach we propose differs from interpolation in a subtle way. 
-Interpolation could be described as computing new values at arbitrary positions based on an exising set of known data points. 
-Instead, here we propose to construct a forward model for the data, where we then evaluate that model on the given set of output positions (pixels).
-The \emph{output product} is the same form as what one would expect from an interpolation procedure (e.g., computed values along some array of positions), but the \emph{procedure} differs from interpolation.
-When interpolating data the basis functions are explicitly constructed from the data, whereas here we construct a forward model that is maximally consistent with the data, and evaluated at new positions.
-
 The ideas here aren't novel; many projects have avoided data interpolation previously.
@@
 There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{froe}, \citealt{sp}):
-Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
+Optimal extraction is an unfortunate name for the process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
 One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is a better method that moves this spectral combination method directly into the two-dimensional data.
@@
 
-HOGG: Most important: Say here why you should never want to do this at all! % ARC: I think you've covered this by talking about the two-dimensional data.
-Comment on a spectrum being too low in s/n to measure a Na line...!! % ARC: I don't remember what the context is for this
+HOGG: Most important: Say here why you should never want to do this at all!
+Comment on a spectrum being too low in s/n to measure a Na line...!!
 
@@
 Don't actually construct the very sparse $C_i$ matrices, just do pixel-wise multiplies.
-Don't use \texttt{inv()}, use \texttt{solve()} or \texttt{lstsq()}.
+Don't use \texttt{inv()}; use \texttt{solve()} or \texttt{lstsq()}.
 Check your condition numbers.
@@
 If you use the non-uniform fast Fourier transform, order your frequencies in the way the software wants, and deal with the complex numbers.
-Specific choice of $L$ and $P$ and $M_\star$ in the Fourier case.
+HOGG: Specific choice of $L$ and $P$ and $M_\star$ in the Fourier case.
 
@@
   Jon Holtzman (NMSU), 
+  Jennifer Johnson (OSU),
   Dustin Lang (Perimeter),
@@
 \bibitem[Williams et al.(1996)]{hdf} Williams, R.~E., Blacker, B., Dickinson, M., et al.\ 1996, \aj, 112, 1335. doi:10.1086/118105
-\bibitem[Wilson et al.(2019)]{apogeehardware} Wilson, J.C., Hearty, F.R., Skrutskie, M.F., Majewski, S.R., Holtzman, J.A., Eisenstein, D., et al.\ 2019, \pasp, 131, 055001.
-doi:10.1088/1538-3873/ab0075.
 \bibitem[Wolf et al.(2003)]{combo17} Wolf, C., Meisenheimer, K., Rix, H.-W., et al.\ 2003, \aap, 401, 73. doi:10.1051/0004-6361:20021513

### 2023-10-05 paper/main.tex
add comment about weak Na, citations to Zackay, Kelson, etc

--- 2023-09-21 paper/main.tex
@@
 % text macros
-\shorttitle{don't interpolate your data}
+\shorttitle{don't interpolate the data}
 \shortauthors{hogg \& casey}
@@
 For one example, the \textsl{COMBO-17} survey (\citealt{combo17}) took many images of the sky in many bands, and did not co-add the data prior to making measurements, in part to avoid mixing different point-spread functions and pixel samplings.
-For another, what we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models.
+For another, some projects have sought to co-add background-dominated images in a near optimal way, but presume that all observations are already registered and re-sampled to the same grid \citep{zackay}.
+What we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models.
 Plus it learns (it isn't told) the epoch-to-epoch spectral shifts.
-There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{froe}, \citealt{sp}):
+There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{kelson}, \citealt{froe}, \citealt{sp}):
 Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
@@
 HOGG: Most important: Say here why you should never want to do this at all! % ARC: I think you've covered this by talking about the two-dimensional data.
-Comment on a spectrum being too low in s/n to measure a Na line...!! % ARC: I don't remember what the context is for this
 
@@
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
-The residuals look correlated (there is some ringing which mimics weak spectral features), and the residuals are larger near strong spectral features, especially in the case of the poorly sampled input data.
+The residuals look correlated (there is ringing which mimics weak spectral features), and the residuals are larger near strong spectral features, especially in the case of the poorly sampled input data.
 
@@
 
+HOGG: What is the third attitude?
+
+The combined spectrum computed with the Forward Model(tm) has minimal pixel-to-pixel covariances based on repeated experiments.
+In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
+This is particularly evident in the second panel of Figure~\ref{fig:noise} near $\ln\lambda = 8.7024$.
+Here the pixel-to-pixel covariances resulting from Standard Practice(tm) are no longer academic, as the covariances here mimic weak spectral features.
+For example, if a weak Na line existed near this wavelength then a spectroscopist might readily interpret these covariances as real spectral features and report an enhanced Na abundance.
+Checking the strength of other (stronger) Na features would be a sensible check for this interpretation, but often there are few spectral lines available for a given element.
+
 One side note about epoch-to-epoch variations in instrument or data-processing parameters:
@@
 Matt Daunt (NYU),
+Dylan Green (UC Irvine),
 Jon Holtzman (NMSU), 
@@
 \bibitem[Ivezi{\'c} et al.(2019)]{lsst} Ivezi{\'c}, {\v{Z}}., Kahn, S.~M., Tyson, J.~A., et al.\ 2019, \apj, 873, 111. doi:10.3847/1538-4357/ab042c
+\bibitem[Kelson(2003)]{kelson} Kelson, D.~D.\ 2003, \pasp, 115, 688. doi:10.1086/375502
 \bibitem[Majewski et al.(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2017, \aj, 154, 94. doi:10.3847/1538-3881/aa784d
@@
 \bibitem[Zechmeister et al.(2014)]{froe} Zechmeister, M., Anglada-Escud{\'e}, G., \& Reiners, A.\ 2014, \aap, 561, A59. doi:10.1051/0004-6361/201322746
+\bibitem[Zackay, B. \& Ofek (2017)]{zackay} Zackay, B. \& Ofek, E. O.\ 2017, \apj, 836, 188. doi:10.3847/1538-4357/836/2/188
+
 \end{thebibliography}

### 2023-11-02 paper/main.tex
Updates from Overleaf
start de-escalation of the language

@@
-% Copyright 2022 David W. Hogg (NYU) and Andy Casey (Monash)
+% Copyright 2022, 2023, [add your years here] David W. Hogg (NYU) and Andy Casey (Monash)
 
@@
 % -----------
+% - Change the tone to "here's a method; it is principled and has great properties" and remove all suggestion that it is THE BEST THING.
+% - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
+% - Make comments directly about Drizzle to justify the name.
+% - discuss conservation of flux vs intensity.
+% - Blanton asks: Is this identical to sinc shift and average? Hogg thinks it MIGHT be.
+% - Add Hogg response to Blanton: Interpolate with (say) cubic spline and then average is NOT THE SAME as fit all the data simultaneously with a cubic spline function on a particular set of knots.
+% - Dustin asks: Is this effectively producing a sinc LSF? Hogg thinks it DEFINITELY IS NOT. Explain why not in the text.
+% - Hogg notes: Is it clear in the text that we are taking everything to a standard instrument?
+And the user can decide what that instrument is?
+That isn't in our current implementation, but it COULD BE.
+Discussion point?
+% - Related Hogg issue: Any latent spectrum (like our combined spectrum) should be something that COULD BE OBSERVED by some instrument in the future. Hence spectro-perfectionism is wrong (ish).
+% - Read and cite the DESI spectral pipeline paper; Dustin says it's relevant.
 % - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
@@
 
-\title{\Huge Don't interpolate the data!}
-% ARC: I changed this to "Don't interpolate the data!"? because you can't like, own data, man.
-% (but you can change it back if you prefer)
+\title{\large Fowizzle: Combining spectra or images by forward modeling}
 
@@
 \begin{abstract}\noindent
-When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is often tempting to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
-Bound-saturating measurements are made not by manipulating data, but instead by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
-Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that no data are ever harmed, and yet the output is the shift-adjusted mean.
+When there are many observations of an astronomical source---many images with different dithers, or many spectra taken at different barycentric velocities---it is standard practice to shift and stack the data, to (for example) make a high signal-to-noise average image or mean spectrum.
+Bound-saturating measurements are made by manipulating a likelihood function, where the data are treated as fixed, and model parameters are modified to fit the data.
+Traditional shifting and stacking of data can be converted into a model-fitting procedure, such that the data are not modified, and yet the output is the shift-adjusted mean.
 The key component of this conversion is a spectral model that is completely flexible but also a continuous function of wavelength (or position in the case of imaging) that can represent any signal being measured by the device after any reasonable translation (or rotation or field distortion).
@@
 Noise maps, data gaps, and bad-data masks don't require interpolation.
-The output can take the form of a spectrum evaluated on a pixel grid, as is traditional.
-The noise in the output spectrum becomes uncorrelated across neighboring pixels. 
+The output can take the form of an image or spectrum evaluated on a pixel grid, as is traditional.
+In addition to shifts, the model can account for line-spread or point-spread function variations, world-coordinate-system variations, and calibration or normalization variations.
+The noise in the output becomes uncorrelated across neighboring pixels as the shifts deliver good coverage in some sense.
 The only cost is a small increase in computational complexity.
-We demonstrate all of these things with a small data example and we provide open-source sample code for re-use.
+We demonstrate the method with a small data example and we provide open-source sample code for re-use.
 \end{abstract}
@@
 and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.
+This technology is encoded in tools such as \textsl{Drizzle} (\citealt{drizzle}) and HOGG.
 
-Is it astronomical blasphemy to say that this practice is \emph{wrong?}
-It is wrong because minimum-variance unbiased estimators require care in their construction; they are (often) maximum-likelihood estimators (\citealt{something}). % HOGG
-The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of the image or spectrum, given the noisy data of the individual exposures.
-Noisy data are combined to make a measurement by building and optimizing a likelihood, not by shifting and stacking the input data.
+These traditional practices are reliable and fast.
+However, they will---in some cases---sacrifice some information.
+Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{something}). % HOGG
+The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of some kind of ``true'' or high signal-to-noise image or spectrum, given the noisy data of the individual exposures.
+That is, a combined image is a kind of \emph{prediction}: It is the prediction of what \emph{would have been observed} with a much longer exposure with a much better device.
+In measurement theory, the practice is to combine noisy data into a measurement by building and optimizing a likelihood, not by shifting and stacking (or performing other arithmetic operations on) the input data.
 Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
@@
 Indeed it has just as much freedom as the shifting-and-stacking standard practice.
-The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation.
+The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation, by means of a justified likelihood function.
 
-The data are never manipulated when using an estimator (in contrast to standard practice), but (for us) the functional requirements are the same as with standard practice:
-We consider here only procedures that start with the individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
+A boundary condition of the work presented here is that the functional requirements of anything we provide must be the same as with standard practice:
+That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
 That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.

### 2024-03-05 paper/main.tex
Updates from Overleaf
combination not necessary

@@
 % - Change the tone to "here's a method; it is principled and has great properties" and remove all suggestion that it is THE BEST THING.
+We don't need to win on BEST.
+We can win on sensible, well justified, good, implemented, demonstrated.
 % - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
@@
 The noise in the output becomes uncorrelated across neighboring pixels as the shifts deliver good coverage in some sense.
-The only cost is a small increase in computational complexity.
+The only cost is a small increase in computational complexity over that of traditional methods.
 We demonstrate the method with a small data example and we provide open-source sample code for re-use.
@@
 
-% Hogg: what do you want to do about this?
+% Hogg: what do you want to do about this? HOGG SAY: I want to explicitly address it, maybe in a first section after the Introduction. This introduction is too long anyway!
 
 The approach we propose differs from interpolation in a subtle way. 
-Interpolation could be described as computing new values at arbitrary positions based on an exising set of known data points. 
+Interpolation could be described as computing new values at arbitrary positions based on an existing set of known data points. 
 Instead, here we propose to construct a continuous spectral model for the data, where we then evaluate that model on the given set of output positions (pixels).
@@
 
-HOGG: Most important: Say here why you should never want to do this at all! % ARC: I think you've covered this by talking about the two-dimensional data.
+Finally it is worthy of remark that, in some conceptual sense, there is no reason to \emph{ever} combine the individual spectroscopic exposures at all, for any reason, except possibly visualization and good vibes.
+No information is gained by performing a combination.
+Any line or feature that becomes visible under spectral combination must---at least in principle---be measurable in the uncombined individual spectra if those individual spectra are analyzed appropriately simultaneously.
+That was the insight and motivation for the unusual \textsl{COMBO-17} (\citealt{combo17}) approach in which the individual exposures were never co-added or combined; it is also the underlying idea behind work we have done on the proper motions and parallaxes of very faint sources (\citealt{undetectable}).
+Averaging the data doesn't ever create any new information.
+It can be satisfying though; it meets astronomical community expectations; and it might provide insights or inspire discoveries.
 
@@
 
+%% HOGG SAY: This is not quite correct: We are not assuming no time var. We are assuming that the user wants the mean.
 \paragraph{Time variability}
@@
 \bibitem[Kelson(2003)]{kelson} Kelson, D.~D.\ 2003, \pasp, 115, 688. doi:10.1086/375502
+\bibitem[Lang et al.(2009)]{undetectable} Lang, D., Hogg, D.~W., Jester, S., et al.\ 2009, \aj, 137, 4400. doi:10.1088/0004-6256/137/5/4400
 \bibitem[Majewski et al.(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2017, \aj, 154, 94. doi:10.3847/1538-3881/aa784d

### 2024-03-07 paper/main.tex
fixed some inconsistencies about noise and sizes

@@
 \paragraph{Multi-epoch spectra with shifts}
-We will assume that there are $n$ observations (epochs) of the same star (or maybe of different stars or quasars, say, that are assumed to be more-or-less identical, intrinsically, but let's think of a single star for now).
+We will assume that there are $N$ observations (epochs) of the same star (or maybe of different stars or quasars, say, that are assumed to be more-or-less identical, intrinsically, but let's think of a single star for now).
 Each of these observations $i$ (with $1\leq i\leq n$) produces a one-dimensional observed (noisy) image $y_i$.
-The $y_i$ can be thought of as $m_i$-vectors in what follows, where $m_i$ is the number of pixels in the $i$-th spectrum.
+The $y_i$ can be thought of as $M_i$-vectors in what follows, where $M_i$ is the number of pixels in the $i$-th spectrum.
 The different observations are made at different relative velocities (star minus spectrograph) such that the images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
@@
 \paragraph{Wavelength calibration}
-We will assume that, for every $m_i$-pixel image $y_i$ we also have a $m_i$-vector (or list) $x_i$ of pixel positions, such that each element of $x_i$ gives an accurate and precise value for the wavelength (or position in the spectrograph, in some settings) for each pixel of the image $y_i$.
+We will assume that, for every $M_i$-pixel image $y_i$ we also have a $M_i$-vector (or list) $x_i$ of pixel positions, such that each element of $x_i$ gives an accurate and precise value for the wavelength (or position in the spectrograph, in some settings) for each pixel of the image $y_i$.
 That is, the device is well calibrated, and we know all the housekeeping data we need for image $y_i$.
+
+If the data being combined were images rather than spectra, the relevant calibration would be astrometric calibration; to wit, that each pixel has a known position on the sky.
 
@@
 Each observation is noisy.
-We will assume that the noise has known properties, and especially that it is additive, (nearly) zero-mean and (nearly) Gaussian in form, with known variance.
-Under these assumptions, each spectrum $y_i$ has an associated variance tensor $C_i$.
-If $y_i$ is an $m_i$-vector, then $C_i$ is a nonnegative-definite $m_i\times m_i$ 2-tensor.
-We don't need to assume that the different pixels of the spectrum received independent noise, but we do assume that each epoch $i$ has an independent noise draw.
+We will assume that the noise has good properties, and especially that it is additive, (nearly) zero-mean and (nearly) Gaussian in form.
+We will \emph{not} assume that you know the variance of the noise accurately, but you can do better with this method if you do know the noise variances accurately for each pixel.
+In what follows, each spectrum $y_i$ is given an associated (but possibly inaccurate) noise-variance tensor $C_i$.
+If $y_i$ is an $M_i$-vector, then $C_i$ must be a nonnegative-definite $M_i\times m_i$ 2-tensor.
+In this description, there is no need to assume that the different pixels of the spectrum received independent noise, but we do assume that each epoch $i$ has an independent noise draw.
 
@@
 This mask has value 1 in the locations in all good pixels, and 0 in the locations of all bad pixels.
-
-\paragraph{Bit-flags}
-A bit-flag is often used to store meta data about a pixel that otherwise cannot be encoded by the spectrum $y_i$ or its asssociated variance tensor $C_i$ (e.g., poorly described line spread function).
-We often want to propagate this meta data to appropriate pixels in the combined spectrum. 
-These bit-flag masks are usually encoded as integers with up to $F$ potential flags, allowing for all (or a subset of) flags to be encoded as a single value $\sum^{F}_{f=0} 2^f$. 
-This bit-flag mask can be separated into $F$ masks where each has value 1 in the locations where the flag is marked, and 0 in the locations where it is not.
+There are many different ways in which a pixel can be bad; in what follows the assumption is that the relevant judgements have been made more-or-less correctly.
 
 \paragraph{Data gaps}
-There is no assumption here that all the $y_i$ have the same number of pixels $m_i$, nor that the pixel grid is in any sense uniform or complete.
+There is no assumption here that all the $y_i$ have the same number of pixels $M_i$, nor that the pixel grid is in any sense uniform or complete.
 That is, there can be gaps and spaces in the wavelength coverage of the spectra, as there are at chip gaps and where the detector has bad columns.
@@
 \paragraph{Pixel-convolved line-spread function}
-The spectrograph has something like a point-spread function or line-spread function, which, in the case of spectrosccopy, sets the shape of an unresolved spectral line.
+The spectrograph has something like a point-spread function or line-spread function, which, in the case of spectroscopy, sets the shape of an unresolved spectral line.
 In general it depends on the slit or fiber width, the spectrograph resolution and optics, instrument focus, and the properties of the detector.
@@
 This leads to spectral (or image) variations, and violate the assumptions of this model (and also the implicit assumptions underlying standard practice).
-We won't address this kind of variability here, except in some discussion at the end.
+We won't address this kind of variability here, except in some discussion at the end:
+The method presented here can be straightforwardly modified to account for LSF or PSF variations.
 
 \paragraph{Continuous spectral model} 
-In what follows we are going to avoid interpolating the \emph{data} by having a spectral \emph{model} that can be arbitrarily and losslessly interpolated.
+In what follows we are going to avoid interpolating the data.
+We can avoid it by having a spectral \emph{model} instead that can be arbitrarily and losslessly interpolated.
 That means that the spectral model must be a representation of a continuous one-dimensional function.
@@
 The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination.
-Thus we will use a linear basis.
+Thus we will make use of a linear basis.
 
 \paragraph{Band limit}
-We say that a signal is \emph{band-limited} if it has no frequency content above some highest possible cutoff frequency.
+We are not assuming, in what follows, that the result we are looking for will be band-limited;
+we say that a signal is \emph{band-limited} if it has no frequency content above some highest possible cutoff frequency.
 There is a weird sense in which data from a spectrograph both is, and is not, band limited.
@@
 It might surprise the reader to learn that we prefer to let the model capture higher frequencies, even when we think they shouldn't be there for any reasons other than noise, given the hardware.
-More on this below.
+More on this below. %% HOGG DO WE COME BACK TO THIS?
 
+% HOGG SAY: IT is not at all clear here that we are NOT assuming critical sampling. We aren't! Make that clear.
+% In general this section has issues.
 \paragraph{Critical sampling}
-Because of considerations related to the Nyquist frequency on a uniform grid, it is valuable for a spectrograph to have a pixel spacing on the device that is smaller than the width of the PCLSF.
+We are not assuming, in what follows, that the data are critically sampled.
+Because of considerations related to the Nyquist frequency on a uniform grid, it is valuable for spectra to be represented with a pixel spacing that is smaller than the width of the PCLSF.
 Specifically, it is valuable for there to be two or three pixels across the full-width at half-maximum (FWHM) of the PCLSF.
 This keeps the band-limited part of the spectroscopic signal well-sampled in the data.
-There are spectrograph that don't have pixel scales this fine (for example, \textsl{APOGEE}; \citealt{apogee}); in these cases what we are proposing here is \emph{even more important} than it is in the well-sampled case.
-This will become obvious in the experiments below.
+In general, interpolation works much better on critically sampled data than it does on ill-sampled data.
+
+The fact that we are not interpolating the data means that we don't require critical sampling in the raw data.
+And indeed, there are spectrographs that don't have critically sampled pixel scales (for example, \textsl{APOGEE}; \citealt{apogee}).
+Thus we will not assume that the raw data are critically sampled, but we will produce the output mean spectrum on a grid that is critically sampled.
+The input and output pixel grids do not need to coincide when the model is a continuous function.
 
@@
 In the special case that the spectral model is a Fourier series, there are very useful and fast tools for interpolation, including especially the non-uniform fast Fourier transform (for example, \citealt{finufft}).
+These tools make the spectral combination substantially faster than naive linear algebra estimates would suggest.
 
-%% HOGG SAY: This is not quite correct: We are not assuming no time var. We are assuming that the user wants the mean.
 \paragraph{Time variability}
-When we average data (or build a static model of multi-epoch data), we are implicitly assuming that there is no intrinsic variability to the source being observed.
-That is, we are assuming that the changes from observation to observation are solely from either the shifts $\Delta x_i$ or from the noise.
-We interpolate the model to account for the shifts, and we combine data to beat down the noise.
-There are situations in which there is substantial time variability and these approaches are still valid, and there are modified approaches.
-We will discuss these changes and generalizations at the end.
+When we average data (or build a static model of multi-epoch data), we are implicitly assuming that we do not care about any of the intrinsic variability of the source being observed.
+That is, we are assuming that the changes from observation to observation are solely from either the shifts $\Delta x_i$ or else from effects that we consider to be noise.
+We interpolate the model to account for the shifts, and we combine data to average away this noise.
+In many cases the stellar variability is itself the subject of study, and then the variability is not entirely noise.
+But from the perspective of this project, any stellar variability is considered to be part of the total additive noise in the spectrum.
 
@@
 Again, the assumption is that the different calibrated or processed observations are of the same underlying signal.
-If there are multiple signals participating, the model would have to expand into something like \textsl{wobble} (\citealt{wobble}).
+
+Many of these effects can be accounted for in straightforward generalizations of what we deliver below.
+For example, if there are multiple signals participating, the model would have to expand into something like \textsl{wobble} (\citealt{wobble}).
+For another, if there are normalization variations, a normalization model can be added to the image combination procedure.
+We return to these issues at the end. %% HOGG DO WE?
 
@@
 
+HOGG Maybe consider the case of $P\sim\sum_i M_i$. Double descent and all? Or say this earlier in the method section?
+
 One side note about epoch-to-epoch variations in instrument or data-processing parameters:

### 2024-03-08 paper/main.tex
Implementation notes and references

@@
 % -----------
-% - Change the tone to "here's a method; it is principled and has great properties" and remove all suggestion that it is THE BEST THING.
-We don't need to win on BEST.
-We can win on sensible, well justified, good, implemented, demonstrated.
+% - What's our name? Choose and propagate into where "Forward Model(tm)" is used.
 % - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
@@
 However, they will---in some cases---sacrifice some information.
-Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{something}). % HOGG
+Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{mvue}). % HOGG
 The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of some kind of ``true'' or high signal-to-noise image or spectrum, given the noisy data of the individual exposures.
@@
 Each of these observations $i$ (with $1\leq i\leq n$) produces a one-dimensional observed (noisy) image $y_i$.
-The $y_i$ can be thought of as $M_i$-vectors in what follows, where $M_i$ is the number of pixels in the $i$-th spectrum.
+The $y_i$ can be thought of as $M_i$-vectors in what follows, where $M_i$ is the number of pixels in spectrum $i$.
 The different observations are made at different relative velocities (star minus spectrograph) such that the images are shifted by a different shift $\Delta x_i$ relative to the detector or extracted wavelength grid.
@@
 Importantly, the bad data get zeros in the \emph{inverse} covariance matrix, not the covariance matrix!
-We will propagate the meta data from the bit-flag mask later.
 
@@
     g_j(x) & = \left\{\begin{array}{cl}\displaystyle\cos\left(\frac{\pi\,[j-1]}{L}\,x\right) & \mbox{for $j$ odd} \\[3ex]
-                                       \displaystyle\sin\left(\frac{\pi\,j}{L}\,x\right) & \mbox{for $j$ even}\end{array}\right. ~,
+    \displaystyle\sin\left(\frac{\pi\,j}{L}\,x\right) & \mbox{for $j$ even}\end{array}\right. ~,\label{eq:basis}
 \end{align}
@@
 \begin{align}
-    \hat\theta &= (X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1}\,Y ~,
+    \hat\theta &= (X^\top\,C^{-1}\,X)^{-1}\,X^\top\,C^{-1}\,Y ~, \label{eq:xtxinvxty}
 \end{align}
@@
 \paragraph{Implementation notes}
-Don't actually construct the very sparse $C_i$ matrices, just do pixel-wise multiplies.
-Don't use \texttt{inv()}, use \texttt{solve()} or \texttt{lstsq()}.
-Check your condition numbers.
-If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra; research those.
-If you use the non-uniform fast Fourier transform, order your frequencies in the way the software wants, and deal with the complex numbers.
-Specify your choice of $L$ and $P$ and $M_\star$ in the Fourier case.
+Although the equations like \eqref{eq:xtxinvxty} show matrix multiplies between $C^{-1}$ and $X$, it is not a good idea to perform these multplies as true matrix multiplies if the $C^{-1}$ matrix is diagonal (as it often is).
+When $C^{-1}$ is diagonal, the matrix multiply can be done as a broadcast of the vector of diagonal values instead, which is faster (and way less memory intensive).
+
+Although the equations like \eqref{eq:xtxinvxty} show inverses, it is almost never a good idea to use the native linear-algebra inverse function \texttt{inv()} in any implementation of this method.
+The reason is that we don't want the inverse matrix itself, we want the outcome of the inverse matrix multiplied by $X^\top\,C^{-}\,Y$.
+For this use case, it is more precise and more more numerically stable to use the linear algebra \texttt{solve()} or \texttt{lstsq()} functions.
+
+Whenever performing\texttt{inv()}, \texttt{solve()}, or \texttt{lstsq()} operations, it is sensible to begin by checking the condition number (the ratio of the largest to the smallest singular value) of the matrix in play.
+Condition numbers larger than $10^8$ are usually trouble, even when the mathematics is being performed at double precision.
+When condition numbers get large, maybe a different basis or a re-scale of some basis functions might be a good idea.
+
+If $X$ is either very sparse (as it is in a b-spline basis) or Fourier (as it is in a sine-and-cosine basis) then there are very fast ways to do the linear algebra.
+We could have used the \texttt{finufft} package (\citealt{finufft}) to do the mathematics fast, because we have a Fourier basis here.
+Sparse linear algebra packages like that in \texttt{scipy} (\citealt{scipy}) are available; they would similarly speed up the code when using spline-like or wavelet-like bases.
+
+If you use any non-uniform fast Fourier transform package, the programming interface of that package will constrain how the frequencies must be ordered in the code, and it will probably require complex-number implementation.
+These have impacts on how the code is written.
+In particular the fact that the output spectrum $y_\ast$ is real-valued constrains the complex Fourier amplitudes.
+Continuing on the Fourier theme:
+Fourier series, technically, model \emph{periodic} functions.
+The length scale $L$ of that periodicity in \eqref{eq:basis} must be set to a length longer than the $x$-space span of any of the input or output pixel grids.
+
+We recommended setting $P=M_\ast$.
+We recommend very strongly \emph{against} setting $P\approx M$, where $M=\sum_i M_i$.
+That is, you want the number of parameters to equal your number of output pixels, but you \emph{don't} want your number of parameters to be close to the number of \emph{input} pixels.
+The reason for this is the ``double descent'' phenomenon, which de-stabilizes linear fitting (\citealt{doubledescent}).
+If you find yourself in the situation that $M_\ast\approx M$, reconsider your life choices.
 
@@
 
-\paragraph{Forward Model (tm)}
-As our main experiment or result, we apply the Forward Model (tm) as described above in \sectionname~\ref{sec:method} to the input data shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
+\paragraph{Forward Model(tm)}
+As our main experiment or result, we apply the Forward Model(tm) as described above in \sectionname~\ref{sec:method} to the input data shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
 In each case, the output pixel grid $x_\star$ is chosen to be well sampled, with a pixel spacing of $1/R$ in the (natural) logarithm of wavelength.
@@
     \end{center}
-    \caption{The top two panels show the result of running the Forward Model (tm) on the poorly sampled input data, plus the residual away from the truth.
+    \caption{The top two panels show the result of running the Forward Model(tm) on the poorly sampled input data, plus the residual away from the truth.
 The bottom two panels show the same but for the well sampled input data.
@@
 \end{figure}
-The combined spectrum resulting from the Forward Model (tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
+The combined spectrum resulting from the Forward Model(tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
 It is compared to the true spectral expectation model that was used to generate the data.
-Also shown in \figurename~\ref{fig:forward} is the residuals, Forward Model (tm) minus truth.
+Also shown in \figurename~\ref{fig:forward} is the residuals, Forward Model(tm) minus truth.
 The residuals look stationary, approximately Gaussian, and uncorrelated (more on this below).
 
-\paragraph{Standard Practice (tm)}
+\paragraph{Standard Practice(tm)}
 In standard practice, the individual spectra $y_i$ are interpolated to rest-frame spectra $y'_i$ using an interpolator.
@@
 
-This censorship of the bad pixels requires some attention:
+In Standard Practice(tm), this censorship of the bad pixels requires some attention:
 Naive interpolation of an integer (binary actually) bad-pixel mask $b_i$ will not deliver an integer mask $b'_i$.
@@
     \end{center}
-    \caption{As part of Standard Practice (tm), it is necessary to interpolate the individual epoch spectra before averaging them.
+    \caption{As part of Standard Practice(tm), it is necessary to interpolate the individual epoch spectra before averaging them.
 This shows one epoch of the poorly sampled data, interpolated to the output $x_\star$ pixel grid in the rest frame.
@@
     \end{center}
-    \caption{Same as \figurename~\ref{fig:forward} but for the output of Standard Practice (tm).
+    \caption{Same as \figurename~\ref{fig:forward} but for the output of Standard Practice(tm).
 Note that in the poorly sampled case, the residuals appear to be spatially correlated, and the residuals appear to be larger near strong spectral features; more about these issues in the text and in \figurename~\ref{fig:noise}.\label{fig:standard}}
@@
 \end{figure}
-We show the final result of Standard Practice (tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
+We show the final result of Standard Practice(tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
@@
 We repeated the above experiments to empirically measure the noise covariances, and to estimate how correlated neighboring pixels are in the output spectrum.
-For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using the Forward Model (tm) and Standard Practice (tm).
+For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using the Forward Model(tm) and Standard Practice(tm).
 We then computed the empirical covariance in the output spectrum minus the true spectrum, which are shown in \figurename~\ref{fig:noise}.
-In each case, Standard Practice (tm) has a lower noise variance at zero lag, but shows spatially correlated noise over many pixels.
-The Forward Model (tm), in contrast, shows no pixel-to-pixel noise covariances, even when the input data are ill sampled.
+In each case, Standard Practice(tm) has a lower noise variance at zero lag, but shows spatially correlated noise over many pixels.
+The Forward Model(tm), in contrast, shows no pixel-to-pixel noise covariances, even when the input data are ill sampled.
 \begin{figure}[t!]
@@
 The variances were estimated by performing the experiments repeated 64 times, with the same spectral expectations and sampling but unique noise draws.
-The Forward Model (tm) has slightly larger variance per pixel at zero lag, but vanishing pixel-to-pixel correlations.
-In contrast, the Standard Practice (tm) leads to correlated pixel values, with correlations extending out to many pixels.\label{fig:noise}}
+The Forward Model(tm) has slightly larger variance per pixel at zero lag, but vanishing pixel-to-pixel correlations.
+In contrast, the Standard Practice(tm) leads to correlated pixel values, with correlations extending out to many pixels.\label{fig:noise}}
     \end{mdframed}
@@
 
-HOGG: Note that if we choose $P<M$ we start to see noise correlations in the Forward Model (tm) too.
-That's one of a few reasons we like $P=M$.
+HOGG: Add intuition here about WHY the output noise model is close to uncorrelated?
+HOGG: Note that if we choose $P<M_\ast$ we start to see noise correlations in the Forward Model(tm) too.
+That's one of a few reasons we like $P=M_\ast$.
 
@@
 
-Give some summary of the above.
-% ARC: Did you want to add summary of the above, or is this comment stale and the summary is now below?
+We have presented a forward-modeling approach to combining spectra or images.
+This method has good properties:
+It doesn't involve interpolating or otherwise distorting the input data;
+it produces a combined or mean spectrum in the format that traditional methods employ;
+it makes no assumptions about the form of the combined spectrum (except that it can be represented on an $M_\ast$-pixel grid);
+and it has excellent noise properties in real circumstances.
+The method can work on a wide variety of different kinds of data, but it is especially valuable in projects where the individual input exposures are ill-sampled, or not as well sampled as one might like.
+It is also good for situations in which the input data have heterogeneous bad pixels or data gaps.
 
@@
 But it is not different, conceptually, from what is presented above.
+That is, once the (now more complicated) shift operators $\Delta x_i$ are known, the problem proceeds by the same mathematics.
 
@@
 The source can vary, the instrument can vary, there can be variable foregrounds or backgrounds, and there can be mistakes or variations in calibration or data processing.
-All of these things happen in real data sets, but the Forward Model (tm) is no more impacted by this than what is done in Standard Practice (tm)!
+All of these things happen in real data sets, but the Forward Model(tm) is no more impacted by this than what is done in Standard Practice(tm)!
 
@@
 The first is that the procedures we design are designed to deliver the exposure (or signal-to-noise-squared) weighted mean of the observed epochs.
-That is, the fact that the data are varying does not mean that the methodology needs to be modified; without modification, the procedures given above---both the Forward Model (tm) and the Standard Practice (tm)---will produce something like the mean spectrum.
+That is, the fact that the data are varying does not mean that the methodology needs to be modified; without modification, the procedures given above---both the Forward Model(tm) and the Standard Practice(tm)---will produce something like the mean spectrum.
+With this attitude, the spectral variations are just another source of noise.
 
@@
 The LSF adjustment is then applied to the spectral model before it is compared to each epoch of data, just as the shift is applied in the current methodological scope.
-Similar model adustments can account for intrinsic spectral variations, additive or multiplicative nuisance signals in the data (like backgrounds or telluric absorptions), and mistakes in data pre-processing (like continuum modeling errors).
+Similar model adjustments can account for intrinsic spectral variations, additive or multiplicative nuisance signals in the data (like backgrounds or telluric absorptions), and mistakes in data pre-processing (like continuum modeling errors).
 
-HOGG: What is the third attitude?
-
-The combined spectrum computed with the Forward Model(tm) has minimal pixel-to-pixel covariances based on repeated experiments.
-In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
-This is particularly evident in the second panel of Figure~\ref{fig:noise} near $\ln\lambda = 8.7024$.
-Here the pixel-to-pixel covariances resulting from Standard Practice(tm) are no longer academic, as the covariances here mimic weak spectral features.
-For example, if a weak Na line existed near this wavelength then a spectroscopist might readily interpret these covariances as real spectral features and report an enhanced Na abundance.
-Checking the strength of other (stronger) Na features would be a sensible check for this interpretation, but often there are few spectral lines available for a given element.
-
-HOGG Maybe consider the case of $P\sim\sum_i M_i$. Double descent and all? Or say this earlier in the method section?
-
-One side note about epoch-to-epoch variations in instrument or data-processing parameters:
-It is usually a good idea to model not the full instrument or data properties (as does \citealt{sp}) if that can be avoided, but rather just the \emph{differences} in instrument or data properties from one epoch to the next, taking those differences away from some fiducial data or instrument state.
-In the case of the LSF, for example, that obviates the need to model the infinite-resolution spectrum; the spectrum only needs to be modeled at fiducial resolution and convolved to other resolutions with difference kernels.
-
-Although \emph{conceptually} changes to account for data variability aren't that big, \emph{in practice} they can be very hard to implement.
+Although \emph{conceptually} it is not hard to account for such variabilities, \emph{in practice} they can make life hard.
 For one, the method we have presented assumes that we \emph{know} the individual epoch shifts \textsl{a priori}, and all new epoch-to-epoch adjustments would also depend on per-epoch housekeeping data or parameters that might, in practice, be hard to know.
@@
 For example, we could imagine forward-modeling generalizations of some of sparse methods (for example, \citealt{candes}) for applications in which the variability is localized in time or space; or we could imagine generalizations of matrix factorizations (for example, \citealt{hmf}) for applications in which variability is low dimensional (the \textsl{wobble} method has some such functionality; \citealt{wobble}).
-% HOGG: need candes and hmf citations
+
+The combined spectrum computed with the Forward Model(tm) has minimal pixel-to-pixel covariances based on repeated experiments.
+In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
+This is particularly evident in the second panel of Figure~\ref{fig:noise} near $\ln\lambda = 8.7024$.
+Here the pixel-to-pixel covariances resulting from Standard Practice(tm) are not purely of academic interest, as the covariances here mimic weak spectral features.
+For example, if a weak Na line existed near this wavelength then a spectroscopist might readily interpret these covariances as real spectral features and report an enhanced Na abundance.
+Checking the strength of other (stronger) Na features would be a sensible check for this interpretation, but often there are few spectral lines available for a given element.
+
+One side note about epoch-to-epoch variations in instrument or data-processing parameters:
+It is usually a good idea to model not the full instrument or data properties (as does \citealt{sp}) if that can be avoided, but rather just the \emph{differences} in instrument or data properties from one epoch to the next, taking those differences away from some fiducial data or instrument state.
+In the case of the LSF, for example, that obviates the need to model the infinite-resolution spectrum; the spectrum only needs to be modeled at fiducial resolution and convolved to other resolutions with difference kernels.
 
@@
 
-% What about non-Gaussian noise?
+We effectively assumed that the noise was Gaussian, and zero mean, but these are not always true of astronomical data.
+In principle even in the best-case scenarios there are Poisson noise sources, and cosmic rays and faulty detector pixels tend to add in highly skewed, non-zero-mean noise.
+The most important noise assumption made in all this is the zero-mean assumption; when this is wrong, all current spectral (or image) combination methods will be biased (including Standard Practice), unless they employ a likelihood function that accurately represents the detailed noise statistics.
 
-We assumed that the noise was Gaussian, but this is not always true of astronomical data.
-In the case of Poisson noise that is not background dominated, the variance tensor $C_i$ will be biased, and all our assumptions about additive Gaussian variance will be broken.
-% ARC: what else should we say about this?
-
-Code is available at \url{https://github.com/davidwhogg/NoDataInterpolation} and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
+All code used for this project is available at \url{https://github.com/davidwhogg/NoDataInterpolation} and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
 
@@
 \texttt{numpy} \citep{numpy} ---
+\texttt{scipy} \citep{scipy} ---
 \texttt{matplotlib} \citep{matplotlib}.
@@
 \bibitem[Bolton \& Schlegel(2010)]{sp} Bolton, A.~S. \& Schlegel, D.~J.\ 2010, \pasp, 122, 248. doi:10.1086/651008
+\bibitem[Candes \& Romberg(2007)]{candes} Candes, E., \& Romberg, J., 2007, Inverse problems, 23, 969.
+\bibitem[Fruchter \& Hook(2002)]{drizzle} Fruchter, A.~S. \& Hook, R.~N.\ 2002, \pasp, 114, 144. doi:10.1086/338393
 \bibitem[Harris et al.(2020)]{numpy} Harris, C.~R., Millman, K.~J., van der Walt, S.~J., et al.\ 2020, \nat, 585, 357. doi:10.1038/s41586-020-2649-2
+\bibitem[Hastie et al.(2022)]{doubledescent} Hastie, T., Montanari, A., Rosset, S., \& Tibshirani, R. J., 2022, Annals of Statistics, 50, 949. doi:10.1214/21-AOS2133
 \bibitem[Hewett et al.(1985)]{oe} Hewett, P.~C., Irwin, M.~J., Bunclark, P., et al.\ 1985, \mnras, 213, 971. doi:10.1093/mnras/213.4.
@@
 \bibitem[Majewski et al.(2017)]{apogee} Majewski, S.~R., Schiavon, R.~P., Frinchaboy, P.~M., et al.\ 2017, \aj, 154, 94. doi:10.3847/1538-3881/aa784d
+\bibitem[Tsalmantza \& Hogg(2012)]{hmf} Tsalmantza, P. \& Hogg, D.~W.\ 2012, \apj, 753, 122. doi:10.1088/0004-637X/753/2/122
+\bibitem[Virtanen et al.(2020)]{scipy} Virtanen, P., Gommers, R., Oliphant, T.~E., et al.\ 2020, Nature Methods, 17, 261. doi:10.1038/s41592-019-0686-2
+\bibitem[Voinov \& Nikulin(2012)]{mvue} Voinov, V. G., \& Nikulin, M. S., 2012, Unbiased Estimators and Their Applications: Volume 1: Univariate Case, Springer Science \& Business Media.
 \bibitem[Williams et al.(1996)]{hdf} Williams, R.~E., Blacker, B., Dickinson, M., et al.\ 1996, \aj, 112, 1335. doi:10.1086/118105

### 2024-03-09 paper/main.tex
this isn't interpolation; flux conservation?

@@
 % - discuss conservation of flux vs intensity.
-% - Blanton asks: Is this identical to sinc shift and average? Hogg thinks it MIGHT be.
-% - Add Hogg response to Blanton: Interpolate with (say) cubic spline and then average is NOT THE SAME as fit all the data simultaneously with a cubic spline function on a particular set of knots.
-% - Dustin asks: Is this effectively producing a sinc LSF? Hogg thinks it DEFINITELY IS NOT. Explain why not in the text.
 % - Hogg notes: Is it clear in the text that we are taking everything to a standard instrument?
@@
 % - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
-% - Audit for its / it's. Hogg has issues with this.
 % - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.
@@
 and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.
-This technology is encoded in tools such as \textsl{Drizzle} (\citealt{drizzle}) and HOGG.
+This technology is encoded in tools such as \textsl{Drizzle} (\citealt{drizzle}).
 
@@
 We return to these issues at the end. %% HOGG DO WE?
+
+\paragraph{Flux or photon conservation}
+HOGG SHOULD WE PUT WORDS HERE ABOUT WHY WE DON'T CARE ABOUT THIS CONSIDERATION? (1) normalized spectra. (2) It is intensity that a telescope imager measures.
 
@@
 It doesn't involve interpolating or otherwise distorting the input data;
-it produces a combined or mean spectrum in the format that traditional methods employ;
+it produces a combined or mean spectrum in the format that astronomers traditionally expect;
 it makes no assumptions about the form of the combined spectrum (except that it can be represented on an $M_\ast$-pixel grid);
 and it has excellent noise properties in real circumstances.
-The method can work on a wide variety of different kinds of data, but it is especially valuable in projects where the individual input exposures are ill-sampled, or not as well sampled as one might like.
+The method can work on a wide variety of different kinds of data, but it is especially valuable in projects where the individual input exposures are ill-sampled, or not as well sampled as one might like; our original motivation was the individual \textsl{APOGEE} spectrograph (\citealt{apogeehardware}) exposures.
 It is also good for situations in which the input data have heterogeneous bad pixels or data gaps.
+
+The method does not involve interpolating the data.
+However, it does fit a very flexible model to the data, and there is a limit in which fitting very flexible models is equivalent to interpolation (see, eg, \citealt{fittingflexible}).
+So is this method, implicitly, interpolating the data?
+Our answer is no.
+Interpolation of the data involves non-parametric models with control points that are set by the pixel grids of the input data.
+Our model choice never refers to the properties of the input data pixel grids; it is chosen based on the desired properties of the output, combined-spectrum grid.
+Furthermore, interpolation functions (like cubic spline, say, or Lanczos-5, say) pass through every raw-data pixel value exactly.
+There is no point in this method at which the function being fit passes through every (or even any) of the raw-data pixel values.
 

### 2024-03-09 paper/main.tex
desi paper
less aggressive introduction

@@
 % - Related Hogg issue: Any latent spectrum (like our combined spectrum) should be something that COULD BE OBSERVED by some instrument in the future. Hence spectro-perfectionism is wrong (ish).
-% - Read and cite the DESI spectral pipeline paper; Dustin says it's relevant.
 % - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
@@
 \renewcommand{\paragraph}[1]{\medskip\par\noindent\textbf{#1}~---}
+\sloppy\sloppypar\raggedbottom\frenchspacing
 
@@
 
-% notes
-
-\sloppy\sloppypar\raggedbottom\frenchspacing
 \begin{document}
@@
 \affiliation{Max-Planck-Institut f\"ur Astronomie, Heidelberg}
-\affiliation{Flatiron Institute, a division of the Simons Foundation}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute}
 
@@
 
-\keywords{Foo --- Bar --- HOGG}
+\keywords{HOGG --- Foo --- Bar}
 
-\section*{}\clearpage
 \section{Introduction}\label{sec:intro}
@@
 and the future Rubin Observatory \textsl{LSST} (\citealt{lsst}), among countless other projects.
-This technology is encoded in tools such as \textsl{Drizzle} (\citealt{drizzle}).
+This technology is encoded in tools like \texttt{Drizzle} (\citealt{drizzle}).
 
@@
 That is, a combined image is a kind of \emph{prediction}: It is the prediction of what \emph{would have been observed} with a much longer exposure with a much better device.
-In measurement theory, the practice is to combine noisy data into a measurement by building and optimizing a likelihood, not by shifting and stacking (or performing other arithmetic operations on) the input data.
+In measurement theory, the practice is to combine noisy data into a measurement by building and optimizing a likelihood.
 Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
-It is just a simultaneous estimate of a lot of different components (a lot of different pixels; all the pixels in the combined image or spectrum).
+It is just a simultaneous estimate of a lot of different components---a lot of different pixels---all the pixels in the combined image or spectrum.
 
@@
 Spectral combination methods that distort absorption (or emission) lines will translate into biases, errors, or increased variance in these measurements.
-Spectral combination methods that correlate the noise in the outputs will translate into increased variance, and necessitate more sophisticated downstream measurement techniques (which don't really exist at present).
+Spectral combination methods that correlate the noise in the outputs will translate into increased variance, and necessitate more sophisticated downstream measurement techniques.
 
@@
 Indeed it has just as much freedom as the shifting-and-stacking standard practice.
-The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation, by means of a justified likelihood function.
+The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation, by means of a justifiable likelihood function.
 
 A boundary condition of the work presented here is that the functional requirements of anything we provide must be the same as with standard practice:
-That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates, bad-pixel masks, and perhaps bit-flags; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
+That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
 That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
 Unlike the shift-and-coadd approach, the estimator we will deliver will be insensitive to under- or poorly-sampled data, which is not uncommon when astronomical seeing accidentally gets excellent, or in spectrographs that have limited detector pixels (for example, \citealt{apogeehardware}).
-As we will show, an estimator also often yields a more faithful estimate of the \emph{true} spectrum, and almost always better noise properties.
-Both of these results make our results better for any subsequent precision measurements (e.g., radial velocities, chemical abundances).
-But importantly, we aren't proposing a change to the \emph{format} of the input or outputs of spectral or image combination.
+As we will show with artificial data, the estimator we propose yields a faithful estimate of the \emph{true} spectrum, and usually with good noise properties.
+Both of these results make our results good for subsequent precision measurements (e.g., radial velocities, chemical abundances).
 
-% Hogg's email epiphany:
-% 1.
-I think I figured out what Blanton means by saying that interpolation of the data is a kind of fitting.
-It is!
-But those fits can’t be combined well because the fits are infinitely over-parameterized.
-If the interpolation was done approximately with a finite-parameter fit, we could interpolate and then average just fine.
-I’m writing this here to remind myself of this epiphany for when I come back online and write a subsection about this in the paper.
-
-% Hogg: what do you want to do about this? HOGG SAY: I want to explicitly address it, maybe in a first section after the Introduction. This introduction is too long anyway!
-
-The approach we propose differs from interpolation in a subtle way. 
+The approach we propose differs from interpolation-and-averaging in a subtle way. 
 Interpolation could be described as computing new values at arbitrary positions based on an existing set of known data points. 
-Instead, here we propose to construct a continuous spectral model for the data, where we then evaluate that model on the given set of output positions (pixels).
-The \emph{output product} is the same form as what one would expect from an interpolation procedure (e.g., computed values along some array of positions), but the \emph{procedure} differs from interpolation.
-When interpolating data the basis functions are explicitly constructed from the data, whereas here we construct a forward model that is maximally consistent with the data, and evaluated at new positions.
+Instead, here we propose to construct a continuous spectral model for the data, such that we can evaluate that model on either the input (raw-data) pixel grid or else on the output (combined spectrum) pixel grid.
+The \emph{output product} is the same form as what one would expect from an interpolation-and-averaging procedure (values at a set of pixel positions), but the \emph{procedure} differs from interpolation.
+When interpolating the input data the basis functions are explicitly constructed from the input-data pixel grid, whereas here we construct a forward model that is maximally consistent with the data, and can be evaluated at any pixel locations.
 
@@
 For another, some projects have sought to co-add background-dominated images in a near optimal way, but presume that all observations are already registered and re-sampled to the same grid \citep{zackay}.
+For another, the \textsl{DESI} survey ensures that all spectra are extracted on the same wavelength grid, to avoid any interpolation or resampling \citep{desipipeline}.
 What we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models.
@@
 Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
-One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is a better method that moves this spectral combination method directly into the two-dimensional data.
+One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is almost certainly a better method that moves this spectral combination method directly into the two-dimensional data; that is, there ought to be a method that obviates one-dimensional spectral extraction entirely.
 We hope for a future in which one-dimensional spectra are never extracted from two-dimensional images; in this future the two-dimensional data are the sacred data, not the extractions.
@@
 In real applications, the spectra we extract are often of variable sources, or taken at different position angles on extended objects, such that we don't expect each epoch to be identical.
-Similarly, the \textsl{LSST} data are being taken the way they because the sky (and point-spread function) will vary.
-That means that we are only considering here the very easiest case.
-However, we note that all standard shift-and-coadd methods make these non-variable assumptions implicitly.
-Part of our contribution in this \documentname{} is to make implicit assumptions explicit, in \sectionname~\ref{sec:assumptions}.
+Similarly, the \textsl{LSST} data are being taken the way they are because the sky (and point-spread function) will vary.
+Hence, in assuming that we only want the time-invariant mean of our data, we are considering here the very easiest problem in a range of possible problems.
+However, we note that all standard shift-and-coadd methods make similar assumptions implicitly.
+Part of our contribution in this \documentname{} is to make implicit assumptions explicit; we do this in \sectionname~\ref{sec:assumptions}.
 
@@
 for valuable discussions and input.
+The Flatiron Institute is a division of the Simons Foundation.
+
+HOGG: DID WE USE SDSS DATA?
+
+ARC: DO YOU HAVE GRANTS TO THANK?
 
@@
 \bibitem[Fruchter \& Hook(2002)]{drizzle} Fruchter, A.~S. \& Hook, R.~N.\ 2002, \pasp, 114, 144. doi:10.1086/338393
+\bibitem[Guy et al.(2023)]{desipipeline} Guy, J., Bailey, S., Kremin, A., et al.\ 2023, \aj, 165, 144. doi:10.3847/1538-3881/acb212
 \bibitem[Harris et al.(2020)]{numpy} Harris, C.~R., Millman, K.~J., van der Walt, S.~J., et al.\ 2020, \nat, 585, 357. doi:10.1038/s41586-020-2649-2
@@
 \bibitem[Wolf et al.(2003)]{combo17} Wolf, C., Meisenheimer, K., Rix, H.-W., et al.\ 2003, \aap, 401, 73. doi:10.1051/0004-6361:20021513
+\bibitem[Zackay, B. \& Ofek (2017)]{zackay} Zackay, B. \& Ofek, E. O.\ 2017, \apj, 836, 188. doi:10.3847/1538-4357/836/2/188
 \bibitem[Zechmeister et al.(2014)]{froe} Zechmeister, M., Anglada-Escud{\'e}, G., \& Reiners, A.\ 2014, \aap, 561, A59. doi:10.1051/0004-6361/201322746
-\bibitem[Zackay, B. \& Ofek (2017)]{zackay} Zackay, B. \& Ofek, E. O.\ 2017, \apj, 836, 188. doi:10.3847/1538-4357/836/2/188
 

### 2024-03-12 paper/main.tex
Update main.tex
- very minor word edits (typos)
- wrong figure being referenced
- add ack

@@
 A boundary condition of the work presented here is that the functional requirements of anything we provide must be the same as with standard practice:
-That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity of flux values on a pixel grid, along with uncertainty estimates and metadata flags.
-That is, the only difference between standard practice and the forward model that we advocate is the purely in the nature of the estimator.
+That is, we consider here only procedures that start with individual, single-epoch observations as inputs, along with noise estimates and bad-pixel masks; and we consider only procedures that end with an output ``combined image'' or ``combined spectrum'' that is a set of intensity or flux values on a pixel grid, along with uncertainty estimates and metadata flags.
+That is, the only difference between standard practice and the forward model that we advocate is purely in the nature of the estimator.
 Unlike the shift-and-coadd approach, the estimator we will deliver will be insensitive to under- or poorly-sampled data, which is not uncommon when astronomical seeing accidentally gets excellent, or in spectrographs that have limited detector pixels (for example, \citealt{apogeehardware}).
@@
 
-In many cases the LSF or PSF is actually substantially different in different exposures.
+In many cases the LSF or PSF is actually substantially different between exposures.
 This leads to spectral (or image) variations, and violate the assumptions of this model (and also the implicit assumptions underlying standard practice).
@@
 That is, we are assuming that the changes from observation to observation are solely from either the shifts $\Delta x_i$ or else from effects that we consider to be noise.
-We interpolate the model to account for the shifts, and we combine data to average away this noise.
+We interpolate the model to account for the shifts, and we combine data to average away this noise. % ARC to HOGG: is 'average away' what you want to say here?
 In many cases the stellar variability is itself the subject of study, and then the variability is not entirely noise.
@@
 
-Whenever performing\texttt{inv()}, \texttt{solve()}, or \texttt{lstsq()} operations, it is sensible to begin by checking the condition number (the ratio of the largest to the smallest singular value) of the matrix in play.
+Whenever performing \texttt{inv()}, \texttt{solve()}, or \texttt{lstsq()} operations, it is sensible to begin by checking the condition number (the ratio of the largest to the smallest singular value) of the matrix in play.
 Condition numbers larger than $10^8$ are usually trouble, even when the mathematics is being performed at double precision.
@@
 In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
-This is particularly evident in the second panel of Figure~\ref{fig:noise} near $\ln\lambda = 8.7024$.
+This is particularly evident in the second panel of Figure~\ref{fig:standard} near $\ln\lambda = 8.7024$.
 Here the pixel-to-pixel covariances resulting from Standard Practice(tm) are not purely of academic interest, as the covariances here mimic weak spectral features.
@@
 
-All code used for this project is available at \url{https://github.com/davidwhogg/NoDataInterpolation} and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
+All code used for this project is available at \url{https://github.com/davidwhogg/NoDataInterpolation}.% ANDY to update terminology in code.
+and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
 
@@
 
-ARC: DO YOU HAVE GRANTS TO THANK?
+Parts of this research were supported by the Australian Research Council Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D), through project number CE170100013.
 

### 2024-03-13 paper/main.tex
killing HOGG notes; flux conservation; diagonal noise

--- 2024-03-09 paper/main.tex
@@
-% Copyright 2022, 2023, [add your years here] David W. Hogg (NYU) and Andy Casey (Monash)
+% Copyright 2022, 2023, 2024 David W. Hogg (NYU) and Andy Casey (Monash)
 
@@
 % - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
-% - Make comments directly about Drizzle to justify the name.
 % - discuss conservation of flux vs intensity.
@@
 Discussion point?
-% - Related Hogg issue: Any latent spectrum (like our combined spectrum) should be something that COULD BE OBSERVED by some instrument in the future. Hence spectro-perfectionism is wrong (ish).
-% - Audit for rigid terminology. HOGG is currently liking ``combined spectrum'' for the model expectation, for example. And "standard practice" for the traditions of our field.
-% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code?
+% - Related Hogg issue: Any latent spectrum (like our combined spectrum) should be something that COULD BE OBSERVED by some instrument in the future.
+Hence spectro-perfectionism is wrong (ish).
+Add this to the discussion?
+% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code? HOGG
 % - Write about the two input-data cases.
@@
 
-\keywords{HOGG --- Foo --- Bar}
+\keywords{data analysis --- image processing --- spectroscopy --- statistics}
 
@@
 However, they will---in some cases---sacrifice some information.
-Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{mvue}). % HOGG
+Minimum-variance unbiased estimators require care in their construction; they are generally maximum-likelihood estimators (\citealt{mvue}).
 The deep, high signal-to-noise image or spectrum we seek is a statistical estimate of some kind of ``true'' or high signal-to-noise image or spectrum, given the noisy data of the individual exposures.
@@
 In measurement theory, the practice is to combine noisy data into a measurement by building and optimizing a likelihood.
-Of course we don't always think of an image or a spectrum of the sky as the result of applying an \emph{estimator}, but it is.
+Of course we don't always think of a combined (or mosaiced) image of the sky or of an average spectrum of a star as the result of applying an \emph{estimator}, but it is.
 It is just a simultaneous estimate of a lot of different components---a lot of different pixels---all the pixels in the combined image or spectrum.
@@
 Spectral combination methods that distort absorption (or emission) lines will translate into biases, errors, or increased variance in these measurements.
-Spectral combination methods that correlate the noise in the outputs will translate into increased variance, and necessitate more sophisticated downstream measurement techniques.
+Spectral combination methods that correlate the noise in the output combined spectrum will translate into increased variance, and necessitate more sophisticated downstream measurement techniques.
 
@@
 One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is almost certainly a better method that moves this spectral combination method directly into the two-dimensional data; that is, there ought to be a method that obviates one-dimensional spectral extraction entirely.
-We hope for a future in which one-dimensional spectra are never extracted from two-dimensional images; in this future the two-dimensional data are the sacred data, not the extractions.
+We hope for a future in which one-dimensional spectra are never extracted from two-dimensional images; in this future the two-dimensional data will be the sacred data, not the extractions.
 
@@
 It might surprise the reader to learn that we prefer to let the model capture higher frequencies, even when we think they shouldn't be there for any reasons other than noise, given the hardware.
-More on this below. %% HOGG DO WE COME BACK TO THIS?
+The reason we provide our model the flexibility to go to higher frequencies is that we want our combined spectrum to have the capacity to capture or represent the pixel-to-pixel noise in the estimate; this pixel-to-pixel noise obeys no band limit.
 
-% HOGG SAY: IT is not at all clear here that we are NOT assuming critical sampling. We aren't! Make that clear.
-% In general this section has issues.
 \paragraph{Critical sampling}
@@
 And indeed, there are spectrographs that don't have critically sampled pixel scales (for example, \textsl{APOGEE}; \citealt{apogee}).
-Thus we will not assume that the raw data are critically sampled, but we will produce the output mean spectrum on a grid that is critically sampled.
+Thus we will not assume that the raw data are critically sampled, but we will produce the output combined spectrum on a grid that is critically sampled with respect to the instrument resolution.
 The input and output pixel grids do not need to coincide when the model is a continuous function.
@@
 For another, if there are normalization variations, a normalization model can be added to the image combination procedure.
-We return to these issues at the end. %% HOGG DO WE?
+That is, if different input spectra have different normalizations, the model for the combined spectrum can be multiplied by a normalization model, unique to each input spectrum, prior to evaluation of the likelihood.
 
 \paragraph{Flux or photon conservation}
-HOGG SHOULD WE PUT WORDS HERE ABOUT WHY WE DON'T CARE ABOUT THIS CONSIDERATION? (1) normalized spectra. (2) It is intensity that a telescope imager measures.
+Some methods for image combination, like \texttt{Drizzle}, have as a hard constraint that they conserve \emph{flux} or \emph{photons}.
+We do not agree with this constraint or condition.
+The fundamental observable in a telescope is intensity, or energy per area per time per wavelength (or per frequency).
+This quantity is related to the phase-space density of photons.
+When the pixel grid of an imager is changed, the expectation of the rate of photon arrivals per pixel will change, but the expectation of the intensity will not change.
+Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized measurements of intensity) at the detector surface.
+The combined or average image should be the provide the best estimate of this (possibly normalized) intensity.
 
@@
 \end{align}
-where $X_\star$ is the $M_\star\times P$ evaluation of all of the $P$ basis functions at the locations of the $M_\star$ pixels in the output-spectrum pixel grid $x_\star$.
+where $X_\star$ is the $M_\star\times P$ evaluation of all of the $P$ basis functions at the locations of the $M_\star$ pixels in the output combined-spectrum pixel grid $x_\star$.
 Putting these things together, the output spectrum $y_\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
@@
 \begin{align}
-    C_\star &= [X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top]^{-1} ~.
+    C_\star &= [X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top]^{-1} ~.\label{eq:outcovar}
 \end{align}
@@
 Although the equations like \eqref{eq:xtxinvxty} show inverses, it is almost never a good idea to use the native linear-algebra inverse function \texttt{inv()} in any implementation of this method.
-The reason is that we don't want the inverse matrix itself, we want the outcome of the inverse matrix multiplied by $X^\top\,C^{-}\,Y$.
+The reason is that we don't want the inverse matrix itself, we want the outcome of the inverse matrix multiplied by $X^\top\,C^{-1}\,Y$.
 For this use case, it is more precise and more more numerically stable to use the linear algebra \texttt{solve()} or \texttt{lstsq()} functions.
@@
 These have impacts on how the code is written.
-In particular the fact that the output spectrum $y_\ast$ is real-valued constrains the complex Fourier amplitudes.
+In particular the fact that the output combined spectrum $y_\ast$ is real-valued constrains the complex Fourier amplitudes.
 Continuing on the Fourier theme:
@@
 The interpolated spectra $y'_i$ are, by construction, all on the same rest-frame wavelength grid $x_\star$ so they can be averaged, with a straight average, a median, an inverse-variance-weighted mean, or any more sophisticated algorithm.
-Here we choose an inverse-variance-weighted mean (HOGG: NO WE DON'T), but zeroing out the inverse variance at the locations of the bad pixels.
+Here we choose an unweighted mean, with the exception that we zero out the contributions of any bad pixels in that mean.
 
@@
 \paragraph{Noise and noise covariances}
-
 We repeated the above experiments to empirically measure the noise covariances, and to estimate how correlated neighboring pixels are in the output spectrum.
@@
 
-HOGG: Add intuition here about WHY the output noise model is close to uncorrelated?
-HOGG: Note that if we choose $P<M_\ast$ we start to see noise correlations in the Forward Model(tm) too.
-That's one of a few reasons we like $P=M_\ast$.
+It is interesting to ask \emph{why} the output combined spectrum in the Forward Model(tm) case has uncorrelated noise.
+It won't in general: If the number of epochs is small, or the coverage in the wavelength domain is spotty, there will in general be pixel-to-pixel covariances in the uncertainties on the combined spectrum.
+However, if the number of epochs is large and the distribution of shift operators $\Delta x_i$ is good, as it is in these experiments, the off-diagonal elements of the output combined-spectrum covariance matrix given in \eqref{eq:outcovar} will tend towards zero.
+The Standard Practice(tm) output combined spectrum always shows correlated noise, and especially when the input data are ill-sampled.
+The reason for this is that interpolation is very sensitive to noise in the ill-sampled case, such that the inter-pixel value of the interpolated function is moved substantially by small movements in the input pixel values, and these inter-pixel regions in the raw data are large relative to the pixel spacing in the combined spectrum spectral grid.
 

### 2024-03-15 paper/main.tex
tiny

@@
 \paragraph{Flux or photon conservation}
-Some methods for image combination, like \texttt{Drizzle}, have as a hard constraint that they conserve \emph{flux} or \emph{photons}.
+Some methods for image combination, like \texttt{Drizzle} \citep{drizzle}, have as a hard constraint that they conserve \emph{flux} or \emph{photons}.
 We do not agree with this constraint or condition.
@@
 When the pixel grid of an imager is changed, the expectation of the rate of photon arrivals per pixel will change, but the expectation of the intensity will not change.
-Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized measurements of intensity) at the detector surface.
-The combined or average image should be the provide the best estimate of this (possibly normalized) intensity.
+Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.
+The combined or average image should be the provide the best estimate of this intensity.
 

### 2024-03-15 paper/main.tex

@@
 % -----------
-% - What's our name? Choose and propagate into where "Forward Model(tm)" is used.
-% - Put in notes about our hypothesis for WHY we get uncorrelated pixels.
-% - discuss conservation of flux vs intensity.
+% - What's our name? Frizzle? Choose and propagate into where "Forward Model(tm)" is used.
 % - Hogg notes: Is it clear in the text that we are taking everything to a standard instrument?
@@
 We do not agree with this constraint or condition.
-The fundamental observable in a telescope is intensity, or energy per area per time per wavelength (or per frequency).
+The fundamental observable in a telescope is intensity, or energy per area per solid angle per time per wavelength (or per frequency).
 This quantity is related to the phase-space density of photons.
-When the pixel grid of an imager is changed, the expectation of the rate of photon arrivals per pixel will change, but the expectation of the intensity will not change.
+When the mapping between a pixel grid and the sky is changed, the expectation of the rate of photon arrivals per pixel will change (because the pixel effective solid angles change), but the expectation of the intensity will not change.
 Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.

### 2024-03-15 paper/main.tex
tiny edits and notes to self; ready to submit

--- 2024-03-15 paper/main.tex
@@
 % -----------
-% - What's our name? Frizzle? Choose and propagate into where "Forward Model(tm)" is used.
-% - Hogg notes: Is it clear in the text that we are taking everything to a standard instrument?
+% - HOGG check that the code checks out and "just runs".
+% - Is it clear in the text that we are taking everything to a standard instrument?
 And the user can decide what that instrument is?
@@
 Add this to the discussion?
-% - Audit and fix the sign applied to $\Delta x_i$ everywhere. Wrong relative to code? HOGG
-% - Write about the two input-data cases.
-% - See bug list in code for other issues with the text--code relationship (noise model; line-formation model).
-% - Deal with all HOGG, ANDY, ARC, XXX, YYY, CITE, etc.
+% - HOGG: Is the sign of Delta x_i wrong in the code? It is correct in the document now.
+% - See bug list in code for other issues with the text--code relationship (noise model; line-formation model). HOGG
+% - submit to AAS Journals
+% - post to arXiv
 
@@
 
-\title{\large Fowizzle: Combining spectra or images by forward modeling}
+\title{\large \textsl{Frizzle}: Combining spectra or images by forward modeling}
 
@@
 That is, we are assuming that the changes from observation to observation are solely from either the shifts $\Delta x_i$ or else from effects that we consider to be noise.
-We interpolate the model to account for the shifts, and we combine data to average away this noise. % ARC to HOGG: is 'average away' what you want to say here?
+We interpolate the model (not the data!) to account for the shifts, and we combine data to average down this noise.
 In many cases the stellar variability is itself the subject of study, and then the variability is not entirely noise.
 But from the perspective of this project, any stellar variability is considered to be part of the total additive noise in the spectrum.
+We return to this subject below in the Discussion.
 
@@
 \begin{align}
-    y_i &= f(x_i + \Delta x_i;\theta) + \mbox{noise} \\
+    y_i &= f(x_i - \Delta x_i;\theta) + \mbox{noise} \\
     f(x;\theta) &= \sum_{j=1}^P a_j\,g_j(x) ~,
 \end{align}
-where we are being a bit loose with terminology (since $x_i$ is a set of positions, not just one position), the noise is additive (but not yet specified), and each of the $P$ functions $g_j(x)$ is a basis function of the representation.
+where we are being a bit loose with terminology (since $y_i$ is a set of fluxes, and $x_i$ is a set of positions, not just one position), the noise is additive (but not yet specified), and each of the $P$ functions $g_j(x)$ is a basis function of the representation.
 
@@
 where $X_\star$ is the $M_\star\times P$ evaluation of all of the $P$ basis functions at the locations of the $M_\star$ pixels in the output combined-spectrum pixel grid $x_\star$.
-Putting these things together, the output spectrum $y_\star$ can be thought of as just a kind of linear shifting-and-averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
+Putting these things together, the output spectrum $y_\star$ can be thought of as just a kind of linear weighted averaging operation on the input spectra $y_i$ (which have been packed into $Y$) according to
 \begin{align}
@@
 But, \emph{very importantly}, this is only a correct usage when the input $C_i$ matrices really represented the true uncertainties going in.
-If you have any concerns about that, then it is way safer to get your uncertainty estimates from bootstrap or jackknife (\citealt{fittingflexible}).
+If you have any concerns about that, then it is way safer to get your uncertainty estimates from bootstrap or jackknife (for a related example, see \citealt{fittingflexible}).
 
@@
 For this use case, it is more precise and more more numerically stable to use the linear algebra \texttt{solve()} or \texttt{lstsq()} functions.
+The only use for \texttt{inv()} is where the full inverse of a matrix is required, as when estimating the full $C_\star$ by \eqref{eq:outcovar}.
 
@@
 These have impacts on how the code is written.
-In particular the fact that the output combined spectrum $y_\ast$ is real-valued constrains the complex Fourier amplitudes.
+In particular the fact that the output combined spectrum $y_\star$ is real-valued constrains the complex Fourier amplitudes.
 Continuing on the Fourier theme:
@@
 
-We recommended setting $P=M_\ast$.
+We recommended setting $P=M_\star$.
 We recommend very strongly \emph{against} setting $P\approx M$, where $M=\sum_i M_i$.
@@
 The reason for this is the ``double descent'' phenomenon, which de-stabilizes linear fitting (\citealt{doubledescent}).
-If you find yourself in the situation that $M_\ast\approx M$, reconsider your life choices.
+If you find yourself in the situation that $M_\star\approx M$, reconsider your life choices.
 
@@
 it produces a combined or mean spectrum in the format that astronomers traditionally expect;
-it makes no assumptions about the form of the combined spectrum (except that it can be represented on an $M_\ast$-pixel grid);
+it makes no assumptions about the form of the combined spectrum (except that it can be represented on an $M_\star$-pixel grid);
 and it has excellent noise properties in real circumstances.
@@
 In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
-This is particularly evident in the second panel of Figure~\ref{fig:standard} near $\ln\lambda = 8.7024$.
+This is particularly evident in the second panel of Figure~\ref{fig:standard} (say near $\ln\lambda = 8.7024$).
 Here the pixel-to-pixel covariances resulting from Standard Practice(tm) are not purely of academic interest, as the covariances here mimic weak spectral features.
@@
 The Flatiron Institute is a division of the Simons Foundation.
-
-HOGG: DID WE USE SDSS DATA?
-
-Parts of this research were supported by the Australian Research Council Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D), through project number CE170100013.
+This research was supported in part by the Australian Research Council Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D), through project number CE170100013.
 

### 2024-03-16 paper/main.tex

@@
 % -----------
-% - HOGG check that the code checks out and "just runs".
 % - Is it clear in the text that we are taking everything to a standard instrument?
@@
 Hence spectro-perfectionism is wrong (ish).
-Add this to the discussion?
+Add this to the Discussion?
 % - HOGG: Is the sign of Delta x_i wrong in the code? It is correct in the document now.
@@
 % text macros
-\shorttitle{don't interpolate the data}
+\shorttitle{combining spectra by forward modeling}
 \shortauthors{hogg \& casey}
@@
 \newcommand{\sectionname}{Section}
+\newcommand{\secref}[1]{\sectionname~\ref{#1}}
+\newcommand{\name}{\textsl{Frizzle}}
 
@@
 
-\title{\large \textsl{Frizzle}: Combining spectra or images by forward modeling}
+\title{\large \name: Combining spectra or images by forward modeling}
 
@@
 For another, the \textsl{DESI} survey ensures that all spectra are extracted on the same wavelength grid, to avoid any interpolation or resampling \citep{desipipeline}.
-What we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models.
-Plus it learns (it isn't told) the epoch-to-epoch spectral shifts.
+What we propose below is actually a sub-problem of the problem solved by the \textsl{wobble} method for determining stellar radial velocities (\citealt{wobble}); \textsl{wobble} is more general because it also includes a tellurics model and also tellurics and stellar variability models,
+plus it learns (it isn't told) the epoch-to-epoch spectral shifts.
 There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{kelson}, \citealt{froe}, \citealt{sp}):
 Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
-One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is almost certainly a better method that moves this spectral combination method directly into the two-dimensional data; that is, there ought to be a method that obviates one-dimensional spectral extraction entirely.
+One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is almost certainly an even better method that moves this spectral combination method directly into the two-dimensional data; that is, there ought to be a method that obviates one-dimensional spectral extraction entirely.
 We hope for a future in which one-dimensional spectra are never extracted from two-dimensional images; in this future the two-dimensional data will be the sacred data, not the extractions.
@@
 However, we note that all standard shift-and-coadd methods make similar assumptions implicitly.
-Part of our contribution in this \documentname{} is to make implicit assumptions explicit; we do this in \sectionname~\ref{sec:assumptions}.
+Part of our contribution in this \documentname{} is to make implicit assumptions explicit; we do this in \secref{sec:assumptions}.
 
@@
 We will assume here that (from some external information) we know these shift operators very accurately.
-This is often not true, of course; we will return to this point in the discussion.
+This is often not true, of course; we will return to this point in \secref{sec:discussion} below.
 
@@
 This leads to spectral (or image) variations, and violate the assumptions of this model (and also the implicit assumptions underlying standard practice).
-We won't address this kind of variability here, except in some discussion at the end:
+We won't address this kind of variability here, except for some comments in \secref{sec:discussion}:
 The method presented here can be straightforwardly modified to account for LSF or PSF variations.
@@
 But from the perspective of this project, any stellar variability is considered to be part of the total additive noise in the spectrum.
-We return to this subject below in the Discussion.
+We return to this subject below in \secref{sec:discussion}.
 
@@
 Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.
-The combined or average image should be the provide the best estimate of this intensity.
+The combined image should provide a good mean estimate of this intensity.
 
-\section{Method}\label{sec:method}
+\section{Method: \name}\label{sec:method}
 
@@
 
-The spectral expectation model $f(x;\theta)$, which will be a representation for our combined spectrum, is a function of $x$ controlled by a list $\theta$ of $P$ linear parameters $a_j$ (with $1\leq j\leq P$).
+The spectral expectation model $f(x;\theta)$, which will be a representation of our combined spectrum, is a function of $x$ controlled by a list $\theta$ of $P$ linear parameters $a_j$ (with $1\leq j\leq P$).
 This function will be very flexible; possibly even an interpolation of control points or a Fourier series; the number $P$ will be roughly the number of pixels we want in our output combined spectrum.
@@
 \paragraph{Fake data}
-We demonstrate the effectiveness of the spectral combination method on artificial data.
+We demonstrate the effectiveness of \name{}---the spectral combination method by forward modeling---on artificial data.
 The spectral data were generated from a mean spectral expectation that is a continuum plus a randomly generated line list, where each line equivalent width was set by a random draw from a power law. 
@@
 
-\paragraph{Forward Model(tm)}
-As our main experiment or result, we apply the Forward Model(tm) as described above in \sectionname~\ref{sec:method} to the input data shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
+\paragraph{\name{}}
+As our main experiment or result, we apply the forward model described above in \secref{sec:method} to the input data shown in \figurename~\ref{fig:data1} and \figurename~\ref{fig:data2}.
 In each case, the output pixel grid $x_\star$ is chosen to be well sampled, with a pixel spacing of $1/R$ in the (natural) logarithm of wavelength.
@@
     \end{center}
-    \caption{The top two panels show the result of running the Forward Model(tm) on the poorly sampled input data, plus the residual away from the truth.
+    \caption{The top two panels show the result of running \name{} on the poorly sampled input data, plus the residual away from the truth.
 The bottom two panels show the same but for the well sampled input data.
@@
 \end{figure}
-The combined spectrum resulting from the Forward Model(tm) described in \sectionname~\ref{sec:method} is shown in \figurename~\ref{fig:forward}.
+The combined spectrum resulting from the forward model described in \secref{sec:method} is shown in \figurename~\ref{fig:forward}.
 It is compared to the true spectral expectation model that was used to generate the data.
-Also shown in \figurename~\ref{fig:forward} is the residuals, Forward Model(tm) minus truth.
+Also shown in \figurename~\ref{fig:forward} is the residuals, \name{} minus truth.
 The residuals look stationary, approximately Gaussian, and uncorrelated (more on this below).
@@
 We repeated the above experiments to empirically measure the noise covariances, and to estimate how correlated neighboring pixels are in the output spectrum.
-For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using the Forward Model(tm) and Standard Practice(tm).
+For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using \name{} and Standard Practice(tm).
 We then computed the empirical covariance in the output spectrum minus the true spectrum, which are shown in \figurename~\ref{fig:noise}.
 In each case, Standard Practice(tm) has a lower noise variance at zero lag, but shows spatially correlated noise over many pixels.
-The Forward Model(tm), in contrast, shows no pixel-to-pixel noise covariances, even when the input data are ill sampled.
+\name{}, in contrast, shows no pixel-to-pixel noise covariances, even when the input data are ill sampled.
 \begin{figure}[t!]
@@
 The variances were estimated by performing the experiments repeated 64 times, with the same spectral expectations and sampling but unique noise draws.
-The Forward Model(tm) has slightly larger variance per pixel at zero lag, but vanishing pixel-to-pixel correlations.
+\name{} has slightly larger variance per pixel at zero lag, but vanishing pixel-to-pixel correlations.
 In contrast, the Standard Practice(tm) leads to correlated pixel values, with correlations extending out to many pixels.\label{fig:noise}}
@@
 
-It is interesting to ask \emph{why} the output combined spectrum in the Forward Model(tm) case has uncorrelated noise.
+It is interesting to ask \emph{why} the output combined spectrum in the forward-modeling case shows uncorrelated noise.
 It won't in general: If the number of epochs is small, or the coverage in the wavelength domain is spotty, there will in general be pixel-to-pixel covariances in the uncertainties on the combined spectrum.
@@
 
-We have presented a forward-modeling approach to combining spectra or images.
-This method has good properties:
+We have presented a forward-modeling approach to combining spectra or images;
+we name it \name.
+\name{} has good properties:
 It doesn't involve interpolating or otherwise distorting the input data;
@@
 
-In the Introduction, we were agnostic about whether we were talking about imaging or spectroscopic data.
+In the Introduction (\secref{sec:intro}), we were agnostic about whether we were talking about imaging or spectroscopic data.
 In the end, our examples are spectroscopic.
@@
 
-Everything presented above depends on a set of strong assumptions, listed in \sectionname~\ref{sec:assumptions}.
+Everything presented above depends on a set of strong assumptions, listed in \secref{sec:assumptions}.
 Perhaps the strongest assumption is that the spectrum or image is not a function of time.
@@
 The source can vary, the instrument can vary, there can be variable foregrounds or backgrounds, and there can be mistakes or variations in calibration or data processing.
-All of these things happen in real data sets, but the Forward Model(tm) is no more impacted by this than what is done in Standard Practice(tm)!
+All of these things happen in real data sets, but \name{} is no more impacted by this than Standard Practice(tm)!
 
@@
 The first is that the procedures we design are designed to deliver the exposure (or signal-to-noise-squared) weighted mean of the observed epochs.
-That is, the fact that the data are varying does not mean that the methodology needs to be modified; without modification, the procedures given above---both the Forward Model(tm) and the Standard Practice(tm)---will produce something like the mean spectrum.
+That is, the fact that the data are varying does not mean that the methodology needs to be modified; without modification, the procedures given above---both \name{} and Standard Practice(tm)---will produce something like the mean spectrum.
 With this attitude, the spectral variations are just another source of noise.
@@
 
-The combined spectrum computed with the Forward Model(tm) has minimal pixel-to-pixel covariances based on repeated experiments.
-In contrast, the combined spectrum computed with Standard Practice(tm) has substantial pixel-to-pixel covariances.
+The combined spectrum created by \name{} has minimal pixel-to-pixel covariances based on repeated experiments.
+In contrast, the combined spectrum created by Standard Practice(tm) has substantial pixel-to-pixel covariances.
 This is particularly evident in the second panel of Figure~\ref{fig:standard} (say near $\ln\lambda = 8.7024$).
@@
 Adrian Price-Whelan (Flatiron),
-and the Astronomical Data Group at the Flatiron Institute,
+and the Astronomical Data Group at the Flatiron Institute
 for valuable discussions and input.

### 2024-03-18 paper/main.tex
new figure locations
changes for AAS submisssion

@@
 
-\documentclass[modern]{aastex631}
+\documentclass[modern, linenumbers]{aastex631}
 \usepackage[utf8]{inputenc}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=1.3\figurewidth]{notebooks/data1.pdf}\\
-    \includegraphics[width=\figurewidth]{notebooks/datazoom.pdf}
+    \includegraphics[width=1.3\figurewidth]{data1.pdf}\\
+    \includegraphics[width=\figurewidth]{datazoom.pdf}
     \end{center}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=1.3\figurewidth]{notebooks/data2.pdf}
+    \includegraphics[width=1.3\figurewidth]{data2.pdf}
     \end{center}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/forward1.pdf}\\
-    \includegraphics[width=\figurewidth]{notebooks/forward2.pdf}
+    \includegraphics[width=\figurewidth]{forward1.pdf}\\
+    \includegraphics[width=\figurewidth]{forward2.pdf}
     \end{center}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/datazoom_interpolated.pdf}
+    \includegraphics[width=\figurewidth]{datazoom_interpolated.pdf}
     \end{center}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/standard1.pdf}\\
-    \includegraphics[width=\figurewidth]{notebooks/standard2.pdf}
+    \includegraphics[width=\figurewidth]{standard1.pdf}\\
+    \includegraphics[width=\figurewidth]{standard2.pdf}
     \end{center}
@@
     \begin{mdframed}\begin{center}
-    \includegraphics[width=\figurewidth]{notebooks/noise.pdf}
+    \includegraphics[width=\figurewidth]{noise.pdf}
     \end{center}
@@
 
-\paragraph{Acknowledgements}
+\begin{acknowledgments}
 It is a pleasure to thank
@@
 This research was supported in part by the Australian Research Council Centre of Excellence for All Sky Astrophysics in 3 Dimensions (ASTRO 3D), through project number CE170100013.
+\end{acknowledgments}
 

### 2025-08-14 paper/main.tex
Ruffio
note re green
Loredo

@@
 
-% To-Do items
-% -----------
+% To-Do items -- at respond to referee
+% ------------------------------------
+% - SEE email from Tom Loredo 2024-05-08.
+% - SEE email from Dylan Green in March 2023
+% - SEE email from "jruffio" about Ruffio et al 2023.
 % - Is it clear in the text that we are taking everything to a standard instrument?
@@
 
+\newcommand{\modified}[1]{\textcolor{orange}{#1}}
+
 \begin{document}
@@
 The output can take the form of an image or spectrum evaluated on a pixel grid, as is traditional.
-In addition to shifts, the model can account for line-spread or point-spread function variations, world-coordinate-system variations, and calibration or normalization variations.
+In addition to shifts, the model \modified{could be extended to} account for line-spread or point-spread function variations, world-coordinate-system variations, and calibration or normalization variations.
 The noise in the output becomes uncorrelated across neighboring pixels as the shifts deliver good coverage in some sense.
@@
 One of the great things about the standard practice---the simple process of shifting and stacking the data---is that this procedure makes almost no assumptions about what the output, combined image or spectrum will look like.
-It isn't really a ``model'' for the data; it is more like a ``summary'' of the data.
+It isn't really a ``model'' for the data; it is more like a ``summary'' of the data \modified{in that the output pixels are summary statistics of the input data}.
 When we switch to a statistical-inference framework, we have to choose some form or basis for expressing the combined image.
@@
 There is a heritage for what's presented here in the ``optimal extraction'' literature (for example, \citealt{oe}, \citealt{kelson}, \citealt{froe}, \citealt{sp}):
-Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.
+Optimal extraction describes a process of extracting a one-dimensional spectral signal from a two-dimensional spectrograph image when that extraction proceeds by forward modeling the two-dimensional data.\footnote{Forward modeling refers to predicting data given some model parameters.
+Inverse modeling refers to inferring some model parameters given some data.}
 One amusing point, which is out of scope here, is that if the data that are being combined are extracted one-dimensional spectra, and those are, in turn, optimal extractions, then there is almost certainly an even better method that moves this spectral combination method directly into the two-dimensional data; that is, there ought to be a method that obviates one-dimensional spectral extraction entirely.
@@
 
-What we do here can apply to almost any astronomical signals, but we are going to specialize to multi-epoch astronomical spectra for specificity.
+What we do here can apply to \modified{many kinds of} astronomical signals.
+\modified{In addition to the direct applications of images and spectroscopy from ultraviolet to the infrared, Frizzle could represent an unseen image plane where complex visibilities are partially observed through interferometry \citep[e.g.,][]{Hilder:2025}.
+However, here we are going to specialize to multi-epoch astronomical spectra for specificity.}
 We are going to make many other assumptions too, all of which are discussed in this \sectionname.
@@
 We will assume that there are $N$ observations (epochs) of the same star (or maybe of different stars or quasars, say, that are assumed to be more-or-less identical, intrinsically, but let's think of a single star for now).
-Each of these observations $i$ (with $1\leq i\leq n$) produces a one-dimensional observed (noisy) image $y_i$.
+Each of these observations $i$ (with $1\leq i\leq \modified{N}$) produces a one-dimensional observed (noisy) image $y_i$.
 The $y_i$ can be thought of as $M_i$-vectors in what follows, where $M_i$ is the number of pixels in spectrum $i$.
@@
 In what follows, each spectrum $y_i$ is given an associated (but possibly inaccurate) noise-variance tensor $C_i$.
-If $y_i$ is an $M_i$-vector, then $C_i$ must be a nonnegative-definite $M_i\times m_i$ 2-tensor.
+If $y_i$ is an $M_i$-vector, then $C_i$ must be a nonnegative-definite $M_i\times M_i$ 2-tensor.
 In this description, there is no need to assume that the different pixels of the spectrum received independent noise, but we do assume that each epoch $i$ has an independent noise draw.
@@
 We won't address this kind of variability here, except for some comments in \secref{sec:discussion}:
-The method presented here can be straightforwardly modified to account for LSF or PSF variations.
+The method presented here \modified{could be modified to account for differential LSF or PSF variations by having the model represent a `highest resolution' spectrum or image, and applying different LSF or PSF kernels to each observation before comparing to data, but this extension is outside the scope of this paper.}
 
@@
 They could be wavelets.
-The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination.
+The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination\modified{, even if we do not require the basis to be orthogonal}.
 Thus we will make use of a linear basis.
@@
 In particular, we will assume that the images or spectra have been bias-corrected, sky-subtracted, flat-fielded, and tellurics-corrected sufficiently well that the different observations $y_i$ can be seen as observations of the same underlying or intrinsic signal.
-Alternatively, the input data might be continuum-normalized or pseudo-continuum-normalized.
+Alternatively, the input data might be continuum-normalized\modified{, pseudo-continuum-normalized, or flux-calibrated}.
 Again, the assumption is that the different calibrated or processed observations are of the same underlying signal.
+\modified{No strict accuracy level is required in these pre-processing steps for \texttt{frizzle} to be executed, but the consistency of the combined spectrum will naturally depend on the consistency of upstream processing steps.}
 
@@
 We do not agree with this constraint or condition.
-The fundamental observable in a telescope is intensity, or energy per area per solid angle per time per wavelength (or per frequency).
+The fundamental observable in a telescope\modified{'s detector} is intensity, or energy per area per solid angle per time per wavelength (or per frequency).
 This quantity is related to the phase-space density of photons.
 When the mapping between a pixel grid and the sky is changed, the expectation of the rate of photon arrivals per pixel will change (because the pixel effective solid angles change), but the expectation of the intensity will not change.
-Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.
-The combined image should provide a good mean estimate of this intensity.
+\modified{Our position is that the spectra or images being combined make the same measurements at the detector surface, and the combined image should provide a good mean estimate of this intensity.}
+%Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.
+%The combined image should provide a good mean estimate of this intensity.
 
@@
 It is no worse a choice than it was (in the past) when you averaged your input data without thinking about noise variances or weights.
-Another sensible default choice would be to set each inverse matrix $C_i^{-1}$ to the identity times the exposure time, such that the eventual math we do will weight the data by the exposure times.
+Another sensible default choice \modified{for photon noise-limited observations} would be to set each inverse matrix $C_i^{-1}$ to the identity times the exposure time, such that the eventual math we do will weight the data by the exposure times.
 That is, you do not need to know the noise variances accurately (or even at all) to execute the forthcoming procedure; but if you do know them, it is good to use them.
@@
 We make a $M\times M$ block-diagonal matrix $C^{-1}$ from all the individual inverse variance matrices $C_i^{-1}$.
-We make a $M\times P$ design matrix $X$ which is the evaluation of all the $P$ basis functions at all the pixel locations in all of the $x_i$.
+We make a $M\times P$ design matrix $X$ which is the evaluation of all the $P$ basis functions at all the \modified{rest frame} pixel locations \modified{$x_i - \Delta{}x_i$}.
 Once we have these things, the best-fit parameters $\hat\theta$ are
@@
 The uncertainty on the combined spectrum $y_\star$ can be estimated in a few ways.
-The information-theoretic answer is that the covariance matrix $C_\star$ delivering the uncertainties on $y_\star$ is
+The information-theoretic answer is that the \modified{inverse covariance matrix $C_\star^{-1}$} delivering the uncertainties on $y_\star$ is
+\modified{
 \begin{align}
-    C_\star &= [X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top]^{-1} ~.\label{eq:outcovar}
+    C_\star^{-1} &= X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top ~.\label{eq:outcovar}
 \end{align}
-The naive uncertainty variances are the diagonals of this matrix (and not the inverses of the diagonals of its inverse).
+}
+The naive uncertainty variances are the \modified{inverse} diagonals of this matrix.
 But, \emph{very importantly}, this is only a correct usage when the input $C_i$ matrices really represented the true uncertainties going in.
@@
 The reason is that we don't want the inverse matrix itself, we want the outcome of the inverse matrix multiplied by $X^\top\,C^{-1}\,Y$.
-For this use case, it is more precise and more more numerically stable to use the linear algebra \texttt{solve()} or \texttt{lstsq()} functions.
-The only use for \texttt{inv()} is where the full inverse of a matrix is required, as when estimating the full $C_\star$ by \eqref{eq:outcovar}.
+For this use case, in \modified{\texttt{Python}} it is more precise and more numerically stable to use the linear algebra \texttt{solve()} or \texttt{lstsq()} functions\modified{, or comparable functions in other languages}.
+The only use for \texttt{inv()} is where the full inverse of a matrix is required, as when estimating the full $C_\star$ by \eqref{eq:outcovar}. % ANDY: or not bother..
 
@@
 When condition numbers get large, maybe a different basis or a re-scale of some basis functions might be a good idea.
+
+\modified{The condition number of the matrix $(X^\top\,C^{-1}\,X)$ will increase when multiple observations sample nearly identical rest-frame wavelengths $x_i - \Delta{}x_i$.
+This degeneracy can occur when two exposures happen to place pixels at very similar rest-frame positions, or in the pathological case of combining spectra that have already been resampled onto a common wavelength grid, which could render $(X^\top\,C^{-1}\,X)$ nearly singular.
+Provided we maintain $P < M$ (where $M = \sum_i M_i$ is the total number of input pixels) and employ robust \texttt{lstsq()} solvers, such near-degeneracies pose no computational difficulties in practice.}
 
@@
 The reason for this is the ``double descent'' phenomenon, which de-stabilizes linear fitting (\citealt{doubledescent}).
+\modified{In practice}
 If you find yourself in the situation that $M_\star\approx M$, reconsider your life choices.
@@
 The spectral expectation model is built on the assumption of a normalized (unit) continuum and unresolved spectral lines with a Gaussian line-spread function (LSF) with one-sigma width $1/R$, where $R=135\,000$ is the spectrograph resolution.
+\modified{We arbitrarily assumed a mean density (Poisson rate) of 2$\times{}10^4$ spectral lines per unit in $x$ (natural logarithm of wavelength).}
 
@@
 In order to give the model maximum flexibility, the number $P$ of Fourier modes in the spectral model is set to be equal to the number of pixels $M_\star$ in the pixel representation.
+\modified{In our experiments here, $M_\star = 338$.
+}
 The log-wavelength scale $L$ of the Fourier basis was set to the pixel spacing times the number of pixels.
@@
 In general this substantially grows the binary mask, and reduces the amount of data available for combination.
-The non-conservative approach is just as bad: if we restrict the number (or fraction) of pixels in the combined spectrum to equal those in the individual epochs, we risk un-masking genuinely bad pixels while growing the binary mask elsewhere!
+The non-conservative approach is just as bad: if we restrict the number (or fraction) of pixels in the combined spectrum to equal those in the individual epochs, we risk \modified{using} genuinely bad pixels while growing the binary mask elsewhere and \modified{marking good output pixels as bad}!
 
@@
 \end{figure}
-We show the final result of Standard Practice(tm)---interpolating the input data and then averaging the interpolations---for the two data sets, in \figurename~\ref{fig:standard}.
+We show the final result of Standard Practice(tm)---interpolating the input data and then \modified{simple averaging}---for the two data sets, in \figurename~\ref{fig:standard}.
 We also show the comparison with the true spectral expectation model that was employed to make the fake data.
@@
 We repeated the above experiments to empirically measure the noise covariances, and to estimate how correlated neighboring pixels are in the output spectrum.
-For the poorly sampled case and the well sampled case, we generated 64 multi-epoch data sets and computed the mean spectrum using \name{} and Standard Practice(tm).
+For the poorly sampled case and the well sampled case, we generated 64 \modified{sets of 8}-epoch data sets and computed the mean spectrum using \name{} and Standard Practice(tm).
 We then computed the empirical covariance in the output spectrum minus the true spectrum, which are shown in \figurename~\ref{fig:noise}.
@@
 The reason for this is that interpolation is very sensitive to noise in the ill-sampled case, such that the inter-pixel value of the interpolated function is moved substantially by small movements in the input pixel values, and these inter-pixel regions in the raw data are large relative to the pixel spacing in the combined spectrum spectral grid.
+\modified{If the sampling in the input data are perfectly sampled, where one epoch is sampled precisely half a pixel offset from the first, then in these situations standard interpolation will not suffer from correlated noise, because interpolation methods are specifically designed for regular sampling patterns.
+However, such situations are rare.
+It would require a stable spectrograph to observe a star with a relative velocity that precisely compensated for the difference in barycentric motion between two (and only two) epochs.}
 
@@
 
-All code used for this project is available at \url{https://github.com/davidwhogg/NoDataInterpolation}.% ANDY to update terminology in code.
-and is registered in the Python Package Index as \texttt{NoDataInterpolation}.
+All code used for this project is available at \url{https://github.com/davidwhogg/NoDataInterpolation} \modified{and is registered in the Python Package Index as \texttt{frizzle}}.
 
@@
 \texttt{scipy} \citep{scipy} ---
-\texttt{matplotlib} \citep{matplotlib}.
+\texttt{matplotlib} \citep{matplotlib} ---
+\texttt{pylops} \citep{pylops} ---
+\texttt{finufft} \citep{finufft}.
 
 \begin{acknowledgments}
-It is a pleasure to thank
+It is a pleasure to thank \modified{the anonymous referee for their detailed review, which helped us to improve this manuscript. We also thank}
     Mike Blanton (NYU),
     Matt Daunt (NYU),
+    \modified{Will Farr (Stony Brook/Flatiron)},
     Dylan Green (UC Irvine),
+    \modified{Thomas Hilder (Monash)},
     Jon Holtzman (NMSU), 
@@
 \bibitem[Wolf et al.(2003)]{combo17} Wolf, C., Meisenheimer, K., Rix, H.-W., et al.\ 2003, \aap, 401, 73. doi:10.1051/0004-6361:20021513
-\bibitem[Zackay, B. \& Ofek (2017)]{zackay} Zackay, B. \& Ofek, E. O.\ 2017, \apj, 836, 188. doi:10.3847/1538-4357/836/2/188
+\bibitem[Zackay \& Ofek (2017)]{zackay} Zackay, B. \& Ofek, E. O.\ 2017, \apj, 836, 188. doi:10.3847/1538-4357/836/2/188
 \bibitem[Zechmeister et al.(2014)]{froe} Zechmeister, M., Anglada-Escud{\'e}, G., \& Reiners, A.\ 2014, \aap, 561, A59. doi:10.1051/0004-6361/201322746

### 2025-08-15 paper/main.tex
tiny tweaks to first parts

@@
 The only cost is a small increase in computational complexity over that of traditional methods.
-We demonstrate the method with a small data example and we provide open-source sample code for re-use.
+We demonstrate the method with a small data example and we provide \modified{fast} open-source sample code for re-use.
 \end{abstract}
@@
 We are in a period in astrophysics in which exceedingly precise measurements are expected, in both imaging and spectroscopy projects.
-In spectroscopy, in particular, high signal-to-noise spectral measurements are expected to obtain surface-abundance measurements with precisions a few percent, and radial-velocity measurements with precisions of up to one ten-thousandth of a \emph{pixel} (in terms of the Doppler shift).
+In spectroscopy, in particular, high signal-to-noise spectral measurements are expected to obtain surface-abundance measurements with precisions of a few percent, and radial-velocity measurements with precisions of up to one ten-thousandth of a \emph{pixel} (in terms of the Doppler shift).
 Spectral combination methods that distort absorption (or emission) lines will translate into biases, errors, or increased variance in these measurements.
@@
 Indeed it has just as much freedom as the shifting-and-stacking standard practice.
-The forward-modeling we recommend gets its power from the fact that it generates the data directly and without manipulation, by means of a justifiable likelihood function.
+The forward-modeling we recommend gets its power from the fact that it generates the input data directly and without manipulation, by means of a justifiable likelihood function.
 
@@
 The \emph{output product} is the same form as what one would expect from an interpolation-and-averaging procedure (values at a set of pixel positions), but the \emph{procedure} differs from interpolation.
-When interpolating the input data the basis functions are explicitly constructed from the input-data pixel grid, whereas here we construct a forward model that is maximally consistent with the data, and can be evaluated at any pixel locations.
+When interpolating, the basis functions are explicitly constructed from the input-data pixel grid, whereas here we construct a forward model that is maximally consistent with the data, and can be evaluated at any pixel locations.
 
@@
 They could be wavelets.
-The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination\modified{, even if we do not require the basis to be orthogonal}.
+The point is that if we use a linear basis, and assume that noises are Gaussian, we will get closed-form expressions for the data combination\modified{, even if the basis is not orthogonal}.
 Thus we will make use of a linear basis.

### 2025-08-17 paper/main.tex
small adjustments

@@
 When the mapping between a pixel grid and the sky is changed, the expectation of the rate of photon arrivals per pixel will change (because the pixel effective solid angles change), but the expectation of the intensity will not change.
-\modified{Our position is that the spectra or images being combined make the same measurements at the detector surface, and the combined image should provide a good mean estimate of this intensity.}
+\modified{Our position is that the spectra or images being combined make measurements of the same intensity at the detector surface; the combined image should provide a good mean estimate of this intensity.}
 %Our position is that the key idea is not the conservation of flux or photons, but that the spectra or images being combined make the same measurements of the intensity (or similarly normalized or otherwise consistently calibrated measurements of intensity) at the detector surface.
@@
 \begin{align}
-    C_\star^{-1} &= X_\star\,(X^\top\,C^{-1}\,X)^{-1}\,X_\star^\top ~.\label{eq:outcovar}
+    C_\star^{-1} &= X_\star\,(X^\top\,C^{-1}\,X)\,X_\star^\top ~.\label{eq:outcovar}
 \end{align}
-}
-The naive uncertainty variances are the \modified{inverse} diagonals of this matrix.
+}The naive uncertainty variances are the \modified{inverse} diagonals of this matrix.
 But, \emph{very importantly}, this is only a correct usage when the input $C_i$ matrices really represented the true uncertainties going in.
@@
 The reason for this is the ``double descent'' phenomenon, which de-stabilizes linear fitting (\citealt{doubledescent}).
-\modified{In practice}
-If you find yourself in the situation that $M_\star\approx M$, reconsider your life choices.
+\modified{In practice if}
+you find yourself in the situation that $M_\star\approx M$, reconsider your life choices.
 
